### 2019-12-10 ms/ms.Rmd
initial commit

---
title: "A stomatal model of anatomical tradeoffs between photosynthesis and pathogen defense"
runtitle: "Stomata tradeoff photosynthesis for defense"
documentclass: frontiersSCNS
author:
  - name: Christopher D. Muir
    affiliation: '1'
    etal: Muir
    email: cdmuir@hawaii.edu
affiliation:
  - id: '1'
    department: School of Life Sciences
    institution: University of Hawaii
    city: Honolulu
    state: Hawaii
    country: USA
output: 
  rticles::frontiers_article:
    includes:
      in_header: header.tex
csl: frontiers.csl
bibliography: stomata-tradeoff.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{abstract}

Stomatal pores control both leaf gas exchange and are an entry for many plant pathogens, setting up the potential for tradeoffs between photosynthesis and defense. To prevent colonization, plants close their stomata after recognizing pathogens to limit infection. Anatmoical shifts to lower stomatal density and/or size may also limit pathogen colonization, but would permanently reduce the gas exchange capacity for the life of a leaf. I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as function of stomatal size and density. The model predicts that changes in stomatal density or size increase the probability of host plant colonization most when stomatal cover is low. I derived scaling relationships between stomatal size and density that preserves a given probability of colonization. These scaling relationships set up a potential conflict between maximizing defense and minimizing stomatal cover. To my knowledge, this is first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy. It makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants. The model is generalizable to many types of pathogens, but lacks significant biological realism that may need to be added in future extensions.

\tiny
 \keyFont{ \section{Keywords:} anatomy, leaf gas exchange, model, pathogen, photosynthesis, scaling, stomata, tradeoff }

\end{abstract}

# Introduction {-}

<!-- overview -->
Stomata evolved to let plants regulate gas exchange in and out of the leaf [@hetherington_role_2003; @berry_stomata:_2010; @chater_origins_2017], but many plant pathogens take advantage of these chinks in the leaf cuticular armor to infect prospective hosts [@zeng_plant_2010; @mclachlan_gate_2014; @melotto_stomatal_2017]. The density and size of stomata set the anatomical maximum rate of stomatal conductance to CO$_2$ and water vapor [@brown_static_1900; @parlange_stomatal_1970; @franks_effect_2001; @franks_maximum_2009; @lehmann_effects_2015; @sack_developmental_2016; @harrison_influence_2019], but the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019]. Many plant pathogens, including viruses [@murray_plant_2016], bacteria [@melotto_plant_2006; @underwood_role_2007], protists [@fawke_oomycete_2015], and fungi [@hoch_signaling_1987; @zeng_plant_2010] use stomatal pores to gain entry to the leaf. Since stomatal conductance is a major limitation on photosynthesis [@farquhar_stomatal_1982; @jones_partitioning_1985] while pathogens reduce fitness, this sets up a potential tradeoff between increased photosynthesis and defense against pathogens. Although there have been many empirical studies on the effect that pathogens have on stomata, there is no theoretical framework in which to place these findings. Lack of theoreatical framework makes it difficult to answer general questions about how selection for pathogen defense constrains maximum rates of gas exchange.

<!-- paragraph on stomatal anatomy and gsmax -->
Stomatal anatomy is the key link between gas exchange and pathogen colonization. The density and size of stomata not only determines the theoretical maximum stomatal conductance (\gsmax; @harrison_influence_2019), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019]. Therefore, I use anatomical \gsmax{} as a proxy for \gsop{} and do not address dynamic changes in stomatal aperture (see [Discussion]). @harrison_influence_2019 recently reviewed the relationship between stomatal anatomy and gas exchange in detail.

<!-- paragraph on stomatal anatomy and pathogen defense -->
Many pathogens rely on stomata to gain entry into the leaf, but it is unclear how anatomical traits that determine \gsmax{} (size and density) affect the ability of pathogens to colonize leaves. The impact of pathogens on host fitness is complex, but after a pathogen reaches a host, the first major step is colonization. Once infected, a pathogen can reduce fitness. Susceptible hosts can lose much of their biomass or die, but even resistant hosts must allocate resources to defense or reduce photosynthesis because of necrosis around sites of infection. Plants can limit colonization physiologically by closing stomata after they recognize pathogens, called stomatal defense [@melotto_stomatal_2017]. Anatomy may be another layer of defense; plants can reduce pathogen colonization by developing leaves with lower stomatal density and/or size. Infection increases in leaves with higher stomatal density [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019], which suggests that both anatomy and physiological responses (stomatal closure) affect host colonization. By the same logic, if stomatal density were held constant, larger stomata should also increase colonization because they occupy more area. One key difference between physiological and anatomical defenses is that stomatal closure reduces gas exchange transiently, whereas anatomy constrains gas exchange throught the life of a leaf. Because we do not understand the relationship between stomatal anatomy and colonization well, we do not know the relationship between \gsmax{} and infection.

<!-- paragraph on size-density scaling? -->
If stomatal size and density affect pathogen colonization, selection to limit colonization may shape stomatal size-density scaling relationships. Botanists have long recognized that stomatal size and density are inversely correlated [@weiss_untersuchungen_1865; @ticha_photosynthetic_1982; @hetherington_role_2003; @sack_hydrology_2003; @franks_co_2009; brodribb_unified_2013; @de_boer_optimal_2016], but the evolutionary origin of this relationship is not yet known. Here I argue that pathogens could shape selection on this relationship. Explanations for inverse size-density scaling are usually cast in terms of preserving \gsmax{} and/or total epidermal area allocated to stomata [@de_boer_optimal_2016] because there are many combinations of stomatal size and density that have same \gsmax{} or same \fs{}:

\begin{align}
  g_\text{s,max} & = b m D S ^ {0.5} \label{eq:gsmax} \\
  f_\text{S} & = D S \label{eq:fs}.
\end{align}

$b$ and $m$ are assumed to be biophysical and morphological constants, *sensu* @sack_developmental_2016 (see [Supplementary Material]. If size and density also affect pathogen colonization, then selection from foliar pathogens could significantly alter the size-density scaling relationship. The empirical size-density scaling relationship is linear on a log-log scale, determined by an intercept $\alpha$ and slope $\beta$:

\begin{align}
  D & = e ^ \alpha S ^ {-\beta}; \\
  d & = \alpha - \beta S.
\end{align}

For brevity, $d = \text{log}(D)$ and $s = \text{log}(S)$. Rearranging Equations \ref{eq:gsmax} and \ref{eq:fs}, a scaling relationship where $\beta = 0.5$ preserves \gsmax{} while $\beta = 1$ preserves \fs.

How would adding pathogens alter these predicted scaling relationships? For simplicity, imagine two environments, one without foliar pathogens and one with lots. In the absence of foliar pathogens, we expect size-density scaling to preserve \gsmax{}, \fs{}, or some least-cost combination of them. What happens when we introduce pathogens? Assuming that stomatal size and density increase pathogen colonization, then selection will favor reduced size and/or density. This would change the intercept $\alpha$ but not the slope. The effect of foliar pathogens on the slope depends on the relationship between size, density, and probability of colonization If the probability of colonization is proportional to the product of *linear* stomatal size ($S ^ {0.5}$) and density ($\propto D S ^ {0.5}$ as for \gsmax) then it has the same effect on the slope as \gsmax{} because there are many combinations of $D$ and $S ^ {0.5}$ that have same probability of colonization. If the probability of colonization is proportional to the product of *areal* stomatal size ($S$) and density ($\propto D S$ as for \fs) then it has the same effect on the slope as \fs{} because there are many combinations of $D$ and $S$ that have same probability of colonization. Alternatively, the probability of colonization may have a different scaling relationsip (neither 0.5 nor 1) or may be nonlinear on a log-log scale. Unlike \gsmax{} and \fs{}, we do not have theory which predicts a stomatal size-density relationship which preserves the probability of colonization.

<!-- paragraph on what we don't know -->
In summary, the physical relationship between stomatal size, density, and conductance is well established [@harrison_influence_2019]. The same traits likely affect the probability of pathogen colonization, but we do not have a theoretical model that makes quantitative predictions. The inverse stomatal size-density relationship has usually been explained in terms of preserving stomatal conductance and/or stomatal cover, but selection by pathogens might alter scaling. To address these gaps, the goals of this study are to 1) introduce a spatially explicit, probabalistic model pathogen colonization on the leaf surface; 2) use the model to predict the relationship between \gsmax, \fs{}, and the probability of colonization; 3) work out what these relationships predict about stomatal size-density scaling.

# Model {-}

In this section, I introduce a spatially explicity, probabilistic model of pathogen colonization on a leaf surface. I explain the model structure and assumptions. The [Methods] section below describes how I analyzed the model to address the goals of the study. I used Sympy version 1.3 [@meurer_sympy:_2017] for symbolic derivations. For generality, I refer to a generic ``pathogen'' that lands on leaf and moves to a stomate. The model is agnostic to the type of pathogen (virus, bacterium, fungus, etc.) and the specific biological details of how it moves (biotrophy). For example, motile bacterial cells can land and move around where fungi may germinate from a cyst and grow until they form an appresorium for infection. These very different biotropic movements on the leaf are treated identically here.

**Table 1** | Glossary of mathematical symbols. The columns indicate the mathematical Symbol used in the paper, the associated symbol used in R scripts,  scientific Units, and a verbal Description.

 Symbol             | R         | Units                 | Description
--------------------|-----------|-----------------------|-----------------------
 $D,d$              | `D, d`    | $\textrm{mm}^{-2}$    | stomatal density ($d = \text{log}~D$)
 $f_s$              | `f_s`     | none                  | stomatal cover ($f_s = D S$)
 $g_\text{s,max}$   | `g_smax`  | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | anatomical maximum stomatal conductance 
 $g_\text{s,op}$    | `g_sop`   | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | operational stomatal conductance 
 $L$                | `L`       | $\mu\textrm{m}$       | guard cell length ($L = 2 R$)
 $R$                | `R`       | $\mu\textrm{m}$       | stomatal radius ($S = 2 \pi R ^ 2$)
 $S,s$              | `S, s`    | $\mu\textrm{m}^2$     | stomatal size ($s = \text{log}~S$)
 $U$                | `U`       | $\mu\textrm{m}$       | interstomatal distance
 $v_i$              | `v_i`     | $\mu\textrm{m}$       | distance between pathogen and stomate $i$
 $\theta_i$         | `theta_i` | radians               | angles between $x_p,y_p$ and stomate $i$
 $H$                | `H`       | $\mu\textrm{m} ^{-1}$ | death rate of pathogen on leaf surface

## Spatial representation of stomata {-}

Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015; earlier citations of stomatal interference]. Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$. This assumption ignore veins, trichomes, and within-leaf variation in stomatal density. Stomata are therefore arrayed in an evenly spaced grid (Figure \ref{fig:fig1}a). The interstomatal distance $U$, measured as the distance from the center of one stomata to the next, is the maximal diagonal of the hexagon in $\mu\textrm{m}$ that forms an equal area boundary between neighbording stomata. The area of a hexagon is $A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2$. By definition the stomatal density is the inverse of this area, such that $D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}$. Therefore, interstomatal distance can be derived from the stomatal density as:

$$ D = \frac{2}{\sqrt{3}} U ^{-2} $$
$$ U = \bigg(\frac{2}{\sqrt{3}} D ^{-1}\bigg) ^ {0.5} $$

For example, if the density is $D = 10 ^ 2~\textrm{mm}^{-2} = 10 ^ {-4}~\mu\textrm{m}^{-2}$, then $U$ is `r round(sqrt(2 / (sqrt(3) * 1e-4)), 1)` $\mu \textrm{m}$. @parkhurst_diffusion_1994 described this result previously. I also make the simplifying assumption that stomata are perfectly circular with radius $R$. This may be approximately true for fully open stomata with kidney-shaped guard cells.

\begin{figure}
  \centering
	\includegraphics[width=6.5in]{../figures/fig1.jpeg}
	\caption{\textbf{A spatially explicit model of stomatal anatomy and pathogen colonization}. \textbf{a}. Stomata are assumed to be in a homogenous equilateral triangular grid, which means that we can extrapolate from \textbf{b.} a focal triangle to the entire leaf. The circle indicate idealized stomata; the grey lines between them are for visualization. \textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the area. \textbf{d.} The model assumes that a pathogen depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves in a certain angle, depicted by the grey polygons.}
	\label{fig:fig1}
\end{figure}

## Spatial representation of pathogen search {-}

Now imagine that a pathogen lands at a random position on the leaf and must arrive at a stomate to colonize. If it lands on a stomate, then it infects the leaf with probability 1; if it lands between stomata, then it infects the leaf with probability $p_\text{locate}$. This is the probability that it locates a stomate, which I will derive below. The probabilities of landing on and between a stomate are \fs{} and $1 - $ \fs{}, respectively. Hence, the total probability of infection is:

\begin{equation}
  \label{eq:p_colonize}
  p_\text{colonize} = f_\text{S} + (1 - f_\text{S}) p_\text{locate}.
\end{equation}

I assume that the pathogen cannot sense where stomata are and orients at random, thereafter traveling in that direction. If it successfully locates a stomate it colonizes the leaf, but otherwise does not infect. If there is a high density of stomata and/or large stomata, the probability of locating a stomate increases. By assuming that stomata form an equilateral triangular grid (see above), we can extrapolate what happens in a focal triangle (Figure \ref{fig:fig1}b) by symmetry. Further, since an equilateral triangle can be broken up into six identical units (Figure \ref{fig:fig1}c), we can simply calculate $p_\text{locate}$ in this focal area.

Imagine that the pathogens lands in position ($x_p,y_p$) within the triangle. The centroid of the triangle is at position ($x_c,y_c$) and a reference stomate is at position $(0,0)$ (Figure \ref{fig:fig2}). Therefore $x_c = U / 2$ and $y_c = \sqrt{3} U / 6$. The other stomata are at positions $(U/2, \sqrt{3} U /2)$ and $(U,0)$ (Figure \ref{fig:fig2}). $x_p$ and $y_p$ are defined as the horizontal and vertical distances, respectively, from the pathogen to the reference stomate at position $(0,0)$.

Given that the pathogens starts at position ($x_p,y_p$), what's the probability of contacting one of the stomata at the vertices of the focal triangle? I assume the probability of contacting a stomate is equal to the proportion of angular directions that lead to a stomate (Figure \ref{fig:fig1}d). I solved this by finding the angles ($\theta_1, \theta_2, \theta_3$) between lines that are tangent to the outside of the three stomata and pass through ($x_p,y_p$) (Figure \ref{fig:fig2}). If stomata $i$ is centered at ($x_i,y_i$), the two slopes of tangency as function of pathogen position are:

\begin{align}
  t_{i,1}(x_p,y_p) & = \frac{-R e_{i,2}(x_p,y_p) + e_{i,3}(x_p,y_p)}{e_{i,1}(x_p,y_p)} \\
  t_{i,2}(x_p,y_p) & = \frac{R e_{i,2}(x_p,y_p) + e_{i,3}(x_p,y_p)}{e_{i,1}(x_p,y_p)}
\end{align}

where

\begin{align}
  e_{i,1}(x_p,y_p) & = (R^2 - x_i^2 + 2 x_i x_p - x_p^2), \\
  e_{i,2}(x_p,y_p) & = \sqrt{-e_{i,1} + (y_i - y_p) ^ 2}, \\
  e_{i,3}(x_p,y_p) & = - x_i y_i + x_i y_p + x_p y_i - x_p y_p.
\end{align}

Note that $i \in \{1, 2, 3\}$, indexing the three stomata in the focal triangle. The angle in radians between $t_{i,1}(x_p,y_p)$ and $t_{i,2}(x_p,y_p)$ is:

\begin{equation}
  \theta_i(x_p,y_p) = \text{arctan}\bigg(\frac{t_{i,1}(x_p,y_p) - t_{i,2}(x_p,y_p)}{1 + (t_{i,1}(x_p,y_p) t_{i,2}(x_p,y_p))}\bigg)
\end{equation}

\begin{figure}
\centering
\input{../figures/fig2.tex}
\caption{\textbf{Spatial representation of stomata and pathogen.} \textbf{a.} The pathogen starts at a uniform random position within the focal region denoted ($x_p,y_p$). Within the focal triangle, the reference stomata is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$. \textbf{b .} Within the focal region, a pathogen can land within the stomate (white cirlce with grey outline and radius $R$) or in the grey area. The outer borders of this area are shown and depend on $R$ and $U$. For a given position $x$, there is a minimum $y$-value ($y_\text{min}$, dashed line) and maximum $y$-value ($y_\text{max}$, solid line).}
\label{fig:fig2}
\end{figure}

I further assumed that the longer distance a pathogen must travel, the less likely it would be to locate a stomate. For example, if stomata are at very low density, then a pathogen may die before it reaches a stomate because of UV, dessication, or another factor. I included this effect by assuming the probability of reaching a stomata declines exponentially at rate $H$ with the Euclidean distance $v_i(x_p,y_p)$ between the pathogen location and the edge of stomata $i$, which is distance $R$ from its center at $x_i,y_i$:

\begin{equation}
  v_i = \sqrt{(x_i - x_p) ^ 2 + (y_i - y_p) ^ 2} - r.
\end{equation}

The probability of locating a stomate as a function of $x_p$ and $y_p$ ($f_\text{locate}(x_p,y_p)$) is the sum of the angles divided by $2\pi$, discounted by their distance from the stomate:

\begin{equation}
  f_\text{locate}(x_p,y_p) = \frac{1}{2 \pi} \sum_{i=1}^3 e^{-H v_i(x_p,y_p)} \theta_i(x_p,y_p)
\end{equation}

When $H = 0$, $p_\text{locate}$ is the fraction of angles that lead from $(x_p,y_p)$ to a stomate. When $H > 0$, $p_\text{locate}$ is proportional to this fraction, but less than it depending on stomatal density, size, and starting location of the pathogen.

To obtain the average $p_\text{locate}$, we must integrate $f_\text{locate}(x_p,y_p)$ over all possible starting positions $(x_p,y_p)$ within the focal area. The focal area is a 30-60-90 triangle with vertices at the reference stomate $(0,0)$, the midpoint of baseline $(U/2,0)$, and the centroid of the focal trinagle $(U/2, \sqrt{3} / 6 U)$ (Figure \ref{fig:fig1}c). Colonization occurs if the pathogen lands in the reference stomate, so we need to integrate from the edge of this stomate, from $\sqrt{3} / 2 R$ to $U/2$ (Figure \ref{fig:fig2}b). At any $x$, we integrate from the bottom of the focal area ($y_\text{min}$) to the top ($y_\text{max}$):

\begin{align}
  y_\text{min} = f(x) & = \begin{cases}
    \sqrt{R ^ 2 - x ^ 2}, & \text{if } \frac{\sqrt{3}}{2} R < x < R \\ 
    0, & \text{if } R \leq x \leq \frac{U}{2} \end{cases} \\
  y_\text{max} = f(x) & = \frac{\sqrt{3}}{3} x
\end{align}

The integral is:

\begin{equation}
  p_\text{locate} = \int\limits_{\frac{\sqrt{3}}{2} R} ^ {U / 2} \int\limits_{y_\mathrm{min}} ^ {y_\mathrm{max}} f_\text{locate}(x, y) \,dx \,dy
\end{equation}

# Materials and Methods {-}

The [Model] calculates a probability of host colonization (Equation \ref{eq:p_colonize}) as a function stomatal density, size, and position of a pathogen on the leaf. I solved $p_\text{colonize}$ using the `integral2()` function in the **pracma** package version 2.2.5 [@borchers_pracma:_2019] for numerical integration. I used R version 3.6.1 [@r_core_team_r:_2019] for all analyses and the paper using **rmarkdown** version 1.17 [@xie_r_2018; @allaire_rmarkdown:_2019]. Source code is deposited on GitHub (https://github.com/cdmuir/stomata-tradeoff) and will be arcived on Zenodo upon publication.

## What is the relationship between stomatal size, density, and the probability of colonization? {-}

I calculated $p_\text{colonize}$ over a biologically plausible grid of stomatal size and density for hypostomatous species based on @de_boer_optimal_2016. Stomatal density ranges from $10^1 -- 10^{3.5}~\text{mm}^{-2}$; stomatal size ranges from $10^1 -- 10^{3.5}~\mu\text{m}^2$. I only considered combinations of size and density where \fs{} was less than 0.5. For simplicity, I have not extended the current analysis to amphistomatous leaves. I crossed stomatal traits with three levels of $H \in \{0, 0.01, 0.1\}$. When $H = 0$, a pathogen perists indefinitely on the leaf surface. $H = 0.01$ and $H = 0.1$ correspond to low and high death rates, respectively. These values are not necessarily realistic, but illustrate qualitatively how a hostile environment on the leaf surface alters model predictions.

## How do pathogens alter optimal stomatal size-density scaling? {-}

The stomatal size-density scaling relationship can be explained in terms of preserving a constant \gsmax{} that is proportional to $D S ^ {0.5}$ when $bm$ is constant (Equation \ref{eq:gsmax}). In other words, there are infinitely many combinations of $D$ and $S ^ {0.5}$ with the same \gsmax. If \gsmax{} is held constant at $C_g$, then the resulting size-density scaling relationship on a log-log scale is:

$$ d = c_g - 0.5 s $$
where lowercase variables are log-transformed equivalents of their uppercase counterparts. To preseve $C_g$, the scaling exponent $\beta_g = 0.5$.

Next, imagine there is similarly a scaling exponent $\beta_p$ that preserves $p_\text{colonize}$ for the product $D S^{\beta_p}$. If $\beta_p = 0.5$, then $p_\text{colonize}$ is always proportional to \gsmax. If $\beta_p < 0.5$, small, densely packed stomata would be better defended (lower $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax. If $\beta_p > 0.5$, small, densely packed stomata would be less defended (higher $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax. I refer to the three outcomes ($\beta_p = 0.5$, $\beta_p < 0.5$, and $\beta_p > 0.5$) as iso-, hypo-, and hyper-conductance, respectively. I was unable to solve analytically for $\beta_p$, so I numerically calculated isoclines of $p_\text{colonize}$ over the grid of $D$ and $S$ values described in the preceding subsection. I numerically calculated the scaling relationships at a constant $p_\text{colonize} \in \{0.025, 0.05, 0.1, 0.2, 0.4\}$ for $H \in \{0, 0.01, 0.1\}$.

# Results {-}

I analyzed an idealized, spatially explicit [Model] of how a pathogen lands on a leaf and finds a stomate to colonize the leaf using a random search. To my knowledge, this is the first model that makes quantitative predictions about the relationship between stomatal anatomy, the probability of colonization, and their impact on stomatal size-density scaling.

## Nonlinear relationships between colonization, stomatal cover, and conductance {-}

The probability of colonization ($p_\text{colonize}$) is not simply a one-to-one relationship between the fraction of epidermal area allocated to stomata (\fs). At low \fs, $p_\text{colonize}$ is much greater than \fs{} because a pathogen can either land on a stomata with probability \fs{} or locate a stomate after land, which must be added to \fs{} (Equation \ref{eq:p_colonize}); Figure \ref{fig:fig3}a). As greater \fs{} and $p_\text{colonize}$ have a linear relationship. When $H = 0$, any combination of stomatal size ($S$) and density ($D$) with the same \fs{} have the same effect on $p_\text{colonize}$. When $H > 0$, pathogens are less likely to land close enough to a stomate to infect before dying, so $p_\text{colonize}$ is closer to \fs{} (Figure \ref{fig:fig3}a). Furthermore, $p_\text{colonize}$ depends on $D$ and $S$, not just \fs. For the same \fs, leaves with greater $D$ have higher $p_\text{colonize}$ (Figure \ref{fig:fig3}a). Holding \fs{} constant, leaves with lower $D$ and higher $S$ will have a greater distance ($v_i$) between a pathogen and its stomata. When $H > 0$, this extra distance leads more pathogens to die before they can find a stomate. In contrast to \fs, $p_\text{colonize}$ increases at a greater than linear rate with \gsmax{} and greater $D$ (smaller $S$) is associated with lower $p_\text{colonize}$ when \gsmax{} is constant (Figure \ref{fig:fig3}b). This happens because $p_\text{colonize}$ increases approximately linearly with $S$ whereas \gsmax{} is proportional to $S^{0.5}$

\begin{figure}
  \centering
	\includegraphics{../figures/fig3.pdf}
	\caption{\textbf{The probability of colonization increases with both stomatal cover and conductance} \textbf{a.} The probability of colonization ($p_\text{colonize}$, $y$-axis) initially increases rapidly with stomatal cover (\fs), then slows down to a linear relationship. Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$). The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density when $H = 0$. When $H > 0$, higher density (solid lines) increase $p_\text{colonize}$. \textbf{b.} $p_\text{colonize}$ increases expoential with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ decreases at higher density for a given \gsmax.}

	\label{fig:fig3}
\end{figure}

## Hyper-conductance size-density scaling {-}

The scaling relationship between $S$ and $D$ that preserves $p_\text{colonize}$ is always greater 0.5 (hyper-condtance), but usually less than 1. When $H = 0$, the scaling relationship is essentially 1 (Figure \ref{fig:fig4}), which means that an increase \fs{} leads to a proportional increase in $p_\text{colonize}$. Because the scaling relationship is greater than 0.5, leaves with greater stomatal density will have lower $p_\text{colonize}$ than leaves lower stomatal density but the same \gsmax. In other words, increasing $D$ and lowering $S$ allows plants to reduce $p_\text{colonize}$ while maintaining \gsmax. The scaling relationship is slightly less than 1, but still greater than 0.5, when $H > 0$ (Figure \ref{fig:fig4}). In this area of parameter space, lower stomatal density can reduce \fs{} while $p_\text{colonize}$ is constant, but this will still result in lower \gsmax.

\begin{figure}
  \centering
	\includegraphics{../figures/fig4.pdf}
	\caption{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$). In each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.025 (lowest line), 0.05, 0.1, 0.2, or 0.4 (highest line). For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.6$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region. When the death rate on the leaf surface is low ($H = 0$), the scaling exponent is very close to $\beta = 1$. When $H > 0$, $0.5 < \beta < 1$ and is slightly nonlinear on a log-log scale.}
	\label{fig:fig4}
\end{figure}

# Discussion {-}

Stomatal density and size set the upper limit on gas exchange in leaves [@harrison_influence_2019] and is often closely related to operational stomatal conductance in nature [@murray_consistent_2019].  Despite the fact that many ecologically and economically significant plant pathogens infect through stomata, the relationship between stomatal anatomy and susceptibility to foliar pathogens is less clear than it is for gas exchange. To develop testable predictions, we need mathematical models that can clarify the potential for tradeoffs between stomatal conductance, stomatal cover, and disease resistance. I used a spatially explicit model of a pathogen searching for a stomate to colonize the host to derive predictions about the relationship between stomatal anatomy and disease resistance for the first time. The model predicts that the probability of colonization is not always proportional to the surface area of leaf covered by stomata (\fs), as one might intuitively predict. If the leaf surface is a hostile environment and pathogens have a limited time to search, lower stomatal density decreases the probability of colonization even if \fs{} is constant. However, \gsmax{} decreases proportionally more than the probability of colonization. The model reveals the potential for conflicting demands of maximizing disease resistance, minimizing stomatal cover, and maintaining stomatal conductance. Including the effect of anatomy on disease resistance has the potential to change our understading of how stomatal size-density scaling evolves in land plants.

<!-- paragraph on first main result -->
The model predicts that in most cases, increasing stomatal cover should lead to a proportonal increase in susceptibility, which is the implicit assumption of some empirical studies (e.g. @mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019). It also makes new, testable predictions that are less intuitive. At very low \fs, there is a rapid increase in susceptibility (Figure \ref{fig:fig3}a). If there are no stomata, the probability of colonization is 0, so the first few stomata dramatically increase the probability. This is unlikely to be significant for abaxial (lower) leaf surfaces, which usually have most of the stomata [@salisbury_causes_1928; @metcalfe_anatomy_1950; @mott_adaptive_1984; @peat_comparative_1994; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @drake_two_2019]. However, many adaxial (upper) leaf surfaces have zero or very few stomata. Using adaxial leaf surfaces, it should be possible to test if small changes in stomatal size or density have a larger effect on disease susceptibility when \fs{} is low. The nonlinear increase in $p_\text{colonize}$ is less apparent when $H > 0$ (Figure \ref{fig:fig3}a). A more hostile microenvironment (e.g. drier, higher UV) should therefore reduce the effect of increased size or density as low \fs. If true, the dimishing marginal effect of \fs{} on colonization could explain why stomatal ration on the upper and lower surface is bimodal [@muir_making_2015]. The initial cost of adaxial (upper) stomata is high, but if the benefits outweigh the costs, then equal stomatal densities on each surface maximize CO$_2$ supply for photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990].

<!-- paragraph on scaling -->
An effect stomatal size and density on susceptibility to foliar pathogens could profoundly change our understanding of stomatal size-density scaling. Since allocating leaf epidermis to stomata may be costly [@franks_mechanical_2007; @assmann_guard_1987; @dow_physiological_2014; @lehmann_effects_2015; @baresch_competition_2019], selection should favor leaves that achieve a desired \gsmax{} while minimizing \fs{} [@de_boer_optimal_2016]. Because of their different scaling exponents (Equation \ref{eq:gsmax}, \ref{eq:fs}), smaller, densely packed stomata can achieve the same \gsmax{} at minimum \fs. However, many leaves have larger, sparsely packed stomata. Incorporating susceptibility to disease may explain why. If pathogens have a limited time to find stomata before dying ($H > 0$), then the scaling exponent between size and density that keeps $p_\text{colonize}$ constant is between 0.5 and 1, the scaling exponents for \gsmax{} and \fs, respectively (Figure \ref{fig:fig4}). Greater density of smaller stomata can increase \gsmax{} while keeping $p_\text{colonize}$ constant, but this will increase \fs. Conversely, \fs{} could decrease while keeping $p_\text{colonize}$ constant, but this will decrease \gsmax. This sets up the potential for conflict between competing goals. The optimal stomatal size and density will therefore depend on the precise costs and benefits of infection, stomatal conductance, and stomatal cover. This may explain why many leaves have large, sparsely packed stomata despite the fact that they could achieve the same \gsmax{} and lower \fs{} with smaller, more densely packed stomata

<!-- scope of model -->
The model examines the probability of colonization for a single pathogen and is most applicable to diseases where the host has some resistance. The energetic cost and lost photosynthetic capacity (closed stomata, necrosis, etc.) of dealing with a pathogen is assumed to be proportional to the amount of infection. The actual fitness cost will be modulated by the number of pathogens landing on the leaf and the cost of infection. In environments with fewer or less virulent pathogens, the fitness cost of infection will be less than in environments with more abundant, virulent pathogens.
The model is less relevant to very susceptible host plants that can be severely damaged or killed by a small number of colonizations that spread unchecked throughout the host tissue.

<!-- other limitations -->
The purpose of this model is to provide a general foundation to examine the relationship between stomatal size, density, and defense against foliar pathogens. In its generality, it overlooks interesting natural history and biologically important features of specific plants and their pathogens. For example, some pathogens actively seek out and find stomata [@kiefer_host_2002], whereas I have assume that pathogens randomly orient themselves on the leaf. Including sensing should increase the probability of colonization. I also assumed that the plant does not respond by, for example, closing stomata when it senses a pathogen. However, plants can sense and close stomata when pathogens land on the leaf, but pathogens can pry stomata back open [@melotto_plant_2006]. Future work on specific plant-pathogen interactions could build on this model by adding more biological realism to provide more precise predictions.

# Conclusion {-}

The model makes two non-intuitive predictions. First, the effect of increased stomatal density or size on susceptibility to foliar pathogens is greatest when stomatal coverage is very low. Second, maximizing disease resistance sets up a potential conflict between minimizing stomatal cover and maximizing stomatal conductance. The first prediction may be relatively straightforward to test experimental with adaxial (upper) stomata that occur at low and moderate densities within the same or closely related species [@muir_morphological_2014; @mckown_association_2014; @fetter_trade-offs_2019]. The second prediction about size-density scaling is more complex because we would to know the relationships between colonization, stomatal cover, stomatal conductance, and fitness in natural conditions. Testing these predictions in a variety of species would help determine whether pathogens have played an important role shaping  stomatal anatomy in land plants.

<!-- # Acknowledgements {-} -->

# Funding {-}
I am grateful startup funds from the University of Hawaii for supporting this work.

# Conflict of Interest Statement {-}
The author declares that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.

# Supplementary Material {-}

**Table S1** | Glossary of symbols only used in \gsmax{} calculations.

 Symbol | R      | Units  | Description
--------|--------|--------|-----------------------------------------------------
 $D_w$              | `D_w`    | $\textrm{m}^{2}~\textrm{s}^{-1}$ | diffusion coefficient for water vapour in air at $T_\mathrm{leaf}$ and $P$
 $D_{w,0}$          | `D_w0`   | $\textrm{m}^{2}~\textrm{s}^{-1}$ | diffusion coefficient for water vapour in air at 0 $^\circ$ C and air pressure 101.3246 kPa
 $eT$               | `eT`     | none | stomatal cover ()
 $P$                | `P`      | kPa | atmospheric pressure
 $T_\mathrm{leaf}$  | `T_leaf` | leaf temperature |

### 2019-12-10 python/tangents.py

@@ -1,0 +1,59 @@
+# stomata are a circle with radius R, centered at (x_c, y_c)
+# R ^ 2 = (x - x_c) ^ 2 + (y - y_c) ^ 2
+# pathogen is located at point (x_p, y_p)
+# point-slope form of line from pathogen tangent to stomata
+# y - y_p = m * (x - x_p)
+# y = m * x + b, where b = y_p - m * x_p
+#
+# solve for intersection between line and stomata where:
+# R ^ 2 = (x - x_c) ^ 2 + (y - y_c) ^ 2
+# R ^ 2 = (x - x_c) ^ 2 + (m * x + b - y_c) ^ 2
+# 0 = (1 + m ^ 2) * x ^ 2 + 
+#       2 * (b * m - m * y_c - x_c) * x + 
+#       x_c ^ 2 + (b - y_c) ^ 2 - R ^ 2
+#
+# This a quadratic formula where:
+# A = 1 + m^2
+# B = 2 * (b * m - m * y_c - x_c) * x
+# C = x_c ^ 2 + (b - y_c) ^ 2 - R ^ 2
+#
+# Tangents occurs where 0 = B ^ 2 - 4 * A * C
+
+from sympy import solve, sqrt, symbols
+m, R, x_c, x_p, y_c, y_p = symbols('m, R, x_c, x_p, y_c, y_p')
+b = y_p - m * x_p
+A = 1 + m ** 2
+B = 2 * (b * m - m * y_c - x_c)
+C = x_c ** 2 + (b - y_c) ** 2 - R ** 2
+slopes = solve(B ** 2 - 4 * A * C, m)
+
+b = y_p - slopes[0] * x_p
+A = 1 + slopes[0] ** 2
+B = 2 * (b * slopes[0] - slopes[0] * y_c - x_c)
+C = x_c ** 2 + (b - y_c) ** 2 - R ** 2
+check1 = (B ** 2 - 4 * A * C).simplify()
+
+b = y_p - slopes[1] * x_p
+A = 1 + slopes[1] ** 2
+B = 2 * (b * slopes[1] - slopes[1] * y_c - x_c)
+C = x_c ** 2 + (b - y_c) ** 2 - R ** 2
+check2 = (B ** 2 - 4 * A * C).simplify()
+
+# There are two slopes, slopes[0] and slopes[1]
+# Now plug back into solve points of tangency
[…]

### 2019-12-10 r/00_run-python.R

@@ -1,0 +1,26 @@
+# The output from this script is included with the repo, so there is no need to rerun, but I have provided the code for transparency
+library(glue)
+library(reticulate)
+use_python('/anaconda3/bin/python', required = TRUE)
+source_python("python/tangents.py")
+
+get_tangent_slopes <-
+  glue("get_tangent_slopes <- function(R, x_c, y_c, x_p, y_p) {{
+       m1 <- {m1}
+       m2 <- {m2}
+       c(m1, m2)
+     }}", m1 = slopes[[1]], m2 = slopes[[2]]) 
+
+get_tangent_points <-
+  glue("get_tangent_points <- function(R, x_c, y_c, x_p, y_p) {{
+       x1 <- {x1}
+       x2 <- {x2}
+       y1 <- {y1}
+       y2 <- {y2}
+       data.frame(
+         x = c(x1, x2),
+         y = c(y1, y2)
+       )
+     }}", x1 = x1, x2 = x2, y1 = y1, y2 = y2) 
+
+cat(get_tangent_slopes, "\n", get_tangent_points, file = "r/functions-from-python.R")

### 2019-12-10 r/01_make-fig1-2.R

@@ -1,0 +1,330 @@
+source("r/header.R")
+
+# Figure 1 ----
+
+# Panel A: Equilateral triangular grid ----
+s1 <- 1      # interstomatal distance
+n <- 2
+s2 <- n * s1 # side length of hexagonal grid
+lim <- c(-s2, s2)
+
+df_stomata <- data.frame(
+  from = -s1 * c(seq(n / 2, n, 1 / 2), seq(n - 1/2, n / 2, -1 / 2)),
+  to = s1 * c(seq(n / 2, n, 1 / 2), seq(n - 1/2, n / 2, -1 / 2)),
+  length.out = c((n + 1):(2 * n + 1), (2 * n):(n + 1)),
+  n = sqrt(3) / 2 * seq(-n * s1, n * s1, s1)
+) %>%
+  pmap_dfr(~ {data.frame(
+      x = seq(..1, ..2, length.out = ..3),
+      y = rep(..4, ..3)
+  )}) %>%
+  mutate(image = "figures/open-stomate.png")
+
+xy <- combn(nrow(df_stomata), 2)
+df_grid <- bind_cols(
+  df_stomata[xy[1,], ],
+  df_stomata[xy[2,], ]
+) %>%
+  mutate(d = sqrt((x1 - x) ^ 2 + (y1 - y) ^ 2)) %>%
+  filter(abs(d - s1) < 1e-6)
+  
+pA <- ggplot(df_stomata, mapping = aes(x, y)) +
+  geom_segment(
+    data = df_grid,
+    mapping = aes(xend = x1, yend = y1),
+    color = "grey", size = 2,
+  ) +
+  geom_image(
+    mapping = aes(image = image),
+    size = 0.1
+  ) +
+  labs(title = "Stomata in grid of\nequilateral triangles") +
+  xlim(1.1 * lim) +
+  ylim(1.1 * lim) +
+  coord_equal() +
+  theme_void()
+
+pA
+
+# Panel B: Focal Equilateral Triangle ----
+
+df_stomata %<>% 
+  filter(x >= -s1 / 2, x <= s1 / 2, y <= 0, y >= -sqrt(3) / 2 * s1)
+xlim <- range(df_stomata$x) + c(-1, 1) * 0.15
+ylim <- range(df_stomata$y) + c(-1, 1) * 0.15
+
+pB <- ggplot(df_stomata, mapping = aes(x, y)) +
+  geom_polygon(color = "grey", fill = NA, size = 2) +
[…]

### 2019-12-10 r/02_spatially-explicit-model.R

@@ -1,0 +1,44 @@
+source("r/header.R")
+
+df <- crossing(
+  H = c(0, 0.01, 0.1),
+  D = set_units(10 ^ seq(1, 3.5, by = 0.025), 1 / mm ^ 2),
+  S = set_units(10 ^ seq(1, 3.5, by = 0.025), um ^ 2)
+) %>%
+  mutate(f_s = D * S) %>%
+  filter(f_s < set_units(1/3))
+
+# Calculate g_smax ----
+# Based on Monteith and Unsworth Table A.3 (pg. 376)
+D_wv <- set_units(24.9e-6, m ^ 2 / s)
+v <- set_units(2.24e-2, m^3 / mol)
+df %<>%
+  mutate(
+    b = biophysical_constant(D_wv, v),
+    m = morphological_constant(0.5, 0.5, 0.5),
+    # CHeck if sqrt(s) is right here
+    g_smax = set_units(b * m * D * sqrt(S), mol / m ^ 2 / s)
+  )
+
+# p_colonize(df$D[1], df$S[1], df$H[1])
+
+safe_pcolonize <- safely(p_colonize)
+tictoc::tic()
+plan(multiprocess)
+df$p_colonize <- df %>% 
+  select(D, S, H) %>%
+  furrr::future_pmap_dbl(~ {
+    res <- safe_pcolonize(..1, ..2, ..3)
+    if (is.null(res$result)) {
+      return(NA)
+    } else {
+      return(res$result)
+    }
+  }, .progress = TRUE)
+tictoc::toc()
+
+df %<>% 
+  filter(!is.na(p_colonize)) %>%
+  mutate_if(~ {inherits(.x, "units")}, drop_units)
+
+write_rds(df, "objects/spatially-explicit-results.rds")

### 2019-12-10 r/03_plot-spatially-explicit.R

@@ -1,0 +1,59 @@
+source("r/header.R")
+
+df <- read_rds("objects/spatially-explicit-results.rds")
+
+# Effect of f_s and gmax om p_colonize ----
+
+df1 <- df %>%
+  filter(D %in% c(10, 100, 1000)) %>%
+  mutate(H = glue("H = {H}", H = H))
+
+f3a <- ggplot(df1, aes(x = f_s, y = p_colonize, linetype = as.character(D))) +
+  facet_grid(H ~ .) +
+  geom_line(size = 1.1, lineend = "round") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  xlab(expression(paste("Stomatal cover (", italic(f)[s], ")"))) +
+  ylab(expression(paste("Probability of colonization (", italic(p)[colonize], ")"))) +
+  scale_linetype_manual(
+    values = 3:1, 
+    name = expression(paste("Stomatal density [m", m^-2, "]"))
+  ) +
+  theme_bw() +
+  theme(
+    axis.text = element_text(size = 10),
+    axis.title = element_text(size = 12),
+    axis.title.x = element_text(margin = margin(0.3, 0, 0, 0, unit = "cm")),
+    legend.position = "top",
+    strip.background = element_rect(fill = "white"),
+    strip.text = element_text(size = 10)
+  )
+
+l <- get_legend(f3a)
+
+f3a <- f3a + theme(legend.position = "none")
+
+df2 <- df1 %>%
+  filter(g_smax < 5)
+
+f3b <- ggplot(df2, aes(x = g_smax, y = p_colonize, linetype = as.character(D))) +
+  facet_wrap(H ~ ., strip.position = "left", ncol = 1) +
+  geom_line(size = 1.1, lineend = "round") +
+  xlab(expression(paste("Stomatal conductance (", italic(g)[list(s,max)], ") [mol ", m^-2~s^-1, "]"))) +
+  ylab(expression(paste("Probability of colonization (", italic(p)[colonize], ")"))) +
+  scale_linetype_manual(values = 3:1) +
+  scale_y_continuous(position = "right") +
+  theme_bw() +
+  theme(
+    axis.text = element_text(size = 10),
+    axis.title = element_text(size = 12),
+    legend.position = "none",
+    strip.background = element_rect(fill = "white"),
+    strip.text = element_text(size = 10)
+  )
+
+f3b
+
[…]

### 2019-12-10 r/04_find-scaling.R

@@ -1,0 +1,11 @@
+source("r/header.R")
+
+df <- crossing(
+  p_colonize = c(0.025, 0.05, 0.1, 0.2, 0.4),
+  D = set_units(10 ^ seq(1, 3.5, by = 0.025), 1 / mm ^ 2),
+  H = c(0, 0.01, 0.1)
+) %>%
+  rowwise() %>%
+  mutate(S = find_S(p_colonize, D, H))
+
+write_rds(df, "objects/scaling-results.rds")

### 2019-12-10 r/05_plot-scaling.R

@@ -1,0 +1,56 @@
+source("r/header.R")
+
+df <- read_rds("objects/scaling-results.rds") %>%
+  ungroup() %>%
+  mutate(x = D * S) %>%
+  mutate_if(~ {inherits(.x, "units")}, drop_units) %>%
+  mutate(H = glue("H = {H}", H = H))
+
+df_refline <- df %>%
+  filter(p_colonize == median(p_colonize)) %>%
+  group_by(H) %>%
+  summarize(min_D = min(D), max_D = max(D), D = median(D), S = median(S)) %>%
+  mutate(d = log(D), s = log(S)) %>%
+  crossing(slope = c(-1, -2)) %>%
+  mutate(
+    intercept = s - slope * d,
+         min_S = exp(intercept + slope * log(min_D)),
+         max_S = exp(intercept + slope * log(max_D))
+    ) %>%
+  select(H, D = min_D, xend = max_D, S = min_S, yend = max_S) %>%
+  mutate(p_colonize = NA)
+
+df_labels <- df %>%
+  filter(p_colonize %in% c(0.025, 0.1, 0.4)) %>%
+  group_by(H, p_colonize) %>%
+  filter(D == min(D)) 
+
+fig4 <- ggplot(df, aes(D, S, group = p_colonize)) +
+  facet_wrap(H ~ .) +
+  geom_segment(
+    data = df_refline, 
+    mapping = aes(xend = xend, yend = yend),
+    linetype = "dashed", color = "grey", size = 1.1
+  ) +
+  geom_line() +
+  geom_text(
+    data = df_labels,
+    aes(label = p_colonize),
+    hjust = 1.1, vjust = 0.5
+  ) +
+  scale_x_log10(breaks = 10 ^ (1:3), limits = c(1, 3500)) +
+  scale_y_log10(breaks = 10 ^ (1:3)) +
+  xlab(expression(paste("Stomatal density [m", m^-2, "]"))) +
+  ylab(expression(paste("Stomatal size [", mu, m^2, "]"))) +
+  theme_bw() +
+  theme(
+    axis.text = element_text(size = 10),
+    axis.title = element_text(size = 12),
+    panel.grid = element_blank(),
+    strip.background = element_rect(fill = "white"),
+    strip.text = element_text(size = 10)
+  )
+
+fig4
+
+ggsave("figures/fig4.pdf", w = 6.5, h = 3, units = "in")

### 2019-12-10 r/functions-from-python.R

@@ -1,0 +1,15 @@
+get_tangent_slopes <- function(R, x_c, y_c, x_p, y_p) {
+  m1 <- (-R*sqrt(-R**2 + x_c**2 - 2*x_c*x_p + x_p**2 + y_c**2 - 2*y_c*y_p + y_p**2) - x_c*y_c + x_c*y_p + x_p*y_c - x_p*y_p)/(R**2 - x_c**2 + 2*x_c*x_p - x_p**2)
+  m2 <- (R*sqrt(-R**2 + x_c**2 - 2*x_c*x_p + x_p**2 + y_c**2 - 2*y_c*y_p + y_p**2) - x_c*y_c + x_c*y_p + x_p*y_c - x_p*y_p)/(R**2 - x_c**2 + 2*x_c*x_p - x_p**2)
+  c(m1, m2)
+} 
+ get_tangent_points <- function(R, x_c, y_c, x_p, y_p) {
+  x1 <- (x_c*(R**2 - x_c**2 + 2*x_c*x_p - x_p**2)**2 - y_c*(R**2 - x_c**2 + 2*x_c*x_p - x_p**2)*(R*sqrt(-R**2 + x_c**2 - 2*x_c*x_p + x_p**2 + y_c**2 - 2*y_c*y_p + y_p**2) + x_c*y_c - x_c*y_p - x_p*y_c + x_p*y_p) - (x_p*(-R*sqrt(-R**2 + x_c**2 - 2*x_c*x_p + x_p**2 + y_c**2 - 2*y_c*y_p + y_p**2) - x_c*y_c + x_c*y_p + x_p*y_c - x_p*y_p) - y_p*(R**2 - x_c**2 + 2*x_c*x_p - x_p**2))*(R*sqrt(-R**2 + x_c**2 - 2*x_c*x_p + x_p**2 + y_c**2 - 2*y_c*y_p + y_p**2) + x_c*y_c - x_c*y_p - x_p*y_c + x_p*y_p))/((R**2 - x_c**2 + 2*x_c*x_p - x_p**2)**2 + (R*sqrt(-R**2 + x_c**2 - 2*x_c*x_p + x_p**2 + y_c**2 - 2*y_c*y_p + y_p**2) + x_c*y_c - x_c*y_p - x_p*y_c + x_p*y_p)**2)
[…]

### 2019-12-10 r/functions.R

@@ -1,0 +1,435 @@
+# Calculations of gsmax follow Sack and Buckley 2016 ----
+biophysical_constant <- function(D_wv, v) set_units(D_wv / v, mol / m / s)
+
+morphological_constant <- function(c, h, j) {
+  set_units((pi * c ^ 2) / (j ^ 0.5 * (4 * h * j + pi * c)), 1)
+} 
+
+gmax <- function(b, m, d, s) {
+  set_units(b * m * d * sqrt(s), mol / m ^ 2 / s)
+}
+
+# Calculate D_w as a function temperature and pressure
+get_Dw <- function(D_0, Temp, eT, P) {
+  
+  D_0 %<>%
+    set_units(m ^ 2 / s) %>%
+    drop_units()
+  
+  Temp %<>%
+    set_units(K) %>%
+    drop_units()
+  
+  P %<>%
+    set_units(kPa) %>%
+    drop_units()
+  
+  set_units(D_0 * (Temp / 273.15) ^ eT * (101.3246 / P), m ^ 2 / s)
+  
+}
+
+# Functions for converting from D to U, R to S ----
+D2U <- function(D) {
+
+  # D must have units of 1 / mm ^ 2
+  # U = 2 * Rhex where Rhex is major radius of a regular hexagon
+  # Ahex = 3 * sqrt(3) / 2 * Rhex ^ 2
+  Ahex <- 1 / D
+  Rhex <- sqrt(2 * Ahex / (3 * sqrt(3)))
+  U <- 2 * Rhex # in mm
+  1e3 * U # in um
+
+}
+
+U2D <- function(U) {
+
+  # U must have units of um
+  Rhex <- U / 2
+  Ahex <- 3 * sqrt(3) / 2 * Rhex ^ 2
+  D <- 1 / Ahex # 1 / um ^ 2
+  D * 1e6 # 1 / mm ^ 2
+
+}
+
+S2R <- function(S) {
+  
+  # S must be in units of um ^ 2
+  sqrt(S / (2 * pi)) # units of um
+  
+}
+
+R2S <- function(R) {
+  
+  # r must be in units of um
+  2 * pi * R ^ 2 # units of um ^ 2
+  
+}
+
+# Functions for numerical integration ----
+# Minimum and maximum possible positions for y_pathogen, given x_pathogen, r, ue
+get_ymin <- function(x, R, U) {
+  
+  stopifnot(R < 2 * U)
[…]

### 2019-12-10 r/header.R

@@ -1,0 +1,26 @@
+rm(list = ls())
+graphics.off()
+
+# Libraries
+library(cowplot)
+library(dplyr)
+library(furrr)
+library(ggforce)
+library(ggimage)
+library(gginnards)
+library(ggplot2)
+library(glue)
+library(magrittr)
+library(pracma)
+library(purrr)
+library(readr)
+library(stringr)
+library(tikzDevice)
+library(tidyr)
+library(units)
+
+source("r/functions.R")
+source("r/functions-from-python.R")
+
+palette(colorRampPalette(c("tomato", "steelblue"), alpha = TRUE)(5))
+

### 2019-12-10 r/install-packages.R

@@ -1,0 +1,21 @@
+# Packages to generate manuscript
+install.packages("rmarkdown")
+install.packages("rticles")
+
+# Packages for running R scripts
+install.packages("cowplot")
+install.packages("dplyr")
+install.packages("furrr")
+install.packages("ggforce")
+install.packages("ggimage")
+install.packages("gginnards")
+install.packages("ggplot2")
+install.packages("glue")
+install.packages("magrittr")
+install.packages("pracma")
+install.packages("purrr")
+install.packages("readr")
+install.packages("stringr")
+install.packages("tikzDevice")
+install.packages("tidyr")
+install.packages("units")

### 2019-12-10 ms/ms.Rmd
version submitted to Frontiers

@@
 Stomatal pores control both leaf gas exchange and are an entry for many plant pathogens, setting up the potential for tradeoffs between photosynthesis and defense.
-To prevent colonization, plants close their stomata after recognizing pathogens to limit infection.
-Anatmoical shifts to lower stomatal density and/or size may also limit pathogen colonization, but would permanently reduce the gas exchange capacity for the life of a leaf.
-I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as function of stomatal size and density.
-The model predicts that changes in stomatal density or size increase the probability of host plant colonization most when stomatal cover is low.
-I derived scaling relationships between stomatal size and density that preserves a given probability of colonization.
+To prevent colonization and limit infection, plants close their stomata after recognizing pathogens.
+In addition to closing stomata, anatmoical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes would permanently reduce the gas exchange capacity for the life of the leaf.
+I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as a function of stomatal size and density, anatomical traits which determine maximum rates of gas exchange.
+The model predicts greater stomatal size or density increases colonization, but the effect is most pronounced when stomatal cover is low.
+I also derived scaling relationships between stomatal size and density that preserves a given probability of colonization.
 These scaling relationships set up a potential conflict between maximizing defense and minimizing stomatal cover.
-To my knowledge, this is first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy.
+To my knowledge, this is the first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy.
 It makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants.
-The model is generalizable to many types of pathogens, but lacks significant biological realism that may need to be added in future extensions.
+The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for accurate predictions.
 
@@
 <!-- overview -->
-Stomata evolved to let plants regulate gas exchange in and out of the leaf [@hetherington_role_2003; @berry_stomata:_2010; @chater_origins_2017], but many plant pathogens take advantage of these chinks in the leaf cuticular armor to infect prospective hosts [@zeng_plant_2010; @mclachlan_gate_2014; @melotto_stomatal_2017].
+Stomata evolved to regulate gas exchange in and out of the leaf [@hetherington_role_2003; @berry_stomata:_2010; @chater_origins_2017], but many plant pathogens take advantage of these chinks in the leaf cuticular armor to infect prospective hosts [@zeng_plant_2010; @mclachlan_gate_2014; @melotto_stomatal_2017].
 The density and size of stomata set the anatomical maximum rate of stomatal conductance to CO$_2$ and water vapor [@brown_static_1900; @parlange_stomatal_1970; @franks_effect_2001; @franks_maximum_2009; @lehmann_effects_2015; @sack_developmental_2016; @harrison_influence_2019], but the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019].
-Many plant pathogens, including viruses [@murray_plant_2016], bacteria [@melotto_plant_2006; @underwood_role_2007], protists [@fawke_oomycete_2015], and fungi [@hoch_signaling_1987; @zeng_plant_2010] use stomatal pores to gain entry to the leaf.
+Many plant pathogens, including viruses [@murray_plant_2016], bacteria [@melotto_plant_2006; @underwood_role_2007], protists [@fawke_oomycete_2015], and fungi [@hoch_signaling_1987; @zeng_plant_2010] use stomatal pores to gain entry into the leaf.
 Since stomatal conductance is a major limitation on photosynthesis [@farquhar_stomatal_1982; @jones_partitioning_1985] while pathogens reduce fitness, this sets up a potential tradeoff between increased photosynthesis and defense against pathogens.
 Although there have been many empirical studies on the effect that pathogens have on stomata, there is no theoretical framework in which to place these findings.
-Lack of theoreatical framework makes it difficult to answer general questions about how selection for pathogen defense constrains maximum rates of gas exchange.
+Lack of a theoreatical framework makes it difficult to answer general questions about how selection for pathogen defense constrains maximum rates of gas exchange.
 
@@
 Stomatal anatomy is the key link between gas exchange and pathogen colonization.
-The density and size of stomata not only determines the theoretical maximum stomatal conductance (\gsmax; @harrison_influence_2019), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019].
+The density and size of stomata not only determines the theoretical maximum stomatal conductance (\gsmax), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019].
 Therefore, I use anatomical \gsmax{} as a proxy for \gsop{} and do not address dynamic changes in stomatal aperture (see [Discussion]).
@@
 Anatomy may be another layer of defense; plants can reduce pathogen colonization by developing leaves with lower stomatal density and/or size.
-Infection increases in leaves with higher stomatal density [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019], which suggests that both anatomy and physiological responses (stomatal closure) affect host colonization.
+Infection increases in leaves with higher stomatal density [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019], which suggests that both anatomical and physiological responses (stomatal closure) affect host colonization.
 By the same logic, if stomatal density were held constant, larger stomata should also increase colonization because they occupy more area.
-One key difference between physiological and anatomical defenses is that stomatal closure reduces gas exchange transiently, whereas anatomy constrains gas exchange throught the life of a leaf.
-Because we do not understand the relationship between stomatal anatomy and colonization well, we do not know the relationship between \gsmax{} and infection.
+One key difference between physiological and anatomical defenses is that stomatal closure reduces gas exchange transiently, whereas anatomy constrains gas exchange throught the life of the leaf.
+Because we do not understand the relationship between stomatal anatomy and colonization well, we cannot predict the relationship between \gsmax{} and infection.
 
@@
 If stomatal size and density affect pathogen colonization, selection to limit colonization may shape stomatal size-density scaling relationships.
-Botanists have long recognized that stomatal size and density are inversely correlated [@weiss_untersuchungen_1865; @ticha_photosynthetic_1982; @hetherington_role_2003; @sack_hydrology_2003; @franks_co_2009; brodribb_unified_2013; @de_boer_optimal_2016], but the evolutionary origin of this relationship is not yet known.
+Botanists have long recognized that stomatal size and density are inversely correlated [@weiss_untersuchungen_1865; @ticha_photosynthetic_1982; @hetherington_role_2003; @sack_hydrology_2003; @franks_co_2009; @brodribb_unified_2013; @de_boer_optimal_2016], but the evolutionary origin of this relationship is not yet known.
 Here I argue that pathogens could shape selection on this relationship.
-Explanations for inverse size-density scaling are usually cast in terms of preserving \gsmax{} and/or total epidermal area allocated to stomata [@de_boer_optimal_2016] because there are many combinations of stomatal size and density that have same \gsmax{} or same \fs{}:
+Explanations for inverse size-density scaling are usually cast in terms of preserving \gsmax{} and/or total epidermal area allocated to stomata [\fs; @de_boer_optimal_2016] because there are many combinations of stomatal size and density that have same \gsmax{} or same \fs{}:
 
@@
 
-$b$ and $m$ are assumed to be biophysical and morphological constants, *sensu* @sack_developmental_2016 (see [Supplementary Material].
+$b$ and $m$ are assumed to be biophysical and morphological constants, *sensu* @sack_developmental_2016 (see [Supplementary Material]).
 If size and density also affect pathogen colonization, then selection from foliar pathogens could significantly alter the size-density scaling relationship.
@@
 This would change the intercept $\alpha$ but not the slope.
-The effect of foliar pathogens on the slope depends on the relationship between size, density, and probability of colonization If the probability of colonization is proportional to the product of *linear* stomatal size ($S ^ {0.5}$) and density ($\propto D S ^ {0.5}$ as for \gsmax) then it has the same effect on the slope as \gsmax{} because there are many combinations of $D$ and $S ^ {0.5}$ that have same probability of colonization.
+The effect of foliar pathogens on the slope depends on the relationship between size, density, and probability of colonization.
+If the probability of colonization is proportional to the product of *linear* stomatal size ($S ^ {0.5}$) and density ($\propto D S ^ {0.5}$ as for \gsmax) then it has the same effect on the slope as \gsmax{} because there are many combinations of $D$ and $S ^ {0.5}$ that have same probability of colonization.
 If the probability of colonization is proportional to the product of *areal* stomatal size ($S$) and density ($\propto D S$ as for \fs) then it has the same effect on the slope as \fs{} because there are many combinations of $D$ and $S$ that have same probability of colonization.
 Alternatively, the probability of colonization may have a different scaling relationsip (neither 0.5 nor 1) or may be nonlinear on a log-log scale.
-Unlike \gsmax{} and \fs{}, we do not have theory which predicts a stomatal size-density relationship which preserves the probability of colonization.
+Unlike \gsmax{} and \fs{}, we do not have theory to predict a stomatal size-density relationship that preserves the probability of colonization.
 
@@
 The inverse stomatal size-density relationship has usually been explained in terms of preserving stomatal conductance and/or stomatal cover, but selection by pathogens might alter scaling.
-To address these gaps, the goals of this study are to 1) introduce a spatially explicit, probabalistic model pathogen colonization on the leaf surface; 2) use the model to predict the relationship between \gsmax, \fs{}, and the probability of colonization; 3) work out what these relationships predict about stomatal size-density scaling.
+To address these gaps, the goals of this study are to 1) introduce a spatially explicit model pathogen colonization on the leaf surface; 2) use the model to predict the relationship between \gsmax, \fs{}, and the probability of colonization; 3) work out what these relationships predict about stomatal size-density scaling.
 
@@
 
-In this section, I introduce a spatially explicity, probabilistic model of pathogen colonization on a leaf surface.
-I explain the model structure and assumptions.
-The [Methods] section below describes how I analyzed the model to address the goals of the study.
-I used Sympy version 1.3 [@meurer_sympy:_2017] for symbolic derivations.
+In this section, I introduce a spatially explicit model of pathogen colonization on a leaf surface.
+I explain the model structure and assumptions here; the [Materials and Methods] section below describes how I analyzed the model to address the goals of the study.
 For generality, I refer to a generic ``pathogen'' that lands on leaf and moves to a stomate.
@@
 These very different biotropic movements on the leaf are treated identically here.
+I used Sympy version 1.3 [@meurer_sympy:_2017] for symbolic derivations.
 
@@
  Symbol | R      | Units  | Description
---------------------|-----------|-----------------------|-----------------------
+--------|--------|--------|-----------------------------------------------------
  $D,d$              | `D, d`    | $\textrm{mm}^{-2}$    | stomatal density ($d = \text{log}~D$)
@@
  $g_\text{s,op}$    | `g_sop`   | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | operational stomatal conductance 
- $L$                | `L`       | $\mu\textrm{m}$       | guard cell length ($L = 2 R$)
+ $H$                | `H`       | $\mu\textrm{m} ^{-1}$ | death rate of pathogen on leaf surface
  $R$                | `R`       | $\mu\textrm{m}$       | stomatal radius ($S = 2 \pi R ^ 2$)
  $S,s$              | `S, s`    | $\mu\textrm{m}^2$     | stomatal size ($s = \text{log}~S$)
+ $\theta_i$         | `theta_i` | radians               | angles between pathogen ($x_p,y_p$) and lines tangent to the circumfrence of stomate $i$
  $U$                | `U`       | $\mu\textrm{m}$       | interstomatal distance
- $v_i$              | `v_i`     | $\mu\textrm{m}$       | distance between pathogen and stomate $i$
- $\theta_i$         | `theta_i` | radians               | angles between $x_p,y_p$ and stomate $i$
- $H$                | `H`       | $\mu\textrm{m} ^{-1}$ | death rate of pathogen on leaf surface
+ $v_i$              | `v_i`     | $\mu\textrm{m}$       | distance between pathogen ($x_p,y_p$) and stomate $i$
+ $x_i,y_i$          | `x_i,y_i` | $\mu\textrm{m}$       | position of stomate $i$
+ $x_p,y_p$          | `x_p,y_p` | $\mu\textrm{m}$       | starting position of pathogen
 
@@
 
-Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015; earlier citations of stomatal interference].
+Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015].
 Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$.
-This assumption ignore veins, trichomes, and within-leaf variation in stomatal density.
+This assumption ignores veins, trichomes, and within-leaf variation in stomatal density.
 Stomata are therefore arrayed in an evenly spaced grid (Figure \ref{fig:fig1}a).
@@
 This may be approximately true for fully open stomata with kidney-shaped guard cells.
+Although I assume stomata are circular here, in calculating \gsmax, I assume typical allometric relationships between length, width, and pore area  (see [Supplementary Material]).
 
@@
 Stomata are assumed to be in a homogenous equilateral triangular grid, which means that we can extrapolate from \textbf{b.} a focal triangle to the entire leaf.
-The circle indicate idealized stomata; the grey lines between them are for visualization.
+The circles represent idealized stomata; the grey lines between them are for visualization.
 \textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the area.
-\textbf{d.} The model assumes that a pathogen depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves in a certain angle, depicted by the grey polygons.}
+\textbf{d.} The model assumes that a pathogen, depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves at the correct angle, depicted by the grey polygons.}
 	\label{fig:fig1}
@@
 
-Now imagine that a pathogen lands at a random position on the leaf and must arrive at a stomate to colonize.
+Now imagine that a pathogen lands at a uniform random position within the focal region and must arrive at a stomate to colonize.
 If it lands on a stomate, then it infects the leaf with probability 1; if it lands between stomata, then it infects the leaf with probability $p_\text{locate}$.
 This is the probability that it locates a stomate, which I will derive below.
-The probabilities of landing on and between a stomate are \fs{} and $1 - $ \fs{}, respectively.
-Hence, the total probability of infection is:
+The probabilities of landing on or between a stomate are \fs{} and $1 - f_S$, respectively.
+Hence, the total probability of colonization is:
 
@@
 I assume that the pathogen cannot sense where stomata are and orients at random, thereafter traveling in that direction.
-If it successfully locates a stomate it colonizes the leaf, but otherwise does not infect.
+If it successfully locates a stomate, it colonizes the leaf, but otherwise does not infect.
 If there is a high density of stomata and/or large stomata, the probability of locating a stomate increases.
@@
 Further, since an equilateral triangle can be broken up into six identical units (Figure \ref{fig:fig1}c), we can simply calculate $p_\text{locate}$ in this focal area.
+This implicitly assumes that the probability of colonzing stomata outside the focal area is 0 because they are too far away.
 
 Imagine that the pathogens lands in position ($x_p,y_p$) within the triangle.
-The centroid of the triangle is at position ($x_c,y_c$) and a reference stomate is at position $(0,0)$ (Figure \ref{fig:fig2}).
+The centroid of the triangle is at position ($x_c,y_c$) and a reference stomate is at position $(0,0)$ (Figure \ref{fig:fig2}a).
 Therefore $x_c = U / 2$ and $y_c = \sqrt{3} U / 6$.
@@
 
-Given that the pathogens starts at position ($x_p,y_p$), what's the probability of contacting one of the stomata at the vertices of the focal triangle?
+Given that the pathogen starts at position ($x_p,y_p$), what's the probability of contacting one of the stomata at the vertices of the focal triangle?
 I assume the probability of contacting a stomate is equal to the proportion of angular directions that lead to a stomate (Figure \ref{fig:fig1}d).
-I solved this by finding the angles ($\theta_1, \theta_2, \theta_3$) between lines that are tangent to the outside of the three stomata and pass through ($x_p,y_p$) (Figure \ref{fig:fig2}).
-If stomata $i$ is centered at ($x_i,y_i$), the two slopes of tangency as function of pathogen position are:
+I solved this by finding the angles ($\theta_1, \theta_2, \theta_3$) between lines that are tangent to the outside of the three stomata and pass through ($x_p,y_p$) (Figure \ref{fig:fig2}a).
+If stomate $i$ is centered at ($x_i,y_i$), the two slopes of tangency as function of pathogen position are:
 
@@
 Within the focal triangle, the reference stomata is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$.
-\textbf{b .} Within the focal region, a pathogen can land within the stomate (white cirlce with grey outline and radius $R$) or in the grey area.
+\textbf{b.} Within the focal region, a pathogen can land within the stomate (white cirlce with grey outline and radius $R$) or in the grey area.
 The outer borders of this area are shown and depend on $R$ and $U$.
@@
 For example, if stomata are at very low density, then a pathogen may die before it reaches a stomate because of UV, dessication, or another factor.
-I included this effect by assuming the probability of reaching a stomata declines exponentially at rate $H$ with the Euclidean distance $v_i(x_p,y_p)$ between the pathogen location and the edge of stomata $i$, which is distance $R$ from its center at $x_i,y_i$:
+I included this effect by assuming the probability of reaching a stomate declines exponentially at rate $H$ with the Euclidean distance $v_i(x_p,y_p)$ between the pathogen location and the edge of stomata $i$, which is distance $R$ from its center at $x_i,y_i$:
 
 \begin{equation}
-  v_i = \sqrt{(x_i - x_p) ^ 2 + (y_i - y_p) ^ 2} - r.
+  v_i(x_p,y_p) = \sqrt{(x_i - x_p) ^ 2 + (y_i - y_p) ^ 2} - R.
 \end{equation}
@@
 To obtain the average $p_\text{locate}$, we must integrate $f_\text{locate}(x_p,y_p)$ over all possible starting positions $(x_p,y_p)$ within the focal area.
-The focal area is a 30-60-90 triangle with vertices at the reference stomate $(0,0)$, the midpoint of baseline $(U/2,0)$, and the centroid of the focal trinagle $(U/2, \sqrt{3} / 6 U)$ (Figure \ref{fig:fig1}c).
-Colonization occurs if the pathogen lands in the reference stomate, so we need to integrate from the edge of this stomate, from $\sqrt{3} / 2 R$ to $U/2$ (Figure \ref{fig:fig2}b).
+The focal area is a 30-60-90 triangle with vertices at the center of the reference stomate $(0,0)$, the midpoint of baseline $(U/2,0)$, and the centroid of the focal trinagle $(U/2, \sqrt{3} / 6 U)$ (Figure \ref{fig:fig1}c).
+Colonization occurs with probability 1 if the pathogen lands in the reference stomate, so we need to integrate the probability of colonization if it lands elsewhere.
+This region extends from the edge of the stomate, at $\sqrt{3} / 2 R$ to $U/2$ (Figure \ref{fig:fig2}b).
 At any $x$, we integrate from the bottom of the focal area ($y_\text{min}$) to the top ($y_\text{max}$):
@@
 \begin{equation}
-  p_\text{locate} = \int\limits_{\frac{\sqrt{3}}{2} R} ^ {U / 2} \int\limits_{y_\mathrm{min}} ^ {y_\mathrm{max}} f_\text{locate}(x, y) \,dx \,dy
+  p_\text{locate} = \frac{1}{a_\text{focal}}\int\limits_{\frac{\sqrt{3}}{2} R} ^ {U / 2} \int\limits_{y_\mathrm{min}} ^ {y_\mathrm{max}} f_\text{locate}(x, y) \,dx \,dy
 \end{equation}
+
+$a_\text{focal}$ is the area of the focal region depicted in grey in Figure \ref{fig:fig2}b:
+
+$$ a_\text{focal} = \frac{U ^ 2}{8 \sqrt{3}} - \frac{\pi R ^ 2}{12} $$
 
@@
 I solved $p_\text{colonize}$ using the `integral2()` function in the **pracma** package version 2.2.5 [@borchers_pracma:_2019] for numerical integration.
-I used R version 3.6.1 [@r_core_team_r:_2019] for all analyses and the paper using **rmarkdown** version 1.17 [@xie_r_2018; @allaire_rmarkdown:_2019].
+I used R version 3.6.1 [@r_core_team_r:_2019] for all analyses and wrote the paper in **rmarkdown** version 1.17 [@xie_r_2018; @allaire_rmarkdown:_2019].
 Source code is deposited on GitHub (https://github.com/cdmuir/stomata-tradeoff) and will be arcived on Zenodo upon publication.
 
-## What is the relationship between stomatal size, density, and the probability of colonization? {-}
+## What is the relationship between stomatal size, density, and colonization? {-}
 
 I calculated $p_\text{colonize}$ over a biologically plausible grid of stomatal size and density for hypostomatous species based on @de_boer_optimal_2016.
-Stomatal density ranges from $10^1 -- 10^{3.5}~\text{mm}^{-2}$; stomatal size ranges from $10^1 -- 10^{3.5}~\mu\text{m}^2$.
-I only considered combinations of size and density where \fs{} was less than 0.5.
+Stomatal density ranges from $10^1 - 10^{3.5}~\text{mm}^{-2}$; stomatal size ranges from $10^1 - 10^{3.5}~\mu\text{m}^2$.
+I only considered combinations of size and density where \fs{} was less than 1/3.
 For simplicity, I have not extended the current analysis to amphistomatous leaves.
@@
 $$ d = c_g - 0.5 s $$
-where lowercase variables are log-transformed equivalents of their uppercase counterparts. To preseve $C_g$, the scaling exponent $\beta_g = 0.5$.
+where lowercase variables are log-transformed equivalents of their uppercase counterparts. The scaling exponent $\beta_g = 0.5$ preseves $C_g$.
 
@@
 If $\beta_p = 0.5$, then $p_\text{colonize}$ is always proportional to \gsmax.
-If $\beta_p < 0.5$, small, densely packed stomata would be better defended (lower $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax.
-If $\beta_p > 0.5$, small, densely packed stomata would be less defended (higher $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax.
+If $\beta_p > 0.5$, small, densely packed stomata would be better defended (lower $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax.
+If $\beta_p < 0.5$, small, densely packed stomata would be less defended (higher $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax.
 I refer to the three outcomes ($\beta_p = 0.5$, $\beta_p < 0.5$, and $\beta_p > 0.5$) as iso-, hypo-, and hyper-conductance, respectively.
@@
 The probability of colonization ($p_\text{colonize}$) is not simply a one-to-one relationship between the fraction of epidermal area allocated to stomata (\fs).
-At low \fs, $p_\text{colonize}$ is much greater than \fs{} because a pathogen can either land on a stomata with probability \fs{} or locate a stomate after land, which must be added to \fs{} (Equation \ref{eq:p_colonize}); Figure \ref{fig:fig3}a).
-As greater \fs{} and $p_\text{colonize}$ have a linear relationship.
+At low \fs, $p_\text{colonize}$ increases faster rapdily relative to \fs{} at first Figure \ref{fig:fig3}a).
+At higher \fs, the $p_\text{colonize}$ increases linearly with \fs.
 When $H = 0$, any combination of stomatal size ($S$) and density ($D$) with the same \fs{} have the same effect on $p_\text{colonize}$.
@@
 When $H > 0$, this extra distance leads more pathogens to die before they can find a stomate.
-In contrast to \fs, $p_\text{colonize}$ increases at a greater than linear rate with \gsmax{} and greater $D$ (smaller $S$) is associated with lower $p_\text{colonize}$ when \gsmax{} is constant (Figure \ref{fig:fig3}b).
-This happens because $p_\text{colonize}$ increases approximately linearly with $S$ whereas \gsmax{} is proportional to $S^{0.5}$
+In contrast to \fs, $p_\text{colonize}$ increases at a greater than linear rate with \gsmax.
+Greater $D$ (smaller $S$) is associated with lower $p_\text{colonize}$ when \gsmax{} is held constant (Figure \ref{fig:fig3}b).
+This happens because $p_\text{colonize}$ increases approximately linearly with $S$ whereas \gsmax{} is proportional to $S^{0.5}$.
 
@@
 Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$).
-The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density when $H = 0$.
-When $H > 0$, higher density (solid lines) increase $p_\text{colonize}$.
-\textbf{b.} $p_\text{colonize}$ increases expoential with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ decreases at higher density for a given \gsmax.}
-
+The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density when $H = 0$ (upper facet).
+When $H > 0$, higher density (solid lines) increase $p_\text{colonize}$ (lower facets).
+\textbf{b.} $p_\text{colonize}$ increases expoentially with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is much lower at higher densities for a given \gsmax.
+The relationship between \gsmax{} and $p_\text{colonize}$ is similar for all values of $H$.}
 	\label{fig:fig3}
@@
 
-The scaling relationship between $S$ and $D$ that preserves $p_\text{colonize}$ is always greater 0.5 (hyper-condtance), but usually less than 1.
+The scaling relationship between $S$ and $D$ that preserves $p_\text{colonize}$ is always greater 0.5 (hyper-conductance), but usually less than 1.
 When $H = 0$, the scaling relationship is essentially 1 (Figure \ref{fig:fig4}), which means that an increase \fs{} leads to a proportional increase in $p_\text{colonize}$.
@@
 In each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.025 (lowest line), 0.05, 0.1, 0.2, or 0.4 (highest line).
-For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.6$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region.
+For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.5$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region.
 When the death rate on the leaf surface is low ($H = 0$), the scaling exponent is very close to $\beta = 1$.
@@
 To develop testable predictions, we need mathematical models that can clarify the potential for tradeoffs between stomatal conductance, stomatal cover, and disease resistance.
-I used a spatially explicit model of a pathogen searching for a stomate to colonize the host to derive predictions about the relationship between stomatal anatomy and disease resistance for the first time.
+I used a spatially explicit model of a pathogen searching for a stomate to colonize a host.
+From this [Model], I derived predictions about the relationship between stomatal anatomy and disease resistance for the first time.
 The model predicts that the probability of colonization is not always proportional to the surface area of leaf covered by stomata (\fs), as one might intuitively predict.
@@
 The model reveals the potential for conflicting demands of maximizing disease resistance, minimizing stomatal cover, and maintaining stomatal conductance.
-Including the effect of anatomy on disease resistance has the potential to change our understading of how stomatal size-density scaling evolves in land plants.
+Including the effect of anatomy on disease resistance therefore has the potential to change our understading of how stomatal size-density scaling evolves in land plants.
 
@@
 drier, higher UV) should therefore reduce the effect of increased size or density as low \fs.
-If true, the dimishing marginal effect of \fs{} on colonization could explain why stomatal ration on the upper and lower surface is bimodal [@muir_making_2015].
+If true, the dimishing marginal effect of \fs{} on colonization could explain why stomatal ratio on the upper and lower surface is bimodal [@muir_making_2015].
 The initial cost of adaxial (upper) stomata is high, but if the benefits outweigh the costs, then equal stomatal densities on each surface maximize CO$_2$ supply for photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990].
@@
 <!-- paragraph on scaling -->
-An effect stomatal size and density on susceptibility to foliar pathogens could profoundly change our understanding of stomatal size-density scaling.
+An effect of stomatal size and density on susceptibility to foliar pathogens could change our understanding of stomatal size-density scaling.
 Since allocating leaf epidermis to stomata may be costly [@franks_mechanical_2007; @assmann_guard_1987; @dow_physiological_2014; @lehmann_effects_2015; @baresch_competition_2019], selection should favor leaves that achieve a desired \gsmax{} while minimizing \fs{} [@de_boer_optimal_2016].
@@
 The optimal stomatal size and density will therefore depend on the precise costs and benefits of infection, stomatal conductance, and stomatal cover.
-This may explain why many leaves have large, sparsely packed stomata despite the fact that they could achieve the same \gsmax{} and lower \fs{} with smaller, more densely packed stomata
+This may explain why many leaves have large, sparsely packed stomata despite the fact that they could achieve the same \gsmax{} and lower \fs{} with smaller, more densely packed stomata.
 
 <!-- scope of model -->
-The model examines the probability of colonization for a single pathogen and is most applicable to diseases where the host has some resistance.
+The model examines the probability of colonization for a single pathogen.
+The calculated probabilities of colonization should not be interpreted as exact predictions, but rather as depicting qualitative relationships between stomatal anatomy and infection severity.
+The model is most applicable to diseases where the host has some resistance.
 The energetic cost and lost photosynthetic capacity (closed stomata, necrosis, etc.) of dealing with a pathogen is assumed to be proportional to the amount of infection.
@@
 In its generality, it overlooks interesting natural history and biologically important features of specific plants and their pathogens.
-For example, some pathogens actively seek out and find stomata [@kiefer_host_2002], whereas I have assume that pathogens randomly orient themselves on the leaf.
+For example, some pathogens actively seek out and find stomata [@kiefer_host_2002], whereas I assumed that pathogens randomly orient themselves on the leaf.
 Including sensing should increase the probability of colonization.
@@
 The model makes two non-intuitive predictions.
-First, the effect of increased stomatal density or size on susceptibility to foliar pathogens is greatest when stomatal coverage is very low.
+First, the effect of increased stomatal density or size on susceptibility to foliar pathogens is greatest when stomatal cover is very low.
 Second, maximizing disease resistance sets up a potential conflict between minimizing stomatal cover and maximizing stomatal conductance.
-The first prediction may be relatively straightforward to test experimental with adaxial (upper) stomata that occur at low and moderate densities within the same or closely related species [@muir_morphological_2014; @mckown_association_2014; @fetter_trade-offs_2019].
-The second prediction about size-density scaling is more complex because we would to know the relationships between colonization, stomatal cover, stomatal conductance, and fitness in natural conditions.
+The first prediction may be relatively straightforward to test experimentally with adaxial (upper) stomata that occur at low and moderate densities within the same or closely related species [@muir_morphological_2014; @mckown_association_2014; @fetter_trade-offs_2019].
+The second prediction about size-density scaling is more complex because we would need to know the relationships between colonization, stomatal cover, stomatal conductance, and fitness in natural conditions.
 Testing these predictions in a variety of species would help determine whether pathogens have played an important role shaping  stomatal anatomy in land plants.
@@
 # Funding {-}
+
 I am grateful startup funds from the University of Hawaii for supporting this work.
@@
 # Conflict of Interest Statement {-}
+
 The author declares that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
@@
 
-**Table S1** | Glossary of symbols only used in \gsmax{} calculations.
+I calculated \gsmax{} (Equation \ref{eq:gsmax}) to water vapor at a reference leaf temperature ($T_\mathrm{leaf} = 25 ^ \circ$ C) following @sack_developmental_2016.
+They defined a biophysical and morphological constant as:
 
- Symbol | R      | Units  | Description
---------|--------|--------|-----------------------------------------------------
- $D_w$              | `D_w`    | $\textrm{m}^{2}~\textrm{s}^{-1}$ | diffusion coefficient for water vapour in air at $T_\mathrm{leaf}$ and $P$
- $D_{w,0}$          | `D_w0`   | $\textrm{m}^{2}~\textrm{s}^{-1}$ | diffusion coefficient for water vapour in air at 0 $^\circ$ C and air pressure 101.3246 kPa
- $eT$               | `eT`     | none | stomatal cover ()
- $P$                | `P`      | kPa | atmospheric pressure
- $T_\mathrm{leaf}$  | `T_leaf` | leaf temperature |
+\begin{align*}
+  b = & D_\text{wv} / v \\
+  m = & \frac{\pi c ^ 2}{j ^ {0.5} (4 h j + \pi c)}
+\end{align*}
+
+$b$ is the diffusion coefficient of water vapor in air ($D_\text{wv}$) divided by the kinematic viscosity of dry air ($v$).
+$D_\text{wv} = 2.49 \times 10^{-5} ~ \text{m}^2 ~ \text{s}^{-1}$ and $v = 2.24 \times 10 ^ {-2} ~ \text{m}^3 ~ \text{mol} ^ {-1}$ at $25 ^ \circ$ [@monteith_principles_2013].
+For kidney-shaped guard cells, $c = h = j = 0.5$.
+
+# References {-}

### 2019-12-10 ms/ms.tex
version submitted to Frontiers

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

\documentclass[utf8]{frontiersSCNS}

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}

\linenumbers

% BELOW TAKEN FROM rticles plos template
%
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% hyperref package, useful for hyperlinks
\usepackage{hyperref}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% Sweave(-like)
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Below is from frontiers
%
\bibliographystyle{frontiersinSCNS}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace}
%\doublespacing

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

\usepackage{longtable}
\usepackage{tikz}

% short-cut for fs, gmax, and gop
\newcommand{\fs}{$f_\text{S}$}
\newcommand{\gsmax}{$g_\text{s,max}$}
\newcommand{\gsop}{$g_\text{s,op}$}

\def\Authors{
  Christopher D. Muir\,\textsuperscript{1*}}

\def\Address{

  \textsuperscript{1} School of Life Sciences, University of Hawaii,  Honolulu,  Hawaii,  USA
  }

  \def\corrAuthor{Christopher D. Muir}\def\corrAddress{,  }\def\corrEmail{\href{mailto:cdmuir@hawaii.edu}{\nolinkurl{cdmuir@hawaii.edu}}}
  \def\firstAuthorLast{Muir {et~al.}}

\begin{document}
\onecolumn
\firstpage{1}

\title[Stomata tradeoff photosynthesis for defense]{A stomatal model of anatomical tradeoffs between photosynthesis and
pathogen defense}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

\begin{abstract}

Stomatal pores control both leaf gas exchange and are an entry for many plant pathogens, setting up the potential for tradeoffs between photosynthesis and defense. To prevent colonization and limit infection, plants close their stomata after recognizing pathogens. In addition to closing stomata, anatmoical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes would permanently reduce the gas exchange capacity for the life of the leaf. I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as a function of stomatal size and density, anatomical traits which determine maximum rates of gas exchange. The model predicts greater stomatal size or density increases colonization, but the effect is most pronounced when stomatal cover is low. I also derived scaling relationships between stomatal size and density that preserves a given probability of colonization. These scaling relationships set up a potential conflict between maximizing defense and minimizing stomatal cover. To my knowledge, this is the first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy. It makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants. The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for accurate predictions.

\tiny
 \keyFont{ \section{Keywords:} anatomy, leaf gas exchange, model, pathogen, photosynthesis, scaling, stomata, tradeoff }

\end{abstract}

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

Stomata evolved to regulate gas exchange in and out of the leaf
(Hetherington and Woodward, 2003; Berry et al., 2010; Chater et al.,
2017), but many plant pathogens take advantage of these chinks in the
leaf cuticular armor to infect prospective hosts (Zeng et al., 2010;
McLachlan et al., 2014; Melotto et al., 2017). The density and size of
stomata set the anatomical maximum rate of stomatal conductance to
CO\(_2\) and water vapor (Brown and Escombe, 1900; Parlange and
Waggoner, 1970; Franks and Farquhar, 2001; Franks and Beerling, 2009b;
Lehmann and Or, 2015; Sack and Buckley, 2016; Harrison et al., 2019),
but the pore area shrinks and expands in response to internal and
external factors to regulate gas exchange dynamically (Buckley, 2019).
Many plant pathogens, including viruses (Murray et al., 2016), bacteria
(Melotto et al., 2006; Underwood et al., 2007), protists (Fawke et al.,
2015), and fungi (Hoch et al., 1987; Zeng et al., 2010) use stomatal
pores to gain entry into the leaf. Since stomatal conductance is a major
limitation on photosynthesis (Farquhar and Sharkey, 1982; Jones, 1985)
while pathogens reduce fitness, this sets up a potential tradeoff
between increased photosynthesis and defense against pathogens. Although
there have been many empirical studies on the effect that pathogens have
on stomata, there is no theoretical framework in which to place these
findings. Lack of a theoreatical framework makes it difficult to answer
general questions about how selection for pathogen defense constrains
maximum rates of gas exchange.

Stomatal anatomy is the key link between gas exchange and pathogen
colonization. The density and size of stomata not only determines the
theoretical maximum stomatal conductance (\gsmax), but is also
proportional to the operational stomatal conductance (\gsop) in many
circumstances (Franks et al., 2009, 2014; Dow et al., 2014a; McElwain et
al., 2016; Murray et al., 2019). Therefore, I use anatomical \gsmax{} as
a proxy for \gsop{} and do not address dynamic changes in stomatal
aperture (see \protect\hyperlink{discussion}{Discussion}). Harrison et
al. (2019) recently reviewed the relationship between stomatal anatomy
and gas exchange in detail.

Many pathogens rely on stomata to gain entry into the leaf, but it is
unclear how anatomical traits that determine \gsmax{} (size and density)
affect the ability of pathogens to colonize leaves. The impact of
pathogens on host fitness is complex, but after a pathogen reaches a
host, the first major step is colonization. Once infected, a pathogen
can reduce fitness. Susceptible hosts can lose much of their biomass or
die, but even resistant hosts must allocate resources to defense or
reduce photosynthesis because of necrosis around sites of infection.
Plants can limit colonization physiologically by closing stomata after
they recognize pathogens, called stomatal defense (Melotto et al.,
2017). Anatomy may be another layer of defense; plants can reduce
pathogen colonization by developing leaves with lower stomatal density
and/or size. Infection increases in leaves with higher stomatal density
(McKown et al., 2014; Tateda et al., 2019; Dutton et al., 2019; Fetter
et al., 2019), which suggests that both anatomical and physiological
responses (stomatal closure) affect host colonization. By the same
logic, if stomatal density were held constant, larger stomata should
also increase colonization because they occupy more area. One key
difference between physiological and anatomical defenses is that
stomatal closure reduces gas exchange transiently, whereas anatomy
constrains gas exchange throught the life of the leaf. Because we do not
understand the relationship between stomatal anatomy and colonization
well, we cannot predict the relationship between \gsmax{} and infection.

If stomatal size and density affect pathogen colonization, selection to
limit colonization may shape stomatal size-density scaling
relationships. Botanists have long recognized that stomatal size and
density are inversely correlated (Weiss, 1865; Tichá, 1982; Hetherington
and Woodward, 2003; Sack et al., 2003; Franks and Beerling, 2009a;
Brodribb et al., 2013; de Boer et al., 2016), but the evolutionary
origin of this relationship is not yet known. Here I argue that
pathogens could shape selection on this relationship. Explanations for
inverse size-density scaling are usually cast in terms of preserving
\gsmax{} and/or total epidermal area allocated to stomata (\fs; de Boer
et al., 2016) because there are many combinations of stomatal size and
density that have same \gsmax{} or same \fs{}:

\begin{align}
  g_\text{s,max} & = b m D S ^ {0.5} \label{eq:gsmax} \\
  f_\text{S} & = D S \label{eq:fs}.
\end{align}

\(b\) and \(m\) are assumed to be biophysical and morphological
constants, \emph{sensu} Sack and Buckley (2016) (see
\protect\hyperlink{supplementary-material}{Supplementary Material}). If
size and density also affect pathogen colonization, then selection from
foliar pathogens could significantly alter the size-density scaling
relationship. The empirical size-density scaling relationship is linear
on a log-log scale, determined by an intercept \(\alpha\) and slope
\(\beta\):

\begin{align}
  D & = e ^ \alpha S ^ {-\beta}; \\
  d & = \alpha - \beta S.
\end{align}

For brevity, \(d = \text{log}(D)\) and \(s = \text{log}(S)\).
Rearranging Equations \ref{eq:gsmax} and \ref{eq:fs}, a scaling
relationship where \(\beta = 0.5\) preserves \gsmax{} while
\(\beta = 1\) preserves \fs.

How would adding pathogens alter these predicted scaling relationships?
For simplicity, imagine two environments, one without foliar pathogens
and one with lots. In the absence of foliar pathogens, we expect
size-density scaling to preserve \gsmax{}, \fs{}, or some least-cost
combination of them. What happens when we introduce pathogens? Assuming
that stomatal size and density increase pathogen colonization, then
selection will favor reduced size and/or density. This would change the
intercept \(\alpha\) but not the slope. The effect of foliar pathogens
on the slope depends on the relationship between size, density, and
probability of colonization. If the probability of colonization is
proportional to the product of \emph{linear} stomatal size
(\(S ^ {0.5}\)) and density (\(\propto D S ^ {0.5}\) as for \gsmax) then
it has the same effect on the slope as \gsmax{} because there are many
combinations of \(D\) and \(S ^ {0.5}\) that have same probability of
colonization. If the probability of colonization is proportional to the
product of \emph{areal} stomatal size (\(S\)) and density
(\(\propto D S\) as for \fs) then it has the same effect on the slope as
\fs{} because there are many combinations of \(D\) and \(S\) that have
same probability of colonization. Alternatively, the probability of
colonization may have a different scaling relationsip (neither 0.5 nor
1) or may be nonlinear on a log-log scale. Unlike \gsmax{} and \fs{}, we
do not have theory to predict a stomatal size-density relationship that
preserves the probability of colonization.

In summary, the physical relationship between stomatal size, density,
and conductance is well established (Harrison et al., 2019). The same
traits likely affect the probability of pathogen colonization, but we do
not have a theoretical model that makes quantitative predictions. The
inverse stomatal size-density relationship has usually been explained in
terms of preserving stomatal conductance and/or stomatal cover, but
selection by pathogens might alter scaling. To address these gaps, the
goals of this study are to 1) introduce a spatially explicit model
pathogen colonization on the leaf surface; 2) use the model to predict
the relationship between \gsmax, \fs{}, and the probability of
colonization; 3) work out what these relationships predict about
stomatal size-density scaling.

\hypertarget{model}{%
\section*{Model}\label{model}}
\addcontentsline{toc}{section}{Model}

In this section, I introduce a spatially explicit model of pathogen
colonization on a leaf surface. I explain the model structure and
assumptions here; the
\protect\hyperlink{materials-and-methods}{Materials and Methods} section
below describes how I analyzed the model to address the goals of the
study. For generality, I refer to a generic ``pathogen'' that lands on
leaf and moves to a stomate. The model is agnostic to the type of
pathogen (virus, bacterium, fungus, etc.) and the specific biological
details of how it moves (biotrophy). For example, motile bacterial cells
can land and move around where fungi may germinate from a cyst and grow
until they form an appresorium for infection. These very different
biotropic movements on the leaf are treated identically here. I used
Sympy version 1.3 (Meurer et al., 2017) for symbolic derivations.

\textbf{Table 1} \textbar{} Glossary of mathematical symbols. The
columns indicate the mathematical Symbol used in the paper, the
associated symbol used in R scripts, scientific Units, and a verbal
Description.

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.09\columnwidth}\raggedright
Symbol\strut
\end{minipage} & \begin{minipage}[b]{0.09\columnwidth}\raggedright
R\strut
\end{minipage} & \begin{minipage}[b]{0.09\columnwidth}\raggedright
Units\strut
\end{minipage} & \begin{minipage}[b]{0.61\columnwidth}\raggedright
Description\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(D,d\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{D,\ d}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\textrm{mm}^{-2}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
stomatal density (\(d = \text{log}~D\))\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(f_s\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{f\_s}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
none\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
stomatal cover (\(f_s = D S\))\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(g_\text{s,max}\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{g\_smax}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\text{mol}~\text{m}^{-2}~\text{s}^{-1}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
anatomical maximum stomatal conductance\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(g_\text{s,op}\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{g\_sop}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\text{mol}~\text{m}^{-2}~\text{s}^{-1}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
operational stomatal conductance\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(H\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{H}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\mu\textrm{m} ^{-1}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
death rate of pathogen on leaf surface\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(R\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{R}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\mu\textrm{m}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
stomatal radius (\(S = 2 \pi R ^ 2\))\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(S,s\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{S,\ s}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\mu\textrm{m}^2\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
stomatal size (\(s = \text{log}~S\))\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\theta_i\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{theta\_i}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
radians\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
angles between pathogen (\(x_p,y_p\)) and lines tangent to the
circumfrence of stomate \(i\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(U\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{U}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\mu\textrm{m}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
interstomatal distance\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(v_i\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{v\_i}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\mu\textrm{m}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
distance between pathogen (\(x_p,y_p\)) and stomate \(i\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(x_i,y_i\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{x\_i,y\_i}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\mu\textrm{m}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
position of stomate \(i\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright
\(x_p,y_p\)\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\texttt{x\_p,y\_p}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
\(\mu\textrm{m}\)\strut
\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
starting position of pathogen\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{spatial-representation-of-stomata}{%
\subsection*{Spatial representation of
stomata}\label{spatial-representation-of-stomata}}
\addcontentsline{toc}{subsection}{Spatial representation of stomata}

Stomata develop relatively equal spacing to minimize resistance to
lateral diffusion (Morison et al., 2005), allow space between stomata
(Dow et al., 2014b), and prevent stomatal interference (Lehmann and Or,
2015). Here I assume that stomata are arrayed in an equilateral
triangular grid with a density \(D\) and size (area) \(S\). This
assumption ignores veins, trichomes, and within-leaf variation in
stomatal density. Stomata are therefore arrayed in an evenly spaced grid
(Figure \ref{fig:fig1}a). The interstomatal distance \(U\), measured as
the distance from the center of one stomata to the next, is the maximal
diagonal of the hexagon in \(\mu\textrm{m}\) that forms an equal area
boundary between neighbording stomata. The area of a hexagon is
\(A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2\). By definition the
stomatal density is the inverse of this area, such that
\(D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}\).
Therefore, interstomatal distance can be derived from the stomatal
density as:

\[ D = \frac{2}{\sqrt{3}} U ^{-2} \]
\[ U = \bigg(\frac{2}{\sqrt{3}} D ^{-1}\bigg) ^ {0.5} \]

For example, if the density is
\(D = 10 ^ 2~\textrm{mm}^{-2} = 10 ^ {-4}~\mu\textrm{m}^{-2}\), then
\(U\) is 107.5 \(\mu \textrm{m}\). Parkhurst (1994) described this
result previously. I also make the simplifying assumption that stomata
are perfectly circular with radius \(R\). This may be approximately true
for fully open stomata with kidney-shaped guard cells. Although I assume
stomata are circular here, in calculating \gsmax, I assume typical
allometric relationships between length, width, and pore area (see
\protect\hyperlink{supplementary-material}{Supplementary Material}).

\begin{figure}
  \centering
    \includegraphics[width=6.5in]{../figures/fig1.jpeg}
    \caption{\textbf{A spatially explicit model of stomatal anatomy and pathogen colonization}. \textbf{a}. Stomata are assumed to be in a homogenous equilateral triangular grid, which means that we can extrapolate from \textbf{b.} a focal triangle to the entire leaf. The circles represent idealized stomata; the grey lines between them are for visualization. \textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the area. \textbf{d.} The model assumes that a pathogen, depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves at the correct angle, depicted by the grey polygons.}
    \label{fig:fig1}
\end{figure}

\hypertarget{spatial-representation-of-pathogen-search}{%
\subsection*{Spatial representation of pathogen
search}\label{spatial-representation-of-pathogen-search}}
\addcontentsline{toc}{subsection}{Spatial representation of pathogen
search}

Now imagine that a pathogen lands at a uniform random position within
the focal region and must arrive at a stomate to colonize. If it lands
on a stomate, then it infects the leaf with probability 1; if it lands
between stomata, then it infects the leaf with probability
\(p_\text{locate}\). This is the probability that it locates a stomate,
which I will derive below. The probabilities of landing on or between a
stomate are \fs{} and \(1 - f_S\), respectively. Hence, the total
probability of colonization is:

\begin{equation}
  \label{eq:p_colonize}
  p_\text{colonize} = f_\text{S} + (1 - f_\text{S}) p_\text{locate}.
\end{equation}

I assume that the pathogen cannot sense where stomata are and orients at
random, thereafter traveling in that direction. If it successfully
locates a stomate, it colonizes the leaf, but otherwise does not infect.
If there is a high density of stomata and/or large stomata, the
probability of locating a stomate increases. By assuming that stomata
form an equilateral triangular grid (see above), we can extrapolate what
happens in a focal triangle (Figure \ref{fig:fig1}b) by symmetry.
Further, since an equilateral triangle can be broken up into six
identical units (Figure \ref{fig:fig1}c), we can simply calculate
\(p_\text{locate}\) in this focal area. This implicitly assumes that the
probability of colonzing stomata outside the focal area is 0 because
they are too far away.

Imagine that the pathogens lands in position (\(x_p,y_p\)) within the
triangle. The centroid of the triangle is at position (\(x_c,y_c\)) and
a reference stomate is at position \((0,0)\) (Figure \ref{fig:fig2}a).
Therefore \(x_c = U / 2\) and \(y_c = \sqrt{3} U / 6\). The other
stomata are at positions \((U/2, \sqrt{3} U /2)\) and \((U,0)\) (Figure
\ref{fig:fig2}). \(x_p\) and \(y_p\) are defined as the horizontal and
vertical distances, respectively, from the pathogen to the reference
stomate at position \((0,0)\).

Given that the pathogen starts at position (\(x_p,y_p\)), what's the
probability of contacting one of the stomata at the vertices of the
focal triangle? I assume the probability of contacting a stomate is
equal to the proportion of angular directions that lead to a stomate
(Figure \ref{fig:fig1}d). I solved this by finding the angles
(\(\theta_1, \theta_2, \theta_3\)) between lines that are tangent to the
outside of the three stomata and pass through (\(x_p,y_p\)) (Figure
\ref{fig:fig2}a). If stomate \(i\) is centered at (\(x_i,y_i\)), the two
slopes of tangency as function of pathogen position are:

\begin{align}
  t_{i,1}(x_p,y_p) & = \frac{-R e_{i,2}(x_p,y_p) + e_{i,3}(x_p,y_p)}{e_{i,1}(x_p,y_p)} \\
  t_{i,2}(x_p,y_p) & = \frac{R e_{i,2}(x_p,y_p) + e_{i,3}(x_p,y_p)}{e_{i,1}(x_p,y_p)}
\end{align}

where

\begin{align}
  e_{i,1}(x_p,y_p) & = (R^2 - x_i^2 + 2 x_i x_p - x_p^2), \\
  e_{i,2}(x_p,y_p) & = \sqrt{-e_{i,1} + (y_i - y_p) ^ 2}, \\
  e_{i,3}(x_p,y_p) & = - x_i y_i + x_i y_p + x_p y_i - x_p y_p.
\end{align}

Note that \(i \in \{1, 2, 3\}\), indexing the three stomata in the focal
triangle. The angle in radians between \(t_{i,1}(x_p,y_p)\) and
\(t_{i,2}(x_p,y_p)\) is:

\begin{equation}
  \theta_i(x_p,y_p) = \text{arctan}\bigg(\frac{t_{i,1}(x_p,y_p) - t_{i,2}(x_p,y_p)}{1 + (t_{i,1}(x_p,y_p) t_{i,2}(x_p,y_p))}\bigg)
\end{equation}

\begin{figure}
\centering
\input{../figures/fig2.tex}
\caption{\textbf{Spatial representation of stomata and pathogen.} \textbf{a.} The pathogen starts at a uniform random position within the focal region denoted ($x_p,y_p$). Within the focal triangle, the reference stomata is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$. \textbf{b.} Within the focal region, a pathogen can land within the stomate (white cirlce with grey outline and radius $R$) or in the grey area. The outer borders of this area are shown and depend on $R$ and $U$. For a given position $x$, there is a minimum $y$-value ($y_\text{min}$, dashed line) and maximum $y$-value ($y_\text{max}$, solid line).}
\label{fig:fig2}
\end{figure}

I further assumed that the longer distance a pathogen must travel, the
less likely it would be to locate a stomate. For example, if stomata are
at very low density, then a pathogen may die before it reaches a stomate
because of UV, dessication, or another factor. I included this effect by
assuming the probability of reaching a stomate declines exponentially at
rate \(H\) with the Euclidean distance \(v_i(x_p,y_p)\) between the
pathogen location and the edge of stomata \(i\), which is distance \(R\)
from its center at \(x_i,y_i\):

\begin{equation}
  v_i(x_p,y_p) = \sqrt{(x_i - x_p) ^ 2 + (y_i - y_p) ^ 2} - R.
\end{equation}

The probability of locating a stomate as a function of \(x_p\) and
\(y_p\) (\(f_\text{locate}(x_p,y_p)\)) is the sum of the angles divided
by \(2\pi\), discounted by their distance from the stomate:

\begin{equation}
  f_\text{locate}(x_p,y_p) = \frac{1}{2 \pi} \sum_{i=1}^3 e^{-H v_i(x_p,y_p)} \theta_i(x_p,y_p)
\end{equation}

When \(H = 0\), \(p_\text{locate}\) is the fraction of angles that lead
from \((x_p,y_p)\) to a stomate. When \(H > 0\), \(p_\text{locate}\) is
proportional to this fraction, but less than it depending on stomatal
density, size, and starting location of the pathogen.

To obtain the average \(p_\text{locate}\), we must integrate
\(f_\text{locate}(x_p,y_p)\) over all possible starting positions
\((x_p,y_p)\) within the focal area. The focal area is a 30-60-90
triangle with vertices at the center of the reference stomate \((0,0)\),
the midpoint of baseline \((U/2,0)\), and the centroid of the focal
trinagle \((U/2, \sqrt{3} / 6 U)\) (Figure \ref{fig:fig1}c).
Colonization occurs with probability 1 if the pathogen lands in the
reference stomate, so we need to integrate the probability of
colonization if it lands elsewhere. This region extends from the edge of
the stomate, at \(\sqrt{3} / 2 R\) to \(U/2\) (Figure \ref{fig:fig2}b).
At any \(x\), we integrate from the bottom of the focal area
(\(y_\text{min}\)) to the top (\(y_\text{max}\)):

\begin{align}
  y_\text{min} = f(x) & = \begin{cases}
    \sqrt{R ^ 2 - x ^ 2}, & \text{if } \frac{\sqrt{3}}{2} R < x < R \\ 
    0, & \text{if } R \leq x \leq \frac{U}{2} \end{cases} \\
  y_\text{max} = f(x) & = \frac{\sqrt{3}}{3} x
\end{align}

The integral is:

\begin{equation}
  p_\text{locate} = \frac{1}{a_\text{focal}}\int\limits_{\frac{\sqrt{3}}{2} R} ^ {U / 2} \int\limits_{y_\mathrm{min}} ^ {y_\mathrm{max}} f_\text{locate}(x, y) \,dx \,dy
\end{equation}

\(a_\text{focal}\) is the area of the focal region depicted in grey in
Figure \ref{fig:fig2}b:

\[ a_\text{focal} = \frac{U ^ 2}{8 \sqrt{3}} - \frac{\pi R ^ 2}{12} \]

\hypertarget{materials-and-methods}{%
\section*{Materials and Methods}\label{materials-and-methods}}
\addcontentsline{toc}{section}{Materials and Methods}

The \protect\hyperlink{model}{Model} calculates a probability of host
colonization (Equation \ref{eq:p_colonize}) as a function stomatal
density, size, and position of a pathogen on the leaf. I solved
\(p_\text{colonize}\) using the \texttt{integral2()} function in the
\textbf{pracma} package version 2.2.5 (Borchers, 2019) for numerical
integration. I used R version 3.6.1 (R Core Team, 2019) for all analyses
and wrote the paper in \textbf{rmarkdown} version 1.17 (Xie et al.,
2018; Allaire et al., 2019). Source code is deposited on GitHub
(\url{https://github.com/cdmuir/stomata-tradeoff}) and will be arcived
on Zenodo upon publication.

\hypertarget{what-is-the-relationship-between-stomatal-size-density-and-colonization}{%
\subsection*{What is the relationship between stomatal size, density,
and
colonization?}\label{what-is-the-relationship-between-stomatal-size-density-and-colonization}}
\addcontentsline{toc}{subsection}{What is the relationship between
stomatal size, density, and colonization?}

I calculated \(p_\text{colonize}\) over a biologically plausible grid of
stomatal size and density for hypostomatous species based on de Boer et
al. (2016). Stomatal density ranges from
\(10^1 - 10^{3.5}~\text{mm}^{-2}\); stomatal size ranges from
\(10^1 - 10^{3.5}~\mu\text{m}^2\). I only considered combinations of
size and density where \fs{} was less than 1/3. For simplicity, I have
not extended the current analysis to amphistomatous leaves. I crossed
stomatal traits with three levels of \(H \in \{0, 0.01, 0.1\}\). When
\(H = 0\), a pathogen perists indefinitely on the leaf surface.
\(H = 0.01\) and \(H = 0.1\) correspond to low and high death rates,
respectively. These values are not necessarily realistic, but illustrate
qualitatively how a hostile environment on the leaf surface alters model
predictions.

\hypertarget{how-do-pathogens-alter-optimal-stomatal-size-density-scaling}{%
\subsection*{How do pathogens alter optimal stomatal size-density
scaling?}\label{how-do-pathogens-alter-optimal-stomatal-size-density-scaling}}
\addcontentsline{toc}{subsection}{How do pathogens alter optimal
stomatal size-density scaling?}

The stomatal size-density scaling relationship can be explained in terms
of preserving a constant \gsmax{} that is proportional to
\(D S ^ {0.5}\) when \(bm\) is constant (Equation \ref{eq:gsmax}). In
other words, there are infinitely many combinations of \(D\) and
\(S ^ {0.5}\) with the same \gsmax. If \gsmax{} is held constant at
\(C_g\), then the resulting size-density scaling relationship on a
log-log scale is:

\[ d = c_g - 0.5 s \] where lowercase variables are log-transformed
equivalents of their uppercase counterparts. The scaling exponent
\(\beta_g = 0.5\) preseves \(C_g\).

Next, imagine there is similarly a scaling exponent \(\beta_p\) that
preserves \(p_\text{colonize}\) for the product \(D S^{\beta_p}\). If
\(\beta_p = 0.5\), then \(p_\text{colonize}\) is always proportional to
\gsmax. If \(\beta_p > 0.5\), small, densely packed stomata would be
better defended (lower \(p_\text{colonize}\)) compared to larger,
sparsely spaced stomata with the same \gsmax. If \(\beta_p < 0.5\),
small, densely packed stomata would be less defended (higher
\(p_\text{colonize}\)) compared to larger, sparsely spaced stomata with
the same \gsmax. I refer to the three outcomes (\(\beta_p = 0.5\),
\(\beta_p < 0.5\), and \(\beta_p > 0.5\)) as iso-, hypo-, and
hyper-conductance, respectively. I was unable to solve analytically for
\(\beta_p\), so I numerically calculated isoclines of
\(p_\text{colonize}\) over the grid of \(D\) and \(S\) values described
in the preceding subsection. I numerically calculated the scaling
relationships at a constant
\(p_\text{colonize} \in \{0.025, 0.05, 0.1, 0.2, 0.4\}\) for
\(H \in \{0, 0.01, 0.1\}\).

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

I analyzed an idealized, spatially explicit
\protect\hyperlink{model}{Model} of how a pathogen lands on a leaf and
finds a stomate to colonize the leaf using a random search. To my
knowledge, this is the first model that makes quantitative predictions
about the relationship between stomatal anatomy, the probability of
colonization, and their impact on stomatal size-density scaling.

\hypertarget{nonlinear-relationships-between-colonization-stomatal-cover-and-conductance}{%
\subsection*{Nonlinear relationships between colonization, stomatal
cover, and
conductance}\label{nonlinear-relationships-between-colonization-stomatal-cover-and-conductance}}
\addcontentsline{toc}{subsection}{Nonlinear relationships between
colonization, stomatal cover, and conductance}

The probability of colonization (\(p_\text{colonize}\)) is not simply a
one-to-one relationship between the fraction of epidermal area allocated
to stomata (\fs). At low \fs, \(p_\text{colonize}\) increases faster
rapdily relative to \fs{} at first Figure \ref{fig:fig3}a). At higher
\fs, the \(p_\text{colonize}\) increases linearly with \fs. When
\(H = 0\), any combination of stomatal size (\(S\)) and density (\(D\))
with the same \fs{} have the same effect on \(p_\text{colonize}\). When
\(H > 0\), pathogens are less likely to land close enough to a stomate
to infect before dying, so \(p_\text{colonize}\) is closer to \fs{}
(Figure \ref{fig:fig3}a). Furthermore, \(p_\text{colonize}\) depends on
\(D\) and \(S\), not just \fs. For the same \fs, leaves with greater
\(D\) have higher \(p_\text{colonize}\) (Figure \ref{fig:fig3}a).
Holding \fs{} constant, leaves with lower \(D\) and higher \(S\) will
have a greater distance (\(v_i\)) between a pathogen and its stomata.
When \(H > 0\), this extra distance leads more pathogens to die before
they can find a stomate. In contrast to \fs, \(p_\text{colonize}\)
increases at a greater than linear rate with \gsmax. Greater \(D\)
(smaller \(S\)) is associated with lower \(p_\text{colonize}\) when
\gsmax{} is held constant (Figure \ref{fig:fig3}b). This happens because
\(p_\text{colonize}\) increases approximately linearly with \(S\)
whereas \gsmax{} is proportional to \(S^{0.5}\).

\begin{figure}
  \centering
    \includegraphics{../figures/fig3.pdf}
    \caption{\textbf{The probability of colonization increases with both stomatal cover and conductance} \textbf{a.} The probability of colonization ($p_\text{colonize}$, $y$-axis) initially increases rapidly with stomatal cover (\fs), then slows down to a linear relationship. Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$). The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density when $H = 0$ (upper facet). When $H > 0$, higher density (solid lines) increase $p_\text{colonize}$ (lower facets). \textbf{b.} $p_\text{colonize}$ increases expoentially with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is much lower at higher densities for a given \gsmax. The relationship between \gsmax{} and $p_\text{colonize}$ is similar for all values of $H$.}
    \label{fig:fig3}
\end{figure}

\hypertarget{hyper-conductance-size-density-scaling}{%
\subsection*{Hyper-conductance size-density
scaling}\label{hyper-conductance-size-density-scaling}}
\addcontentsline{toc}{subsection}{Hyper-conductance size-density
scaling}

The scaling relationship between \(S\) and \(D\) that preserves
\(p_\text{colonize}\) is always greater 0.5 (hyper-conductance), but
usually less than 1. When \(H = 0\), the scaling relationship is
essentially 1 (Figure \ref{fig:fig4}), which means that an increase
\fs{} leads to a proportional increase in \(p_\text{colonize}\). Because
the scaling relationship is greater than 0.5, leaves with greater
stomatal density will have lower \(p_\text{colonize}\) than leaves lower
stomatal density but the same \gsmax. In other words, increasing \(D\)
and lowering \(S\) allows plants to reduce \(p_\text{colonize}\) while
maintaining \gsmax. The scaling relationship is slightly less than 1,
but still greater than 0.5, when \(H > 0\) (Figure \ref{fig:fig4}). In
this area of parameter space, lower stomatal density can reduce \fs{}
while \(p_\text{colonize}\) is constant, but this will still result in
lower \gsmax.

\begin{figure}
  \centering
    \includegraphics{../figures/fig4.pdf}
    \caption{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$). In each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.025 (lowest line), 0.05, 0.1, 0.2, or 0.4 (highest line). For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.5$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region. When the death rate on the leaf surface is low ($H = 0$), the scaling exponent is very close to $\beta = 1$. When $H > 0$, $0.5 < \beta < 1$ and is slightly nonlinear on a log-log scale.}
    \label{fig:fig4}
\end{figure}

\hypertarget{discussion}{%
\section*{Discussion}\label{discussion}}
\addcontentsline{toc}{section}{Discussion}

Stomatal density and size set the upper limit on gas exchange in leaves
(Harrison et al., 2019) and is often closely related to operational
stomatal conductance in nature (Murray et al., 2019). Despite the fact
that many ecologically and economically significant plant pathogens
infect through stomata, the relationship between stomatal anatomy and
susceptibility to foliar pathogens is less clear than it is for gas
exchange. To develop testable predictions, we need mathematical models
that can clarify the potential for tradeoffs between stomatal
conductance, stomatal cover, and disease resistance. I used a spatially
explicit model of a pathogen searching for a stomate to colonize a host.
From this \protect\hyperlink{model}{Model}, I derived predictions about
the relationship between stomatal anatomy and disease resistance for the
first time. The model predicts that the probability of colonization is
not always proportional to the surface area of leaf covered by stomata
(\fs), as one might intuitively predict. If the leaf surface is a
hostile environment and pathogens have a limited time to search, lower
stomatal density decreases the probability of colonization even if \fs{}
is constant. However, \gsmax{} decreases proportionally more than the
probability of colonization. The model reveals the potential for
conflicting demands of maximizing disease resistance, minimizing
stomatal cover, and maintaining stomatal conductance. Including the
effect of anatomy on disease resistance therefore has the potential to
change our understading of how stomatal size-density scaling evolves in
land plants.

The model predicts that in most cases, increasing stomatal cover should
lead to a proportonal increase in susceptibility, which is the implicit
assumption of some empirical studies (e.g.~McKown et al. (2014); Tateda
et al. (2019); Dutton et al. (2019); Fetter et al. (2019)). It also
makes new, testable predictions that are less intuitive. At very low
\fs, there is a rapid increase in susceptibility (Figure
\ref{fig:fig3}a). If there are no stomata, the probability of
colonization is 0, so the first few stomata dramatically increase the
probability. This is unlikely to be significant for abaxial (lower) leaf
surfaces, which usually have most of the stomata (Salisbury, 1928;
Metcalfe and Chalk, 1950; Mott et al., 1984; Peat and Fitter, 1994;
Jordan et al., 2014; Muir, 2015; Bucher et al., 2017; Drake et al.,
2019). However, many adaxial (upper) leaf surfaces have zero or very few
stomata. Using adaxial leaf surfaces, it should be possible to test if
small changes in stomatal size or density have a larger effect on
disease susceptibility when \fs{} is low. The nonlinear increase in
\(p_\text{colonize}\) is less apparent when \(H > 0\) (Figure
\ref{fig:fig3}a). A more hostile microenvironment (e.g.~drier, higher
UV) should therefore reduce the effect of increased size or density as
low \fs. If true, the dimishing marginal effect of \fs{} on colonization
could explain why stomatal ratio on the upper and lower surface is
bimodal (Muir, 2015). The initial cost of adaxial (upper) stomata is
high, but if the benefits outweigh the costs, then equal stomatal
densities on each surface maximize CO\(_2\) supply for photosynthesis
(Parkhurst, 1978; Gutschick, 1984; Parkhurst and Mott, 1990).

An effect of stomatal size and density on susceptibility to foliar
pathogens could change our understanding of stomatal size-density
scaling. Since allocating leaf epidermis to stomata may be costly
(Franks and Farquhar, 2007; Assmann and Zeiger, 1987; Dow et al., 2014b;
Lehmann and Or, 2015; Baresch et al., 2019), selection should favor
leaves that achieve a desired \gsmax{} while minimizing \fs{} (de Boer
et al., 2016). Because of their different scaling exponents (Equation
\ref{eq:gsmax}, \ref{eq:fs}), smaller, densely packed stomata can
achieve the same \gsmax{} at minimum \fs. However, many leaves have
larger, sparsely packed stomata. Incorporating susceptibility to disease
may explain why. If pathogens have a limited time to find stomata before
dying (\(H > 0\)), then the scaling exponent between size and density
that keeps \(p_\text{colonize}\) constant is between 0.5 and 1, the
scaling exponents for \gsmax{} and \fs, respectively (Figure
\ref{fig:fig4}). Greater density of smaller stomata can increase
\gsmax{} while keeping \(p_\text{colonize}\) constant, but this will
increase \fs. Conversely, \fs{} could decrease while keeping
\(p_\text{colonize}\) constant, but this will decrease \gsmax. This sets
up the potential for conflict between competing goals. The optimal
stomatal size and density will therefore depend on the precise costs and
benefits of infection, stomatal conductance, and stomatal cover. This
may explain why many leaves have large, sparsely packed stomata despite
the fact that they could achieve the same \gsmax{} and lower \fs{} with
smaller, more densely packed stomata.

The model examines the probability of colonization for a single
pathogen. The calculated probabilities of colonization should not be
interpreted as exact predictions, but rather as depicting qualitative
relationships between stomatal anatomy and infection severity. The model
is most applicable to diseases where the host has some resistance. The
energetic cost and lost photosynthetic capacity (closed stomata,
necrosis, etc.) of dealing with a pathogen is assumed to be proportional
to the amount of infection. The actual fitness cost will be modulated by
the number of pathogens landing on the leaf and the cost of infection.
In environments with fewer or less virulent pathogens, the fitness cost
of infection will be less than in environments with more abundant,
virulent pathogens. The model is less relevant to very susceptible host
plants that can be severely damaged or killed by a small number of
colonizations that spread unchecked throughout the host tissue.

The purpose of this model is to provide a general foundation to examine
the relationship between stomatal size, density, and defense against
foliar pathogens. In its generality, it overlooks interesting natural
history and biologically important features of specific plants and their
pathogens. For example, some pathogens actively seek out and find
stomata (Kiefer et al., 2002), whereas I assumed that pathogens randomly
orient themselves on the leaf. Including sensing should increase the
probability of colonization. I also assumed that the plant does not
respond by, for example, closing stomata when it senses a pathogen.
However, plants can sense and close stomata when pathogens land on the
leaf, but pathogens can pry stomata back open (Melotto et al., 2006).
Future work on specific plant-pathogen interactions could build on this
model by adding more biological realism to provide more precise
predictions.

\hypertarget{conclusion}{%
\section*{Conclusion}\label{conclusion}}
\addcontentsline{toc}{section}{Conclusion}

The model makes two non-intuitive predictions. First, the effect of
increased stomatal density or size on susceptibility to foliar pathogens
is greatest when stomatal cover is very low. Second, maximizing disease
resistance sets up a potential conflict between minimizing stomatal
cover and maximizing stomatal conductance. The first prediction may be
relatively straightforward to test experimentally with adaxial (upper)
stomata that occur at low and moderate densities within the same or
closely related species (Muir et al., 2014; McKown et al., 2014; Fetter
et al., 2019). The second prediction about size-density scaling is more
complex because we would need to know the relationships between
colonization, stomatal cover, stomatal conductance, and fitness in
natural conditions. Testing these predictions in a variety of species
would help determine whether pathogens have played an important role
shaping stomatal anatomy in land plants.

\hypertarget{funding}{%
\section*{Funding}\label{funding}}
\addcontentsline{toc}{section}{Funding}

I am grateful startup funds from the University of Hawaii for supporting
this work.

\hypertarget{conflict-of-interest-statement}{%
\section*{Conflict of Interest
Statement}\label{conflict-of-interest-statement}}
\addcontentsline{toc}{section}{Conflict of Interest Statement}

The author declares that the research was conducted in the absence of
any commercial or financial relationships that could be construed as a
potential conflict of interest.

\hypertarget{supplementary-material}{%
\section*{Supplementary Material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary Material}

I calculated \gsmax{} (Equation \ref{eq:gsmax}) to water vapor at a
reference leaf temperature (\(T_\mathrm{leaf} = 25 ^ \circ\) C)
following Sack and Buckley (2016). They defined a biophysical and
morphological constant as:

\begin{align*}
  b = & D_\text{wv} / v \\
  m = & \frac{\pi c ^ 2}{j ^ {0.5} (4 h j + \pi c)}
\end{align*}

\(b\) is the diffusion coefficient of water vapor in air
(\(D_\text{wv}\)) divided by the kinematic viscosity of dry air (\(v\)).
\(D_\text{wv} = 2.49 \times 10^{-5} ~ \text{m}^2 ~ \text{s}^{-1}\) and
\(v = 2.24 \times 10 ^ {-2} ~ \text{m}^3 ~ \text{mol} ^ {-1}\) at
\(25 ^ \circ\) (Monteith and Unsworth, 2013). For kidney-shaped guard
cells, \(c = h = j = 0.5\).

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-allaire_rmarkdown:_2019}{}%
Allaire, J., Xie, Y., McPherson, J., Luraschi, J., Ushey, K., Wickham,
H., Cheng, J., Chang, W., and Iannone, R. (2019). Rmarkdown: Dynamic
Documents for R. Available at:
\url{https://github.com/rstudio/rmarkdown}.

\leavevmode\hypertarget{ref-assmann_guard_1987}{}%
Assmann, S. M., and Zeiger, E. (1987). ``Guard Call Bioenergetics,'' in
\emph{Stomatal Function}, eds. E. Zeiger, G. D. Farquhar, and I. R.
Cowan (Stanford University Press), 163--193.

\leavevmode\hypertarget{ref-baresch_competition_2019}{}%
Baresch, A., Crifò, C., and Boyce, C. K. (2019). Competition for
epidermal space in the evolution of leaves with high physiological
rates. \emph{New Phytologist} 221, 628--639.
doi:\href{https://doi.org/10.1111/nph.15476}{10.1111/nph.15476}.

\leavevmode\hypertarget{ref-berry_stomata:_2010}{}%
Berry, J. A., Beerling, D. J., and Franks, P. J. (2010). Stomata: Key
players in the earth system, past and present. \emph{Current Opinion in
Plant Biology} 13, 232--239.
doi:\href{https://doi.org/10.1016/j.pbi.2010.04.013}{10.1016/j.pbi.2010.04.013}.

\leavevmode\hypertarget{ref-borchers_pracma:_2019}{}%
Borchers, H. W. (2019). Pracma: Practical Numerical Math Functions. R
package version 2.2.5. Available at:
\url{https://CRAN.R-project.org/package=pracma}.

\leavevmode\hypertarget{ref-brodribb_unified_2013}{}%
Brodribb, T. J., Jordan, G. J., and Carpenter, R. J. (2013). Unified
changes in cell size permit coordinated leaf evolution. \emph{New
Phytologist} 199, 559--570.
doi:\href{https://doi.org/10.1111/nph.12300}{10.1111/nph.12300}.

\leavevmode\hypertarget{ref-brown_static_1900}{}%
Brown, H. T., and Escombe, F. (1900). Static diffusion of gases and
liquids in relation to the assimilation of carbon and translocation in
plants. \emph{Proceedings of the Royal Society of London} 67, 124--128.

\leavevmode\hypertarget{ref-bucher_stomatal_2017}{}%
Bucher, S. F., Auerswald, K., Grün-Wenzel, C., Higgins, S. I., Garcia
Jorge, J., and Römermann, C. (2017). Stomatal traits relate to habitat
preferences of herbaceous species in a temperate climate. \emph{Flora}
229, 107--115.
doi:\href{https://doi.org/10.1016/j.flora.2017.02.011}{10.1016/j.flora.2017.02.011}.

\leavevmode\hypertarget{ref-buckley_how_2019}{}%
Buckley, T. N. (2019). How do stomata respond to water status? \emph{New
Phytologist} 224, 21--36.
doi:\href{https://doi.org/10.1111/nph.15899}{10.1111/nph.15899}.

\leavevmode\hypertarget{ref-chater_origins_2017}{}%
Chater, C. C. C., Caine, R. S., Fleming, A. J., and Gray, J. E. (2017).
Origins and Evolution of Stomatal Development. \emph{Plant Physiology}
174, 624--638.
doi:\href{https://doi.org/10.1104/pp.17.00183}{10.1104/pp.17.00183}.

\leavevmode\hypertarget{ref-de_boer_optimal_2016}{}%
de Boer, H. J., Price, C. A., Wagner-Cremer, F., Dekker, S. C., Franks,
P. J., and Veneklaas, E. J. (2016). Optimal allocation of leaf epidermal
area for gas exchange. \emph{New Phytologist} 210, 1219--1228.
doi:\href{https://doi.org/10.1111/nph.13929}{10.1111/nph.13929}.

\leavevmode\hypertarget{ref-dow_integrated_2014}{}%
Dow, G. J., Bergmann, D. C., and Berry, J. A. (2014a). An integrated
model of stomatal development and leaf physiology. \emph{New
Phytologist} 201, 1218--1226.

\leavevmode\hypertarget{ref-dow_physiological_2014}{}%
Dow, G. J., Berry, J. A., and Bergmann, D. C. (2014b). The physiological
importance of developmental mechanisms that enforce proper stomatal
spacing in \emph{Arabidopsis thaliana}. \emph{New Phytologist} 201,
1205--1217.
doi:\href{https://doi.org/10.1111/nph.12586}{10.1111/nph.12586}.

\leavevmode\hypertarget{ref-drake_two_2019}{}%
Drake, P. L., Boer, H. J., Schymanski, S. J., and Veneklaas, E. J.
(2019). Two sides to every leaf: Water and \textless span
style="font-variant:Small-caps;"\textgreater CO\textless/span\textgreater{}
\(_{\textrm{2}}\) transport in hypostomatous and amphistomatous leaves.
\emph{New Phytologist} 222, 1179--1187.
doi:\href{https://doi.org/10.1111/nph.15652}{10.1111/nph.15652}.

\leavevmode\hypertarget{ref-dutton_bacterial_2019}{}%
Dutton, C., Hõrak, H., Hepworth, C., Mitchell, A., Ton, J., Hunt, L.,
and Gray, J. E. (2019). Bacterial infection systemically suppresses
stomatal density. \emph{Plant, Cell \& Environment} 42, 2411--2421.
doi:\href{https://doi.org/10.1111/pce.13570}{10.1111/pce.13570}.

\leavevmode\hypertarget{ref-farquhar_stomatal_1982}{}%
Farquhar, G. D., and Sharkey, T. D. (1982). Stomatal Conductance and
Photosynthesis. \emph{Annual Review of Plant Physiology} 33, 317--345.
doi:\href{https://doi.org/10.1146/annurev.pp.33.060182.001533}{10.1146/annurev.pp.33.060182.001533}.

\leavevmode\hypertarget{ref-fawke_oomycete_2015}{}%
Fawke, S., Doumane, M., and Schornack, S. (2015). Oomycete Interactions
with Plants: Infection Strategies and Resistance Principles.
\emph{Microbiology and Molecular Biology Reviews} 79, 263--280.
doi:\href{https://doi.org/10.1128/MMBR.00010-15}{10.1128/MMBR.00010-15}.

\leavevmode\hypertarget{ref-fetter_trade-offs_2019}{}%
Fetter, K. C., Nelson, D. M., and Keller, S. R. (2019). Trade-offs and
selection conflicts in hybrid poplars indicate the stomatal ratio as an
important trait regulating disease resistance.
doi:\href{https://doi.org/10.1101/814046}{10.1101/814046}.

\leavevmode\hypertarget{ref-franks_co_2009}{}%
Franks, P. J., and Beerling, D. J. (2009a). CO \(_{\textrm{2}}\) -forced
evolution of plant gas exchange capacity and water-use efficiency over
the Phanerozoic. \emph{Geobiology} 7, 227--236.
doi:\href{https://doi.org/10.1111/j.1472-4669.2009.00193.x}{10.1111/j.1472-4669.2009.00193.x}.

\leavevmode\hypertarget{ref-franks_maximum_2009}{}%
Franks, P. J., and Beerling, D. J. (2009b). Maximum leaf conductance
driven by CO\(_{\textrm{2}}\) effects on stomatal size and density over
geologic time. \emph{Proceedings of the National Academy of Sciences}
106, 10343--10347.

\leavevmode\hypertarget{ref-franks_plasticity_2009}{}%
Franks, P. J., Drake, P. L., and Beerling, D. J. (2009). Plasticity in
maximum stomatal conductance constrained by negative correlation between
stomatal size and density: An analysis using \emph{Eucalyptus globulus}.
\emph{Plant, Cell \& Environment} 32, 1737--1748.
doi:\href{https://doi.org/10.1111/j.1365-3040.2009.002031.x}{10.1111/j.1365-3040.2009.002031.x}.

\leavevmode\hypertarget{ref-franks_effect_2001}{}%
Franks, P. J., and Farquhar, G. D. (2001). The effect of exogenous
abscisic acid on stomatal development, stomatal mechanics, and leaf gas
exchange in \emph{Tradescantia virginiana}. \emph{Plant Physiology} 125,
935--942.

\leavevmode\hypertarget{ref-franks_mechanical_2007}{}%
Franks, P. J., and Farquhar, G. D. (2007). The Mechanical Diversity of
Stomata and Its Significance in Gas-Exchange Control. \emph{Plant
Physiology} 143, 78--87.
doi:\href{https://doi.org/10.1104/pp.106.089367}{10.1104/pp.106.089367}.

\leavevmode\hypertarget{ref-franks_new_2014}{}%
Franks, P. J., Royer, D. L., Beerling, D. J., Van de Water, P. K.,
Cantrill, D. J., Barbour, M. M., and Berry, J. A. (2014). New
constraints on atmospheric CO \(_{\textrm{2}}\) concentration for the
Phanerozoic: Franks et al.: New constraints on Phanerozoic CO2.
\emph{Geophysical Research Letters} 41, 4685--4694.
doi:\href{https://doi.org/10.1002/2014GL060457}{10.1002/2014GL060457}.

\leavevmode\hypertarget{ref-gutschick_photosynthesis_1984}{}%
Gutschick, V. P. (1984). Photosynthesis model for C\(_{\textrm{3}}\)
leaves incorporating CO\(_{\textrm{2}}\) transport, propagation of
radiation, and biochemistry 1. Kinetics and their parameterization.
\emph{Photosynthetica} 18, 549--568.

\leavevmode\hypertarget{ref-harrison_influence_2019}{}%
Harrison, E. L., Arce Cubas, L., Gray, J. E., and Hepworth, C. (2019).
The influence of stomatal morphology and distribution on photosynthetic
gas exchange. \emph{The Plant Journal}, tpj.14560.
doi:\href{https://doi.org/10.1111/tpj.14560}{10.1111/tpj.14560}.

\leavevmode\hypertarget{ref-hetherington_role_2003}{}%
Hetherington, A. M., and Woodward, F. I. (2003). The role of stomata in
sensing and driving environmental change. \emph{Nature} 424, 901--908.
doi:\href{https://doi.org/10.1038/nature01843}{10.1038/nature01843}.

\leavevmode\hypertarget{ref-hoch_signaling_1987}{}%
Hoch, H. C., Staples, R. C., Whitehead, B., Comeau, J., and Wolf, E. D.
(1987). Signaling for Growth Orientation and Cell Differentiation by
Surface Topography in Uromyces. \emph{Science, New Series} 235,
1659--1662. Available at: \url{http://www.jstor.org/stable/1698314}.

\leavevmode\hypertarget{ref-jones_partitioning_1985}{}%
Jones, H. G. (1985). Partitioning stomatal and non-stomatal limitations
to photosynthesis. \emph{Plant, Cell \& Environment} 8, 95--104.
doi:\href{https://doi.org/10.1111/j.1365-3040.1985.tb01227.x}{10.1111/j.1365-3040.1985.tb01227.x}.

\leavevmode\hypertarget{ref-jordan_using_2014}{}%
Jordan, G. J., Carpenter, R. J., and Brodribb, T. J. (2014). Using
fossil leaves as evidence for open vegetation. \emph{Palaeogeography,
Palaeoclimatology, Palaeoecology} 395, 168--175.
doi:\href{https://doi.org/10.1016/j.palaeo.2013.12.035}{10.1016/j.palaeo.2013.12.035}.

\leavevmode\hypertarget{ref-kiefer_host_2002}{}%
Kiefer, B., Riemann, M., Büche, C., Kassemeyer, H.-H., and Nick, P.
(2002). The host guides morphogenesis and stomatal targeting in the
grapevine pathogen Plasmopara viticola. \emph{Planta} 215, 387--393.
doi:\href{https://doi.org/10.1007/s00425-002-0760-2}{10.1007/s00425-002-0760-2}.

\leavevmode\hypertarget{ref-lehmann_effects_2015}{}%
Lehmann, P., and Or, D. (2015). Effects of stomata clustering on leaf
gas exchange. \emph{New Phytologist} 207, 1015--1025.
doi:\href{https://doi.org/10.1111/nph.13442}{10.1111/nph.13442}.

\leavevmode\hypertarget{ref-mcelwain_using_2016}{}%
McElwain, J. C., Yiotis, C., and Lawson, T. (2016). Using modern plant
trait relationships between observed and theoretical maximum stomatal
conductance and vein density to examine patterns of plant
macroevolution. \emph{New Phytologist} 209, 94--103.
doi:\href{https://doi.org/10.1111/nph.13579}{10.1111/nph.13579}.

\leavevmode\hypertarget{ref-mckown_association_2014}{}%
McKown, A. D., Guy, R. D., Quamme, L., Klápště, J., La Mantia, J.,
Constabel, C. P., El-Kassaby, Y. A., Hamelin, R. C., Zifkin, M., and
Azam, M. S. (2014). Association genetics, geography and ecophysiology
link stomatal patterning in \emph{Populus trichocarpa} with carbon gain
and disease resistance trade-offs. \emph{Molecular Ecology} 23,
5771--5790.
doi:\href{https://doi.org/10.1111/mec.12969}{10.1111/mec.12969}.

\leavevmode\hypertarget{ref-mclachlan_gate_2014}{}%
McLachlan, D. H., Kopischke, M., and Robatzek, S. (2014). Gate control:
Guard cell regulation by microbial stress. \emph{New Phytologist} 203,
1049--1063.
doi:\href{https://doi.org/10.1111/nph.12916}{10.1111/nph.12916}.

\leavevmode\hypertarget{ref-melotto_plant_2006}{}%
Melotto, M., Underwood, W., Koczan, J., Nomura, K., and He, S. Y.
(2006). Plant Stomata Function in Innate Immunity against Bacterial
Invasion. \emph{Cell} 126, 969--980.
doi:\href{https://doi.org/10.1016/j.cell.2006.06.054}{10.1016/j.cell.2006.06.054}.

\leavevmode\hypertarget{ref-melotto_stomatal_2017}{}%
Melotto, M., Zhang, L., Oblessuc, P. R., and He, S. Y. (2017). Stomatal
Defense a Decade Later. \emph{Plant Physiology} 174, 561--571.
doi:\href{https://doi.org/10.1104/pp.16.01853}{10.1104/pp.16.01853}.

\leavevmode\hypertarget{ref-metcalfe_anatomy_1950}{}%
Metcalfe, C. R., and Chalk, L. (1950). \emph{Anatomy of the
dicotyledons, Vols. 1 \& 2}. First. Oxford: Oxford University Press.

\leavevmode\hypertarget{ref-meurer_sympy:_2017}{}%
Meurer, A., Smith, C. P., Paprocki, M., Čertík, O., Kirpichev, S. B.,
Rocklin, M., Kumar, A., Ivanov, S., Moore, J. K., Singh, S., et al.
(2017). SymPy: Symbolic computing in Python. \emph{PeerJ Computer
Science} 3, e103.
doi:\href{https://doi.org/10.7717/peerj-cs.103}{10.7717/peerj-cs.103}.

\leavevmode\hypertarget{ref-monteith_principles_2013}{}%
Monteith, J. L., and Unsworth, M. H. (2013). \emph{Principles of
environmental physics: Plants, animals, and the atmosphere}. 4th ed.
Amsterdam ; Boston: Elsevier/Academic Press.

\leavevmode\hypertarget{ref-morison_lateral_2005}{}%
Morison, J. I. L., Emily Gallouët, Lawson, T., Cornic, G., Herbin, R.,
and work(s): N. R. B. R. (2005). Lateral Diffusion of CO₂ in Leaves Is
Not Sufficient to Support Photosynthesis. \emph{Plant Physiology} 139,
254--266. Available at: \url{http://www.jstor.org/stable/4281859}.

\leavevmode\hypertarget{ref-mott_adaptive_1984}{}%
Mott, K. A., Gibson, A. C., and O'Leary, J. W. (1984). The adaptive
significance of amphistomatic leaves. \emph{Plant, Cell \& Environment}
5, 455--460.

\leavevmode\hypertarget{ref-muir_making_2015}{}%
Muir, C. D. (2015). Making pore choices: Repeated regime shifts in
stomatal ratio. \emph{Proceedings of the Royal Society B: Biological
Sciences} 282, 20151498.
doi:\href{https://doi.org/10.1098/rspb.2015.1498}{10.1098/rspb.2015.1498}.

\leavevmode\hypertarget{ref-muir_morphological_2014}{}%
Muir, C. D., Hangarter, R. P., Moyle, L. C., and Davis, P. A. (2014).
Morphological and anatomical determinants of mesophyll conductance in
wild relatives of tomato ( \emph{solanum} sect. \emph{Lycopersicon} ,
sect. \emph{Lycopersicoides} ; Solanaceae). \emph{Plant, Cell \&
Environment} 37, 1415--1426.
doi:\href{https://doi.org/10.1111/pce.12245}{10.1111/pce.12245}.

\leavevmode\hypertarget{ref-murray_consistent_2019}{}%
Murray, M., Soh, W. K., Yiotis, C., Spicer, R. A., Lawson, T., and
McElwain, J. C. (2019). Consistent relationship between field-measured
stomatal conductance and theoretical maximum stomatal conductance in
C\(_{\textrm{3}}\) woody angiosperms in four major biomes.
\emph{International Journal of Plant Sciences}, 706260.
doi:\href{https://doi.org/10.1086/706260}{10.1086/706260}.

\leavevmode\hypertarget{ref-murray_plant_2016}{}%
Murray, R. R., Emblow, M. S. M., Hetherington, A. M., and Foster, G. D.
(2016). Plant virus infections control stomatal development.
\emph{Scientific Reports} 6, 34507.
doi:\href{https://doi.org/10.1038/srep34507}{10.1038/srep34507}.

\leavevmode\hypertarget{ref-parkhurst_diffusion_1994}{}%
Parkhurst, D. F. (1994). Diffusion of
CO\textless sub\textgreater2\textless\textbackslash sub\textgreater{}
and Other Gases Inside Leaves. \emph{New Phytologist} 126, 449--479.
Available at: \url{http://www.jstor.org/stable/2557929}.

\leavevmode\hypertarget{ref-parkhurst_adaptive_1978}{}%
Parkhurst, D. F. (1978). The Adaptive Significance of Stomatal
Occurrence on One or Both Surfaces of Leaves. \emph{The Journal of
Ecology} 66, 367.
doi:\href{https://doi.org/10.2307/2259142}{10.2307/2259142}.

\leavevmode\hypertarget{ref-parkhurst_intercellular_1990}{}%
Parkhurst, D. F., and Mott, K. A. (1990). Intercellular Diffusion Limits
to CO \(_{\textrm{2}}\) Uptake in Leaves: Studies in Air and Helox.
\emph{Plant Physiology} 94, 1024--1032.
doi:\href{https://doi.org/10.1104/pp.94.3.1024}{10.1104/pp.94.3.1024}.

\leavevmode\hypertarget{ref-parlange_stomatal_1970}{}%
Parlange, J.-Y., and Waggoner, P. E. (1970). Stomatal Dimensions and
Resistance to Diffusion. \emph{Plant Physiology} 46, 337--342.
doi:\href{https://doi.org/10.1104/pp.46.2.337}{10.1104/pp.46.2.337}.

\leavevmode\hypertarget{ref-peat_comparative_1994}{}%
Peat, H. J., and Fitter, A. H. (1994). A comparative study of the
distribution and density of stomata in the British flora.
\emph{Biological Journal of the Linnean Society} 52, 377--393.

\leavevmode\hypertarget{ref-r_core_team_r:_2019}{}%
R Core Team (2019). \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing
Available at: \url{http://www.R-project.org/}.

\leavevmode\hypertarget{ref-sack_developmental_2016}{}%
Sack, L., and Buckley, T. N. (2016). The developmental basis of stomatal
density and flux. \emph{Plant Physiology}, pp.00476.2016.
doi:\href{https://doi.org/10.1104/pp.16.00476}{10.1104/pp.16.00476}.

\leavevmode\hypertarget{ref-sack_hydrology_2003}{}%
Sack, L., Cowan, P. D., Jaikumar, N., and Holbrook, N. M. (2003). The
'hydrology' of leaves: Co-ordination of structure and function in
temperate woody species. \emph{Plant, Cell and Environment} 26,
1343--1356.
doi:\href{https://doi.org/10.1046/j.0016-8025.2003.01058.x}{10.1046/j.0016-8025.2003.01058.x}.

\leavevmode\hypertarget{ref-salisbury_causes_1928}{}%
Salisbury, E. J. (1928). On the Causes and Ecological Significance of
Stomatal Frequency, with Special Reference to the Woodland Flora.
\emph{Philosophical Transactions of the Royal Society B: Biological
Sciences} 216, 1--65.
doi:\href{https://doi.org/10.1098/rstb.1928.0001}{10.1098/rstb.1928.0001}.

\leavevmode\hypertarget{ref-tateda_host_2019}{}%
Tateda, C., Obara, K., Abe, Y., Sekine, R., Nekoduka, S., Hikage, T.,
Nishihara, M., Sekine, K.-T., and Fujisaki, K. (2019). The Host Stomatal
Density Determines Resistance to \emph{Septoria gentianae} in Japanese
Gentian. \emph{Molecular Plant-Microbe Interactions} 32, 428--436.
doi:\href{https://doi.org/10.1094/MPMI-05-18-0114-R}{10.1094/MPMI-05-18-0114-R}.

\leavevmode\hypertarget{ref-ticha_photosynthetic_1982}{}%
Tichá, I. (1982). Photosynthetic characteristics during ontogenesis of
leaves 7. Stomata density and sizes. \emph{Photosynthetica} 16,
375--471.

\leavevmode\hypertarget{ref-underwood_role_2007}{}%
Underwood, W., Melotto, M., and He, S. Y. (2007). Role of plant stomata
in bacterial invasion. \emph{Cellular Microbiology} 9, 1621--1629.
doi:\href{https://doi.org/10.1111/j.1462-5822.2007.00938.x}{10.1111/j.1462-5822.2007.00938.x}.

\leavevmode\hypertarget{ref-weiss_untersuchungen_1865}{}%
Weiss, A. (1865). Untersuchungen über die Zahlen- und
Grössenverhältnisse der Spaltöffnungen. \emph{Jahrbücher für
Wissenschaftliche Botanik} 4, 125--196.

\leavevmode\hypertarget{ref-xie_r_2018}{}%
Xie, Y., Allaire, J. J., and Grolemund, G. (2018). \emph{R Markdown: The
definitive guide}. Boca Raton: Taylor \& Francis, CRC Press.

\leavevmode\hypertarget{ref-zeng_plant_2010}{}%
Zeng, W., Melotto, M., and He, S. Y. (2010). Plant stomata: A checkpoint
of host immunity and pathogen virulence. \emph{Current Opinion in
Biotechnology} 21, 599--603.
doi:\href{https://doi.org/10.1016/j.copbio.2010.05.006}{10.1016/j.copbio.2010.05.006}.

\end{document}

### 2019-12-10 r/03_plot-spatially-explicit.R

@@ -56,4 +56,5 @@ f3b
 f3 <- plot_grid(f3a, f3b, labels = "auto", align = "hv", axis = "lr")
 
 plot_grid(l, f3, ncol = 1, rel_heights = c(0.1, 0.9))
+# ggsave("figures/fig3.jpeg", width = 6.5, height = 6.5, units = "in")
 ggsave("figures/fig3.pdf", width = 6.5, height = 6.5, units = "in")

### 2019-12-10 r/05_plot-scaling.R

@@ -53,4 +53,5 @@ fig4 <- ggplot(df, aes(D, S, group = p_colonize)) +
 
 fig4
 
+# ggsave("figures/fig4.jpeg", w = 6.5, h = 3, units = "in")
 ggsave("figures/fig4.pdf", w = 6.5, h = 3, units = "in")

### 2019-12-10 r/functions.R

@@ -9,25 +9,6 @@ gmax <- function(b, m, d, s) {
   set_units(b * m * d * sqrt(s), mol / m ^ 2 / s)
 }
 
-# Calculate D_w as a function temperature and pressure
-get_Dw <- function(D_0, Temp, eT, P) {
-  
-  D_0 %<>%
-    set_units(m ^ 2 / s) %>%
-    drop_units()
-  
-  Temp %<>%
-    set_units(K) %>%
-    drop_units()
-  
-  P %<>%
-    set_units(kPa) %>%
-    drop_units()
-  
-  set_units(D_0 * (Temp / 273.15) ^ eT * (101.3246 / P), m ^ 2 / s)
-  
-}
-
 # Functions for converting from D to U, R to S ----
 D2U <- function(D) {
 

### 2020-07-13 ms/ms.Rmd
r1

@@
 ---
-title: "A stomatal model of anatomical tradeoffs between photosynthesis and pathogen defense"
-runtitle: "Stomata tradeoff photosynthesis for defense"
+title: "A stomatal model of anatomical tradeoffs between gas exchange and pathogen colonization"
+runtitle: "Stomatal anatomy and pathogen colonization"
 documentclass: frontiersSCNS
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 
-Stomatal pores control both leaf gas exchange and are an entry for many plant pathogens, setting up the potential for tradeoffs between photosynthesis and defense.
-To prevent colonization and limit infection, plants close their stomata after recognizing pathogens.
-In addition to closing stomata, anatmoical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes would permanently reduce the gas exchange capacity for the life of the leaf.
-I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as a function of stomatal size and density, anatomical traits which determine maximum rates of gas exchange.
-The model predicts greater stomatal size or density increases colonization, but the effect is most pronounced when stomatal cover is low.
+Stomatal pores control both leaf gas exchange and are one route for infection of internal plant tissues by many foliar pathogens, setting up the potential for tradeoffs between photosynthesis and defense.
+Anatomical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes could permanently reduce the gas exchange capacity for the life of the leaf.
+I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as a function of stomatal size and density, anatomical traits which partially determine maximum rates of gas exchange.
+The model predicts greater stomatal size or density increases the probability of colonization, but the effect is most pronounced when the fraction of leaf surface covered by stomata is low.
 I also derived scaling relationships between stomatal size and density that preserves a given probability of colonization.
-These scaling relationships set up a potential conflict between maximizing defense and minimizing stomatal cover.
-To my knowledge, this is the first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy.
-It makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants.
-The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for accurate predictions.
+These scaling relationships set up a potential anatomical conflict between limiting pathogen colonization and minimizing the fraction of leaf surface covered by stomata.
+Although a connection between gas exchange and pathogen defense has been suggested empirically, this is the first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy.
+A limitation of the model is that it does not include variation in innate immunity and stomatal closure in response to pathogens.
+Nevertheless, the model makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants.
+The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for precise predictions.
 
@@
 <!-- overview -->
-Stomata evolved to regulate gas exchange in and out of the leaf [@hetherington_role_2003; @berry_stomata:_2010; @chater_origins_2017], but many plant pathogens take advantage of these chinks in the leaf cuticular armor to infect prospective hosts [@zeng_plant_2010; @mclachlan_gate_2014; @melotto_stomatal_2017].
-The density and size of stomata set the anatomical maximum rate of stomatal conductance to CO$_2$ and water vapor [@brown_static_1900; @parlange_stomatal_1970; @franks_effect_2001; @franks_maximum_2009; @lehmann_effects_2015; @sack_developmental_2016; @harrison_influence_2019], but the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019].
-Many plant pathogens, including viruses [@murray_plant_2016], bacteria [@melotto_plant_2006; @underwood_role_2007], protists [@fawke_oomycete_2015], and fungi [@hoch_signaling_1987; @zeng_plant_2010] use stomatal pores to gain entry into the leaf.
-Since stomatal conductance is a major limitation on photosynthesis [@farquhar_stomatal_1982; @jones_partitioning_1985] while pathogens reduce fitness, this sets up a potential tradeoff between increased photosynthesis and defense against pathogens.
-Although there have been many empirical studies on the effect that pathogens have on stomata, there is no theoretical framework in which to place these findings.
-Lack of a theoreatical framework makes it difficult to answer general questions about how selection for pathogen defense constrains maximum rates of gas exchange.
+Stomata evolved to regulate gas exchange in and out of the leaf [@hetherington_role_2003; @berry_stomata:_2010; @chater_origins_2017], but many foliar pathogens take advantage of these chinks in the leaf cuticular armor to infect prospective hosts [@zeng_plant_2010; @mclachlan_gate_2014; @melotto_stomatal_2017].
+The stomatal and mesophyll conductance to CO$_2$ are two major limits to photosynthesis [@lawson_coordination_2018; @flexas_co2_2018] that are partially determined by stomatal anatomy.
+Since CO$_2$ conductance limits photosynthesis [@farquhar_stomatal_1982; @jones_partitioning_1985] and pathogen infection can reduce fitness [@gilbert_evolutionary_2002], this sets up a potential tradeoff between increased photosynthesis and defense against pathogens mediated by stomatal anatomy [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019].
+For example, plants could increase photosynthetic rate by developing more stomata, but more stomata could result in more pathogen colonization.
+The optimal stomatal density, size, and arrangement on the leaf will depend on the fitness gains from increased gas exchange and fitness losses imposed by foliar pathogens, both of which depend on the environment.
+In the next two paragraphs I will review the relationship between stomatal anatomy, gas exchange, and foliar pathogen colonization.
+Then I will discuss why two anatomical traits, stomatal size and density, might be crucial components of a broader tradeoff between photosynthesis and pathogen defense.
 
-<!-- paragraph on stomatal anatomy and gsmax -->
-Stomatal anatomy is the key link between gas exchange and pathogen colonization.
-The density and size of stomata not only determines the theoretical maximum stomatal conductance (\gsmax), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019].
-Therefore, I use anatomical \gsmax{} as a proxy for \gsop{} and do not address dynamic changes in stomatal aperture (see [Discussion]).
-@harrison_influence_2019 recently reviewed the relationship between stomatal anatomy and gas exchange in detail.
+<!-- anatomy and gas exchange -->
+The stomatal density and maximum pore area set an anatomical upper limit on stomatal conductance [@brown_static_1900; @parlange_stomatal_1970; @franks_effect_2001; @franks_maximum_2009; @lehmann_effects_2015; @sack_developmental_2016; @harrison_influence_2019], but stomatal shape, distribution, and patterning also affect gas exchange.
+Smaller guard cells and dumbbell-shaped stomata of grasses can respond faster to environmental changes [@drake_smaller_2013], but responsiveness is further modulated by subsidiary cell anatomy and physiology [@franks_mechanical_2007; @raissig_mobile_2017; @gray_flanking_2020].
+Stomatal clustering reduces gas exchange and photosynthesis because adjacent stomata interfere with one another [@dow_physiological_2014], diffusion shells overlap [@lehmann_effects_2015], and limitations on lateral diffusion of CO$_2$ in the mesophyll [@lawson_stomatal_2014 and references therein].
+However, sparse clusters of small stomata could allow a leaf with low rates of gas exchange to have faster stomatal response compared to a leaf with large, low-density stomata [@papanatsiou_stomatal_2017].
+Leaves with stomata on both lower and upper surfaces (amphistomatic or amphistomatous) supply more CO$_2$ to the mesophyll than hypostomatic (aka hypostomatous) leaves that only have stomata on the lower surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990; @oguchi_leaf_2018].
+In addition to anatomy, the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019].
+For example, stomata typically open during the day and close at night in C$_3$/C$_4$ plants, but the opposite is true for CAM plants.
+Shade, high vapor pressure deficits, dry soil and other factors can cause stomata to (partially) close even in the middle of the day.
+Variation in how stomata respond to internal and external signals may explain as much of the variation in gas exchange across leaves as anatomy [@lawson_stomatal_2014].
 
-<!-- paragraph on stomatal anatomy and pathogen defense -->
-Many pathogens rely on stomata to gain entry into the leaf, but it is unclear how anatomical traits that determine \gsmax{} (size and density) affect the ability of pathogens to colonize leaves.
-The impact of pathogens on host fitness is complex, but after a pathogen reaches a host, the first major step is colonization.
-Once infected, a pathogen can reduce fitness.
-Susceptible hosts can lose much of their biomass or die, but even resistant hosts must allocate resources to defense or reduce photosynthesis because of necrosis around sites of infection.
-Plants can limit colonization physiologically by closing stomata after they recognize pathogens, called stomatal defense [@melotto_stomatal_2017].
-Anatomy may be another layer of defense; plants can reduce pathogen colonization by developing leaves with lower stomatal density and/or size.
-Infection increases in leaves with higher stomatal density [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019], which suggests that both anatomical and physiological responses (stomatal closure) affect host colonization.
-By the same logic, if stomatal density were held constant, larger stomata should also increase colonization because they occupy more area.
-One key difference between physiological and anatomical defenses is that stomatal closure reduces gas exchange transiently, whereas anatomy constrains gas exchange throught the life of the leaf.
-Because we do not understand the relationship between stomatal anatomy and colonization well, we cannot predict the relationship between \gsmax{} and infection.
+<!-- anatomy and pathogens -->
+Many types of foliar pathogens, including viruses [@murray_plant_2016], bacteria [@melotto_plant_2006; @underwood_role_2007], protists [@fawke_oomycete_2015], and fungi [@hoch_signaling_1987; @zeng_plant_2010] use stomatal pores to gain entry into the leaf.
+For example, rust fungi hyphae recognize the angle at which guard cells project from the leaf surface and use it as a cue for appressorium formation [@allen_appressorium_1991].
+Oomycete pathogens can target open stomata on a leaf [@kiefer_host_2002].
+Plants can limit colonization through innate immunity, called stomatal defense [recently reviewed in @melotto_stomatal_2017], by closing stomata after they recognize microbe-associated molecular patterns (MAMPs) on pathogen cells.
+Some bacterial pathogens have responded by evolving the ability to prevent stomatal closure, increasing their colonization of the leaf interior [@melotto_plant_2006].
+In addition to stomatal closure, anatomical changes in stomatal density and/or size might provide another layer of defense against pathogen colonization.
+For example, infection increases in leaves with higher stomatal density [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019].
+The positive effect of stomatal density on infection suggests that infection is limited by the number or size of locations for colonization, meaning that many individual pathogens must usually be unable to find stomata or other suitable locations for colonization.
+This is actually somewhat surprising given the ability of some pathogens to search for and sense stomata (see above).
 
-<!-- paragraph on size-density scaling? -->
-If stomatal size and density affect pathogen colonization, selection to limit colonization may shape stomatal size-density scaling relationships.
+<!-- Two paragraphs focusing the scope of this study -->
+Stomatal anatomy could be a key link between gas exchange and pathogen colonization.
+Although many anatomical factors and stomatal movement affect gas exchange (see above), here I focus on the density and size of stomata in a hypostomatous leaf.
+Stomatal size refers to both the area of guard cells when fully open, from which one can calculate the pore area for gas exchange (see [Model]).
+For simplicity, I model a hypostomatous leaf, but consider the implications for amphistomatous leaves in the [Discussion].
+Stomatal size and density not only determine the theoretical maximum stomatal conductance (\gsmax), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019].
+\gsop{} is the actual stomatal conductance of plants in the field and is almost always below \gsmax{} because stomata are usually not fully open.
+Although they are not the same, the strong empirical relationship between \gsmax{} and \gsop{} means that anatomical \gsmax{} can be used as a proxy for \gsop{} without explicitly modeling dynamic changes in stomatal aperture (see [Discussion]).
+Stomatal size and density have also been measured on many more species than stomatal responsiveness, which may make it easier to test predictions.
+
+After a pathogen reaches a host, it must survive on the leaf surface and colonize the interior [@beattie_secret_1995; @tucker_surface_2001].
+For analytical tractability, I restrict the focus here to colonization by a pathogen using a random search on a leaf without stomatal defense (i.e.
+a leaf that cannot recognize pathogens and close stomata).
+Obviously, these simplifications ignore a lot of important plant-pathogen interaction biology.
+In the [Discussion], I delve further into these limitations and suggest future work to overcome these limitations.
+In order for pathogen-mediated selection on stomatal anatomy, I assume that the pathogen reduces host fitness once it colonizes [@gilbert_evolutionary_2002].
+Susceptible hosts can lose much of their biomass or die, but even resistant hosts must allocate resources to defense or reduce photosynthesis because of defoliation, biotrophy, or necrosis around sites of infection [@bastiaans_ratio_1991; @mitchell_trophic_2003].
+
+<!-- Purpose of study -->
+The purpose of this study is to develop a theoretical framework to test whether variation in stomatal size and density arises from a tradeoff between gas exchange and pathogen colonization.
+Since stomatal size and density affect both gas exchange and pathogen colonization, selection to balance these competing demands could shape stomatal size-density scaling relationships.
 Botanists have long recognized that stomatal size and density are inversely correlated [@weiss_untersuchungen_1865; @ticha_photosynthetic_1982; @hetherington_role_2003; @sack_hydrology_2003; @franks_co_2009; @brodribb_unified_2013; @de_boer_optimal_2016], but the evolutionary origin of this relationship is not yet known.
-Here I argue that pathogens could shape selection on this relationship.
-Explanations for inverse size-density scaling are usually cast in terms of preserving \gsmax{} and/or total epidermal area allocated to stomata [\fs; @de_boer_optimal_2016] because there are many combinations of stomatal size and density that have same \gsmax{} or same \fs{}:
+Here I argue that deleterious effects of pathogen infection could shape selection on this relationship.
+Explanations for inverse size-density scaling are usually cast in terms of preserving \gsmax{} and/or stomatal cover (\fs), defined at the fraction of epidermal area allocated to stomata [@de_boer_optimal_2016], because there are many combinations of stomatal size and density that have same \gsmax{} or same \fs{}:
 
@@
 
+$D$ and $S$ are stomatal stomatal density and size, respectively (see Table 1 for a glossary of mathematical symbols and units).
 $b$ and $m$ are assumed to be biophysical and morphological constants, *sensu* @sack_developmental_2016 (see [Supplementary Material]).
+\fs{} is proportional to the more widely used stomatal pore area index [@sack_hydrology_2003; see [Supplementary Material]].
 If size and density also affect pathogen colonization, then selection from foliar pathogens could significantly alter the size-density scaling relationship.
@@
 How would adding pathogens alter these predicted scaling relationships?
-For simplicity, imagine two environments, one without foliar pathogens and one with lots.
+For simplicity, consider two environments, one without foliar pathogens and one with lots.
 In the absence of foliar pathogens, we expect size-density scaling to preserve \gsmax{}, \fs{}, or some least-cost combination of them.
 What happens when we introduce pathogens?
-Assuming that stomatal size and density increase pathogen colonization, then selection will favor reduced size and/or density.
+If stomatal size and density increase pathogen colonization, then selection will favor reduced size and/or density.
 This would change the intercept $\alpha$ but not the slope.
@@
 If the probability of colonization is proportional to the product of *areal* stomatal size ($S$) and density ($\propto D S$ as for \fs) then it has the same effect on the slope as \fs{} because there are many combinations of $D$ and $S$ that have same probability of colonization.
-Alternatively, the probability of colonization may have a different scaling relationsip (neither 0.5 nor 1) or may be nonlinear on a log-log scale.
+Alternatively, the probability of colonization may have a different scaling relationship (neither 0.5 nor 1) or may be nonlinear on a log-log scale.
 Unlike \gsmax{} and \fs{}, we do not have theory to predict a stomatal size-density relationship that preserves the probability of colonization.
@@
 In summary, the physical relationship between stomatal size, density, and conductance is well established [@harrison_influence_2019].
-The same traits likely affect the probability of pathogen colonization, but we do not have a theoretical model that makes quantitative predictions.
+Size and density also likely affect the probability of pathogen colonization, but we do not have a theoretical model that makes quantitative predictions.
 The inverse stomatal size-density relationship has usually been explained in terms of preserving stomatal conductance and/or stomatal cover, but selection by pathogens might alter scaling.
-To address these gaps, the goals of this study are to 1) introduce a spatially explicit model pathogen colonization on the leaf surface; 2) use the model to predict the relationship between \gsmax, \fs{}, and the probability of colonization; 3) work out what these relationships predict about stomatal size-density scaling.
+To address these gaps, the goals of this study are to 1) introduce a spatially explicit model pathogen colonization on the leaf surface; 2) use the model to predict the relationship between \gsmax, \fs, and the probability of colonization; 3) work out what these relationships predict about stomatal size-density scaling.
+I analyzed an idealized, spatially explicit [Model] of how a pathogen lands on a leaf and finds a stomate to colonize the leaf using a random search.
+To my knowledge, this is the first model that makes quantitative predictions about the relationship between stomatal anatomy, the probability of colonization, and their impact on stomatal size-density scaling.
+
+<!-- scraps -->
+<!-- {By the same logic, if stomatal density were held constant, larger stomata should also increase colonization because they occupy more area, assuming stomatal responses are similar.
+One key difference between physiological and anatomical defenses is that stomatal closure reduces gas exchange transiently, whereas anatomy constrains gas exchange throughout the life of the leaf.
+-->
+
+<!-- MOVE ELSEWHERE??: Although there have been many empirical studies on the effect that pathogens have on stomata, there is no theoretical framework in which to place these findings.
+Lack of a theoretical framework makes it difficult to answer general questions about how selection for pathogen defense constrains maximum rates of gas exchange.
+Although there is an empirical relationship between stomatal density and pathogen infection (see below), a theoretical framework based on anatomy may help explain empirical observations and generate new predictions.
+-->
+
+<!-- Because we do not understand the relationship between stomatal anatomy and probability of colonization well, we cannot predict the relationship between \gsmax{} and infection.} -->
 
@@
 
-In this section, I introduce a spatially explicit model of pathogen colonization on a leaf surface.
+In this section, I introduce a spatially explicit model of successful pathogen colonization on a leaf surface.
 I explain the model structure and assumptions here; the [Materials and Methods] section below describes how I analyzed the model to address the goals of the study.
 For generality, I refer to a generic ``pathogen'' that lands on leaf and moves to a stomate.
-The model is agnostic to the type of pathogen (virus, bacterium, fungus, etc.) and the specific biological details of how it moves (biotrophy).
-For example, motile bacterial cells can land and move around where fungi may germinate from a cyst and grow until they form an appresorium for infection.
-These very different biotropic movements on the leaf are treated identically here.
-I used Sympy version 1.3 [@meurer_sympy:_2017] for symbolic derivations.
+The model is agnostic to the type of pathogen (virus, bacterium, fungus, etc.) and the specific biological details of how it moves.
+For example, motile bacterial cells can land and move around [@beattie_secret_1995] whereas fungi may germinate from a cyst and grow until they form an appresorium for infection [@tucker_surface_2001].
+These very different tropic movements on the leaf are treated identically here.
+I do not model photosynthesis explicitly, but assume that stomatal conductance limits carbon fixation, even though the relationship is nonlinear.
+I used Sympy version 1.6.1 [@meurer_sympy:_2017] for symbolic derivations.
 
@@
  Symbol | R      | Units    | Description
---------|--------|--------|-----------------------------------------------------
- $D,d$              | `D, d`    | $\textrm{mm}^{-2}$    | stomatal density ($d = \text{log}~D$)
- $f_s$              | `f_s`     | none                  | stomatal cover ($f_s = D S$)
- $g_\text{s,max}$   | `g_smax`  | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | anatomical maximum stomatal conductance 
+--------|--------|----------|---------------------------------------------------
+ $D$                | `D`       | $\textrm{mm}^{-2}$    | stomatal density
+ $d$                | `d`       | $\textrm{mm}^{-2}$    | stomatal density (log-scale, $d = \text{log}~D$)
+ \fs                | `f_s`     | none                  | stomatal cover ($f_s = D S$)
+ $g_\text{s,max}$   | `g_smax`  | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | theoretical maximum stomatal conductance 
  $g_\text{s,op}$    | `g_sop`   | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | operational stomatal conductance 
  $H$                | `H`       | $\mu\textrm{m} ^{-1}$ | death rate of pathogen on leaf surface
- $R$                | `R`       | $\mu\textrm{m}$       | stomatal radius ($S = 2 \pi R ^ 2$)
- $S,s$              | `S, s`    | $\mu\textrm{m}^2$     | stomatal size ($s = \text{log}~S$)
- $\theta_i$         | `theta_i` | radians               | angles between pathogen ($x_p,y_p$) and lines tangent to the circumfrence of stomate $i$
+ $R$                | `R`       | $\mu\textrm{m}$       | stomatal radius ($S = \pi R ^ 2$)
+ $S$                | `S`       | $\mu\textrm{m}^2$     | stomatal size
+ $s$                | `s`       | $\mu\textrm{m}^2$     | stomatal size (log-scale, $s = \text{log}~S$)
+ $\theta_i$         | `theta_i` | radians               | angles between pathogen ($x_p,y_p$) and lines tangent to the circumference of stomate $i$
  $U$                | `U`       | $\mu\textrm{m}$       | interstomatal distance
@@
 Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015].
-Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$.
+Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$ on the abaxial surface only, since most leaves are hypostomatoues [@muir_making_2015; but see [Discussion]].
 This assumption ignores veins, trichomes, and within-leaf variation in stomatal density.
 Stomata are therefore arrayed in an evenly spaced grid (Figure \ref{fig:fig1}a).
-The interstomatal distance $U$, measured as the distance from the center of one stomata to the next, is the maximal diagonal of the hexagon in $\mu\textrm{m}$ that forms an equal area boundary between neighbording stomata.
+The interstomatal distance $U$, measured as the distance from the center of one stomata to the next, is the maximal diagonal of the hexagon in $\mu\textrm{m}$ that forms an equal area boundary between neighboring stomata.
 The area of a hexagon is $A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2$.
@@
 @parkhurst_diffusion_1994 described this result previously.
-I also make the simplifying assumption that stomata are perfectly circular with radius $R$.
-This may be approximately true for fully open stomata with kidney-shaped guard cells.
-Although I assume stomata are circular here, in calculating \gsmax, I assume typical allometric relationships between length, width, and pore area  (see [Supplementary Material]).
+I also make the simplifying assumption that stomata are perfectly circular with radius $R$ when fully open.
+This may be approximately true for fully open stomata with kidney-shaped guard cells [@sack_developmental_2016 and references therein].
+Although I assume stomata are circular here, in calculating \gsmax, I assume typical allometric relationships between length, width, and pore area  [@sack_developmental_2016; see [Supplementary Material]].
 
@@
 The circles represent idealized stomata; the grey lines between them are for visualization.
-\textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the area.
+\textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the triangle.
 \textbf{d.} The model assumes that a pathogen, depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves at the correct angle, depicted by the grey polygons.}
@@
 
-Now imagine that a pathogen lands at a uniform random position within the focal region and must arrive at a stomate to colonize.
+Since stomata are arrayed in a homogeneous grid, we can focus on single focal triangle (Figure \ref{fig:fig1}b-c).
+Suppose that an individual pathogen (e.g.
+bacterial cell or fungal spore) lands at a uniform random position within the focal triangle and must arrive at a stomate to colonize.
 If it lands on a stomate, then it infects the leaf with probability 1; if it lands between stomata, then it infects the leaf with probability $p_\text{locate}$.
@@
 If there is a high density of stomata and/or large stomata, the probability of locating a stomate increases.
-By assuming that stomata form an equilateral triangular grid (see above), we can extrapolate what happens in a focal triangle (Figure \ref{fig:fig1}b) by symmetry.
+By assuming that stomata form an equilateral triangular grid (see above), we can extrapolate what happens in the focal triangle (Figure \ref{fig:fig1}b) by symmetry.
 Further, since an equilateral triangle can be broken up into six identical units (Figure \ref{fig:fig1}c), we can simply calculate $p_\text{locate}$ in this focal area.
-This implicitly assumes that the probability of colonzing stomata outside the focal area is 0 because they are too far away.
+This implicitly assumes that the probability of colonizing stomata outside the focal area is 0 because they are too far away.
+This assumption may be unrealistic for larger pathogens, such as fungi, whose hyphae can travel longer distances on the leaf surface [@perez-martin_tropic_2012].
+In [Appendix 1: Spatially implicit model] I derive a simpler, but spatially *implicit* model that relaxes the assumption the pathogens must colonize a stomate within their focal triangle.
 
-Imagine that the pathogens lands in position ($x_p,y_p$) within the triangle.
+Consider a pathogen that lands in position ($x_p,y_p$) within the triangle.
 The centroid of the triangle is at position ($x_c,y_c$) and a reference stomate is at position $(0,0)$ (Figure \ref{fig:fig2}a).
@@
 \caption{\textbf{Spatial representation of stomata and pathogen.} \textbf{a.} The pathogen starts at a uniform random position within the focal region denoted ($x_p,y_p$).
-Within the focal triangle, the reference stomata is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$.
-\textbf{b.} Within the focal region, a pathogen can land within the stomate (white cirlce with grey outline and radius $R$) or in the grey area.
+Within the focal triangle, the reference stomate is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$.
+\textbf{b.} Within the focal region, a pathogen can land within the stomate (white circle with grey outline and radius $R$) or in the grey area.
 The outer borders of this area are shown and depend on $R$ and $U$.
@@
 I further assumed that the longer distance a pathogen must travel, the less likely it would be to locate a stomate.
-For example, if stomata are at very low density, then a pathogen may die before it reaches a stomate because of UV, dessication, or another factor.
+For example, if stomata are at very low density, then a pathogen may die before it reaches a stomate because of UV, desiccation, or another factor.
 I included this effect by assuming the probability of reaching a stomate declines exponentially at rate $H$ with the Euclidean distance $v_i(x_p,y_p)$ between the pathogen location and the edge of stomata $i$, which is distance $R$ from its center at $x_i,y_i$:
@@
 
-The probability of locating a stomate as a function of $x_p$ and $y_p$ ($f_\text{locate}(x_p,y_p)$) is the sum of the angles divided by $2\pi$, discounted by their distance from the stomate:
+The probability of locating a stomate as a function of pathogen position ($x_p$ and $y_p$) is the sum of the angles divided by $2\pi$, discounted by their distance from the stomate:
 
@@
 
-When $H = 0$, $p_\text{locate}$ is the fraction of angles that lead from $(x_p,y_p)$ to a stomate.
+When there is no pathogen death ($H = 0$), $p_\text{locate}$ is the fraction of angles that lead from $(x_p,y_p)$ to a stomate.
 When $H > 0$, $p_\text{locate}$ is proportional to this fraction, but less than it depending on stomatal density, size, and starting location of the pathogen.
@@
 To obtain the average $p_\text{locate}$, we must integrate $f_\text{locate}(x_p,y_p)$ over all possible starting positions $(x_p,y_p)$ within the focal area.
-The focal area is a 30-60-90 triangle with vertices at the center of the reference stomate $(0,0)$, the midpoint of baseline $(U/2,0)$, and the centroid of the focal trinagle $(U/2, \sqrt{3} / 6 U)$ (Figure \ref{fig:fig1}c).
+The focal area is a 30-60-90 triangle with vertices at the center of the reference stomate $(0,0)$, the midpoint of baseline $(U/2,0)$, and the centroid of the focal triangle $(U/2, \sqrt{3} / 6 U)$ (Figure \ref{fig:fig1}c).
 Colonization occurs with probability 1 if the pathogen lands in the reference stomate, so we need to integrate the probability of colonization if it lands elsewhere.
@@
 
-The [Model] calculates a probability of host colonization (Equation \ref{eq:p_colonize}) as a function stomatal density, size, and position of a pathogen on the leaf.
-I solved $p_\text{colonize}$ using the `integral2()` function in the **pracma** package version 2.2.5 [@borchers_pracma:_2019] for numerical integration.
-I used R version 3.6.1 [@r_core_team_r:_2019] for all analyses and wrote the paper in **rmarkdown** version 1.17 [@xie_r_2018; @allaire_rmarkdown:_2019].
-Source code is deposited on GitHub (https://github.com/cdmuir/stomata-tradeoff) and will be arcived on Zenodo upon publication.
+```{r}
+library(magrittr)
+pkg_info <- sessioninfo::package_info(pkgs = c("pracma", "reticulate", "rmarkdown")) %>%
+  dplyr::filter(package %in% c("pracma", "reticulate", "rmarkdown")) %$%
+  setNames(ondiskversion, package)
+
+```
+
+The [Model] calculates a probability of host colonization (Equation \ref{eq:p_colonize}) as a function of stomatal density, size, and position of a pathogen on the leaf.
+I solved $p_\text{colonize}$ by importing symbolic derivations from Sympy into R with **reticulate** version `r pkg_info["reticulate"]` [@ushey_reticulate_2020] and used the `integral2()` function in the **pracma** package version `r pkg_info["pracma"]` [@borchers_pracma:_2019] for numerical integration.
+I used R version `r getRversion()` [@r_core_team_r:_2020] for all analyses and wrote the paper in **rmarkdown** version `r pkg_info["rmarkdown"]` [@xie_r_2018; @allaire_rmarkdown:_2020].
+Citations for additional R software packages are in Appendix 2.
+Source code is deposited on GitHub (https://github.com/cdmuir/stomata-tradeoff) and will be archived on Zenodo upon publication.
 
@@
 I calculated $p_\text{colonize}$ over a biologically plausible grid of stomatal size and density for hypostomatous species based on @de_boer_optimal_2016.
-Stomatal density ranges from $10^1 - 10^{3.5}~\text{mm}^{-2}$; stomatal size ranges from $10^1 - 10^{3.5}~\mu\text{m}^2$.
-I only considered combinations of size and density where \fs{} was less than 1/3.
-For simplicity, I have not extended the current analysis to amphistomatous leaves.
+Stomatal density ($D$) ranges from $10^1 - 10^{3.5}~\text{mm}^{-2}$; stomatal size ($S$) ranges from $10^1 - 10^{3.5}~\mu\text{m}^2$.
+I only considered combinations of size and density where stomatal cover (\fs) was less than 1/3, which is close to the upper limit in terrestrial plants [@de_boer_optimal_2016].
 I crossed stomatal traits with three levels of $H \in \{0, 0.01, 0.1\}$.
-When $H = 0$, a pathogen perists indefinitely on the leaf surface.
+When $H = 0$, a pathogen persists indefinitely on the leaf surface.
 $H = 0.01$ and $H = 0.1$ correspond to low and high death rates, respectively.
@@
 
-The stomatal size-density scaling relationship can be explained in terms of preserving a constant \gsmax{} that is proportional to $D S ^ {0.5}$ when $bm$ is constant (Equation \ref{eq:gsmax}).
+The stomatal size-density scaling relationship can be explained in terms of preserving a constant stomatal conductance (\gsmax) that is proportional to $D S ^ {0.5}$ when $bm$ is constant (Equation \ref{eq:gsmax}).
 In other words, there are infinitely many combinations of $D$ and $S ^ {0.5}$ with the same \gsmax.
@@
 $$ d = c_g - 0.5 s $$
-where lowercase variables are log-transformed equivalents of their uppercase counterparts. The scaling exponent $\beta_g = 0.5$ preseves $C_g$.
+where lowercase variables are log-transformed equivalents of their uppercase counterparts (Table 1). The scaling exponent $\beta_g = 0.5$ preserves $C_g$.
 
-Next, imagine there is similarly a scaling exponent $\beta_p$ that preserves $p_\text{colonize}$ for the product $D S^{\beta_p}$.
+Next, suppose there is a scaling exponent $\beta_p$ that preserves $p_\text{colonize}$ for the product $D S^{\beta_p}$.
 If $\beta_p = 0.5$, then $p_\text{colonize}$ is always proportional to \gsmax.
-If $\beta_p > 0.5$, small, densely packed stomata would be better defended (lower $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax.
+If $\beta_p > 0.5$, small, densely packed stomata would be more resistant to colonization (lower $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax.
 If $\beta_p < 0.5$, small, densely packed stomata would be less defended (higher $p_\text{colonize}$) compared to larger, sparsely spaced stomata with the same \gsmax.
@@
 I was unable to solve analytically for $\beta_p$, so I numerically calculated isoclines of $p_\text{colonize}$ over the grid of $D$ and $S$ values described in the preceding subsection.
-I numerically calculated the scaling relationships at a constant $p_\text{colonize} \in \{0.025, 0.05, 0.1, 0.2, 0.4\}$ for $H \in \{0, 0.01, 0.1\}$.
+I numerically calculated the scaling relationships at a constant $p_\text{colonize} \in \{0.025, 0.1, 0.4\}$ for $H \in \{0, 0.01, 0.1\}$.
 
@@
 
-I analyzed an idealized, spatially explicit [Model] of how a pathogen lands on a leaf and finds a stomate to colonize the leaf using a random search.
-To my knowledge, this is the first model that makes quantitative predictions about the relationship between stomatal anatomy, the probability of colonization, and their impact on stomatal size-density scaling.
-
 ## Nonlinear relationships between colonization, stomatal cover, and conductance {-}
 
-The probability of colonization ($p_\text{colonize}$) is not simply a one-to-one relationship between the fraction of epidermal area allocated to stomata (\fs).
-At low \fs, $p_\text{colonize}$ increases faster rapdily relative to \fs{} at first Figure \ref{fig:fig3}a).
-At higher \fs, the $p_\text{colonize}$ increases linearly with \fs.
-When $H = 0$, any combination of stomatal size ($S$) and density ($D$) with the same \fs{} have the same effect on $p_\text{colonize}$.
+The probability of colonization ($p_\text{colonize}$) is not simply proportional to stomatal cover (\fs).
+At low \fs, $p_\text{colonize}$ increases rapidly relative to \fs{} at first Figure \ref{fig:fig3}a).
+At higher \fs, $p_\text{colonize}$ increases linearly with \fs.
+When pathogens persist indefinitely ($H = 0$), any combination of stomatal size ($S$) and density ($D$) with the same \fs{} have the same effect on $p_\text{colonize}$.
 When $H > 0$, pathogens are less likely to land close enough to a stomate to infect before dying, so $p_\text{colonize}$ is closer to \fs{} (Figure \ref{fig:fig3}a).
-Furthermore, $p_\text{colonize}$ depends on $D$ and $S$, not just \fs.
+The maximum $p_\text{colonize}$ under the range of parameters considered was $\sim 0.6$ when $H = 0$ and \fs{} is at its maximum value of 1/3.
+When \fs{} is low, $p_\text{colonize}$ is also low.
+The relationship between $p_\text{colonize}$, \fs, and \gsmax is qualitatively similar in the spatially implicit model, but the values for $p_\text{colonize}$ are substantially higher because pathogens can potentially colonize any stomate on the leaf rather than only those in the focal triangle (see [Appendix 1: Spatially implicit model] for more detail).
+Bear in mind that this is the probability for a single individual searching randomly; if enough individuals reach the leaf and/or they can actively find stomata, it’s almost certain that at least some will colonize the leaf.
+However, reducing $p_\text{colonize}$ may help plants limit the damage since fewer total individual pathogens will colonize the leaf interior.
+
+$p_\text{colonize}$ is not directly proportional to \fs{} because it depends on $D$ and $S$ in quantitatively different ways (Figure \ref{fig:figS1}).
 For the same \fs, leaves with greater $D$ have higher $p_\text{colonize}$ (Figure \ref{fig:fig3}a).
@@
 When $H > 0$, this extra distance leads more pathogens to die before they can find a stomate.
-In contrast to \fs, $p_\text{colonize}$ increases at a greater than linear rate with \gsmax.
-Greater $D$ (smaller $S$) is associated with lower $p_\text{colonize}$ when \gsmax{} is held constant (Figure \ref{fig:fig3}b).
+However, this result is inconsistent with the spatially implicit model (Appendix 1) because $S$ and $D$ have identical effects on \fs.
+
+In contrast to \fs, $p_\text{colonize}$ increases at a greater than linear rate with stomatal conductance (\gsmax).
+Greater $D$ (smaller $S$) is associated with lower $p_\text{colonize}$ for a given value of \gsmax{} (Figure \ref{fig:fig3}b).
 This happens because $p_\text{colonize}$ increases approximately linearly with $S$ whereas \gsmax{} is proportional to $S^{0.5}$.
+Therefore, $p_\text{colonize}$ increases exponentially with \gsmax{} at all stomatal densities, but the rate of growth is lower at greater $D$ for a given value of \gsmax.
 
@@
 	\includegraphics{../figures/fig3.pdf}
-	\caption{\textbf{The probability of colonization increases with both stomatal cover and conductance} \textbf{a.} The probability of colonization ($p_\text{colonize}$, $y$-axis) initially increases rapidly with stomatal cover (\fs), then slows down to a linear relationship.
+	\caption{\textbf{The probability of colonization increases with both stomatal cover and conductance} I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of stomatal densities and sizes (see [Materials and Methods]), but a subset of results are shown here.
+Stomatal size and density determine stomatal cover (\fs; Equation \ref{eq:fs}) and theoretical maximum stomatal conductance (\gsmax; Equation \ref{eq:gsmax}).
+\textbf{a.} $p_\text{colonize}$ initially increases rapidly with \fs{} ($x$-axis), then slows down to a linear relationship.
 Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$).
-The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density when $H = 0$ (upper facet).
+The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density when $H = 0$ (upper facet), which is why the lines overlap.
 When $H > 0$, higher density (solid lines) increase $p_\text{colonize}$ (lower facets).
-\textbf{b.} $p_\text{colonize}$ increases expoentially with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is much lower at higher densities for a given \gsmax.
+\textbf{b.} $p_\text{colonize}$ increases exponentially with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is much lower at higher densities for a given \gsmax.
 The relationship between \gsmax{} and $p_\text{colonize}$ is similar for all values of $H$.}
@@
 
-The scaling relationship between $S$ and $D$ that preserves $p_\text{colonize}$ is always greater 0.5 (hyper-conductance), but usually less than 1.
+The scaling relationship between $S$ and $D$ that preserves $p_\text{colonize}$ is always greater than 0.5 (hyper-conductance), but usually less than 1.
 When $H = 0$, the scaling relationship is essentially 1 (Figure \ref{fig:fig4}), which means that an increase \fs{} leads to a proportional increase in $p_\text{colonize}$.
@@
 In this area of parameter space, lower stomatal density can reduce \fs{} while $p_\text{colonize}$ is constant, but this will still result in lower \gsmax.
+In the spatially implicit model, the size-density scaling exponent was always exactly 1 except when $H = 0$ (Appendix 1).
 
@@
 	\includegraphics{../figures/fig4.pdf}
-	\caption{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$).
-In each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.025 (lowest line), 0.05, 0.1, 0.2, or 0.4 (highest line).
+	\caption{\textbf{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$).} In each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.025 (lowest line), 0.1, or 0.4 (highest line).
 For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.5$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region.
@@
 Stomatal density and size set the upper limit on gas exchange in leaves [@harrison_influence_2019] and is often closely related to operational stomatal conductance in nature [@murray_consistent_2019].
-Despite the fact that many ecologically and economically significant plant pathogens infect through stomata, the relationship between stomatal anatomy and susceptibility to foliar pathogens is less clear than it is for gas exchange.
-To develop testable predictions, we need mathematical models that can clarify the potential for tradeoffs between stomatal conductance, stomatal cover, and disease resistance.
+Despite the fact that many foliar pathogens infect through stomata, the relationship between stomatal anatomy and resistance to foliar pathogens is less clear than it is for gas exchange.
 I used a spatially explicit model of a pathogen searching for a stomate to colonize a host.
-From this [Model], I derived predictions about the relationship between stomatal anatomy and disease resistance for the first time.
+From this [Model], I derived predictions about the relationship between stomatal anatomy and the probability of colonization, a component of disease resistance.
 The model predicts that the probability of colonization is not always proportional to the surface area of leaf covered by stomata (\fs), as one might intuitively predict.
@@
 However, \gsmax{} decreases proportionally more than the probability of colonization.
-The model reveals the potential for conflicting demands of maximizing disease resistance, minimizing stomatal cover, and maintaining stomatal conductance.
-Including the effect of anatomy on disease resistance therefore has the potential to change our understading of how stomatal size-density scaling evolves in land plants.
+The model highlights the potential for conflicting demands of minimizing pathogen colonization, minimizing stomatal cover, and maintaining stomatal conductance.
+Including the effect of anatomy on pathogen colonization therefore has the potential to change our understanding of how stomatal size-density scaling evolves in land plants.
 
 <!-- paragraph on first main result -->
-The model predicts that in most cases, increasing stomatal cover should lead to a proportonal increase in susceptibility, which is the implicit assumption of some empirical studies (e.g.
+The model predicts that in most cases, increasing stomatal cover should lead to a proportional increase in colonization, which agrees with empirical studies (e.g.
 @mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019).
-It also makes new, testable predictions that are less intuitive.
-At very low \fs, there is a rapid increase in susceptibility (Figure \ref{fig:fig3}a).
+It also makes new, testable predictions that are less intuitive (Table 2).
+At very low \fs, there is a rapid increase in colonization (Figure \ref{fig:fig3}a).
 If there are no stomata, the probability of colonization is 0, so the first few stomata dramatically increase the probability.
-This is unlikely to be significant for abaxial (lower) leaf surfaces, which usually have most of the stomata [@salisbury_causes_1928; @metcalfe_anatomy_1950; @mott_adaptive_1984; @peat_comparative_1994; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @drake_two_2019].
+This is less likely to be significant for abaxial (lower) leaf surfaces, which usually have most of the stomata [@salisbury_causes_1928; @metcalfe_anatomy_1950; @mott_adaptive_1984; @peat_comparative_1994; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @drake_two_2019].
 However, many adaxial (upper) leaf surfaces have zero or very few stomata.
-Using adaxial leaf surfaces, it should be possible to test if small changes in stomatal size or density have a larger effect on disease susceptibility when \fs{} is low.
+Using adaxial leaf surfaces, it should be possible to test if small changes in stomatal size or density have a larger effect on pathogen colonization when \fs{} is low.
+Such experiments could use natural genetic variation [@mckown_association_2014] or mutant lines [@dow_physiological_2014].
 The nonlinear increase in $p_\text{colonize}$ is less apparent when $H > 0$ (Figure \ref{fig:fig3}a).
 A more hostile microenvironment (e.g.
-drier, higher UV) should therefore reduce the effect of increased size or density as low \fs.
-If true, the dimishing marginal effect of \fs{} on colonization could explain why stomatal ratio on the upper and lower surface is bimodal [@muir_making_2015].
-The initial cost of adaxial (upper) stomata is high, but if the benefits outweigh the costs, then equal stomatal densities on each surface maximize CO$_2$ supply for photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990].
+drier, higher UV) should therefore reduce the effect of increased size or density at low \fs.
+If true, the diminishing marginal effect of \fs{} on colonization could explain why stomatal ratio on the upper and lower surface is bimodal [@muir_making_2015].
+The initial cost of adaxial (upper) stomata is relatively high, but if the benefits outweigh the costs, then equal stomatal densities on each surface maximize CO$_2$ supply for photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990].
+The costs and benefits will certainly vary with environmental conditions as well.
+Future work should extend this model, which considered hypostomatous leaves, to address stomatal size and density in amphistomatous leaves, since leaf surfaces may differ in the type of pathogens present and microenvironment [@mckown_association_2014; @fetter_trade-offs_2019].
+
+**Table 2** | New testable model predictions and suggested experiments to test them.
+
+ Model Prediction | How to test it
+------------------|----------------
+Increasing stomatal size and/or density will have a larger effect on pathogen colonization in leaves with low stomatal cover.
+| Compare the effect of changing stomatal size and/or density on pathogen colonization in leaves with low and high stomatal cover.
+Increasing stomatal size and/or density will have a smaller effect on pathogen colonization in ephiphytic environments more hostile to pathogens.
+| Compare the effect of changing stomatal size and/or density on pathogen colonization in more hostile environments (e.g.
+drier, higher light)
+When selection against pathogen colonization stronger, the stomatal size-density scaling exponent should be lower | Measure stomatal anatomy in environments that differ in pathogen colonization using comparative or experimental approaches
 
 <!-- paragraph on scaling -->
-An effect of stomatal size and density on susceptibility to foliar pathogens could change our understanding of stomatal size-density scaling.
+An effect of stomatal size and density on foliar pathogen colonization could change our understanding of stomatal size-density scaling.
 Since allocating leaf epidermis to stomata may be costly [@franks_mechanical_2007; @assmann_guard_1987; @dow_physiological_2014; @lehmann_effects_2015; @baresch_competition_2019], selection should favor leaves that achieve a desired \gsmax{} while minimizing \fs{} [@de_boer_optimal_2016].
@@
 However, many leaves have larger, sparsely packed stomata.
-Incorporating susceptibility to disease may explain why.
+Incorporating pathogen colonization may explain why.
 If pathogens have a limited time to find stomata before dying ($H > 0$), then the scaling exponent between size and density that keeps $p_\text{colonize}$ constant is between 0.5 and 1, the scaling exponents for \gsmax{} and \fs, respectively (Figure \ref{fig:fig4}).
@@
 The calculated probabilities of colonization should not be interpreted as exact predictions, but rather as depicting qualitative relationships between stomatal anatomy and infection severity.
-The model is most applicable to diseases where the host has some resistance.
 The energetic cost and lost photosynthetic capacity (closed stomata, necrosis, etc.) of dealing with a pathogen is assumed to be proportional to the amount of infection.
-The actual fitness cost will be modulated by the number of pathogens landing on the leaf and the cost of infection.
+The actual fitness cost will be modulated by the number of pathogens landing on the leaf and the cost of infection, all else being equal.
 In environments with fewer or less virulent pathogens, the fitness cost of infection will be less than in environments with more abundant, virulent pathogens.
 The model is less relevant to very susceptible host plants that can be severely damaged or killed by a small number of colonizations that spread unchecked throughout the host tissue.
-
-<!-- other limitations -->
-The purpose of this model is to provide a general foundation to examine the relationship between stomatal size, density, and defense against foliar pathogens.
-In its generality, it overlooks interesting natural history and biologically important features of specific plants and their pathogens.
-For example, some pathogens actively seek out and find stomata [@kiefer_host_2002], whereas I assumed that pathogens randomly orient themselves on the leaf.
-Including sensing should increase the probability of colonization.
-I also assumed that the plant does not respond by, for example, closing stomata when it senses a pathogen.
-However, plants can sense and close stomata when pathogens land on the leaf, but pathogens can pry stomata back open [@melotto_plant_2006].
-Future work on specific plant-pathogen interactions could build on this model by adding more biological realism to provide more precise predictions.
 
@@
 Second, maximizing disease resistance sets up a potential conflict between minimizing stomatal cover and maximizing stomatal conductance.
-The first prediction may be relatively straightforward to test experimentally with adaxial (upper) stomata that occur at low and moderate densities within the same or closely related species [@muir_morphological_2014; @mckown_association_2014; @fetter_trade-offs_2019].
+The first prediction is consistent with results in *Populus trichocarpa* [@mckown_association_2014] and may be relatively straightforward to test experimentally with adaxial (upper) stomata that occur at low and moderate densities within the same or closely related species [@muir_morphological_2014; @fetter_trade-offs_2019].
 The second prediction about size-density scaling is more complex because we would need to know the relationships between colonization, stomatal cover, stomatal conductance, and fitness in natural conditions.
-Testing these predictions in a variety of species would help determine whether pathogens have played an important role shaping  stomatal anatomy in land plants.
+There is growing evidence that stomata mediate tradeoffs between photosynthesis and defense in *Populus trichocarpa* [@mckown_role_2019], but testing these predictions in a variety of species will help determine whether pathogens have played an important role shaping stomatal anatomy in land plants.
 
-<!-- # Acknowledgements {-} -->
+# Acknowledgments {-}
+
+I would like to thank Athena McKown and an anonymous referee for valuable feedback that improved this manuscript.
 
@@
 
+\clearpage
+
 # Supplementary Material {-}
+
+\renewcommand{\theequation}{S\arabic{equation}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
+\setcounter{equation}{0}    
+\setcounter{figure}{0}    
+\setcounter{table}{0}
+
+## $g_\text{s,max}$ calculation {-}
 
@@
 
+## \fs{} is proportional to the stomatal pore area index {-}
+
+The stomatal pore area index [SPI; @sack_hydrology_2003] is calculated as the product of the stomatal density and guard cell length (GCL) squared:
+
+\begin{equation*}
+  \mathrm{SPI} = D \times \mathrm{GCL} ^ 2
+\end{equation*}
+
+Assuming that the stomatal radius $R$ is half the GCL, then stomatal size $S$ is equivalent to:
+
+\begin{equation*}
+  S = \pi R ^ 2 = \frac{\pi \times \mathrm{GCL} ^ 2}{4}
+\end{equation*}
+
+Based on equation \ref{eq:fs}, it follows that \fs{} and SPI are proportional:
+
+\begin{equation*}
+  f_\text{S} = \frac{\pi \times \mathrm{SPI}}{4}
+\end{equation*}
+
+\clearpage
+
+## Figures {-}
+
+\begin{figure}
+  \centering
+	\includegraphics{../figures/figS1.pdf}
+	\caption{\textbf{The probability of colonization increases with both stomatal size ($S$) and density $D$} I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of $S$, $D$, and $H$ (see [Materials and Methods]) \textbf{a.} Each line shows how $p_\text{colonize}$ increases with $S$ ($x$-axis, log-scale) for selected values of $D \in \{10, 100, 1000\}~\text{mm}^{-2}$.
+\textbf{b.} Each line shows how $p_\text{colonize}$ increases with $D$ ($x$-axis, log-scale) for selected values of $S \in \{10, 100, 1000\}~\mu\text{m}^2$.
+The facets show results for different values of $H$}
+	\label{fig:figS1}
+\end{figure}
+
+\begin{figure}
+  \centering
+	\includegraphics{../figures/figS2.pdf}
+	\caption{\textbf{The probability of colonization increases with both stomatal cover and conductance in the spatially implicit model.} As in Figure \ref{fig:fig3}, I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of stomatal densities and sizes (see [Materials and Methods]), but a subset of results are shown here.
+Stomatal size and density determine stomatal cover (\fs; Equation \ref{eq:fs}) and theoretical maximum stomatal conductance (\gsmax; Equation \ref{eq:gsmax}).
+\textbf{a.} $p_\text{colonize}$ initially increases rapidly with \fs{} ($x$-axis), then slows down to a linear relationship.
+Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$).
+The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density for all values of $H$, which is why the lines overlap.
+\textbf{b.} $p_\text{colonize}$ increases sigmoidally with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is lower at higher densities for a given \gsmax.
+The relationship between \gsmax{} and $p_\text{colonize}$ is similar for all values of $H > 0$.}
+	\label{fig:figS2}
+\end{figure}
+
+\begin{figure}
+  \centering
+	\includegraphics{../figures/figS3.pdf}
+	\caption{\textbf{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$) in the spatially implicit model}.
+As in Figure \ref{fig:fig4}, in each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.25 (lowest line), 0.5, or 0.75 (highest line).
+For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.5$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region.
+The scaling exponent is unity $\beta = 1$ when $H > 0$.}
+	\label{fig:figS3}
+\end{figure}
+
+\clearpage
+
+# Appendix 1: Spatially implicit model {-}
+
+A limitation of the spatially explicit model is that a pathogen could only infect stomata in the focal triangle where it landed.
+Here I analyze an alternative, spatially implicit, model that relaxes this assumptions.
+Instead, I assume that a pathogen can potentially infect any stomate on the leaf.
+It searches through a random walk and has a continuous, constant probability of encountering a stomate that is determined by stomatal cover (\fs).
+If $fs << 1$, this can be modeled as a homogeneous Poisson process and the distance $x$ a pathogen must travel before reaching a stomate follows an exponential distribution:
+
+\begin{equation*}
+  f(x) = f_\text{S} e ^ {-f_\text{S} x}
+\end{equation*}
+
+Given the constant death rate per unit distance $H$, the probability of surviving to distance $x$ is $e^{-H x}$.
+The probability of locating a stomate the probability of surviving to distance $x$ multiplied by $f(x)$ and integrated over all $x$ from 0 to $\infty$:
+
+\begin{align*}
+  p_\text{locate} & = \int_0^\infty f(x) e^{-Hx} dx \\
+                  & = \int_0^\infty f_\text{S} e^{-(f_\text{S} + H)x} dx \\
+                  & = \frac{f_\text{S}}{H + f_\text{S}}
+\end{align*}
+
+Substituting $p_\text{locate}$ above into Equation \ref{eq:p_colonize}:
+
+\begin{align*}
+  p_\text{colonize} & = f_\text{S} + (1 - f_\text{S}) p_\text{locate} \\
+                    & = f_\text{S} + (1 - f_\text{S}) \frac{f_\text{S}}{H + f_\text{S}} \\
+                    & = f_\text{S} (1 + \frac{1 - f_\text{S}}{H + f_\text{S}})
+\end{align*}
+
+With the spatially implicit model, because pathogens can potentially reach any stomate on the leaf, $p_\text{colonize}$ is greater than that in the spatially explicit model for the same value of $H$.
+For example, if the pathogen can search forever ($H = 0$), then it will always colonize ($p_\text{colonize} = 1$; Figure \ref{fig:figS2}a).
+But even when $H > 0$, $p_\text{colonize}$ is significantly higher than in the spatially implicit than spatially explicit model for the same \fs{} because pathogens can potentially colonize any stomate on the leaf.
+
+Whereas the spatially explicit model probably underestimates $p_\text{locate}$ for pathogens that can search over long distances, the implicit model overestimates because it assumes that the probability of encountering a stomate is constant (i.e.
+homogeneous Poisson process).
+This is not true because stomata are discrete areas on the leaf.
+If a pathogen is searching far away from a stomate, its probability of encountering a stomate in the near future is lower than that for a pathogen searching near a stomate.
+This should be modeled as a *nonhomogenous* Poisson process.
+Future work should derive $p_\text{locate}$ for the stomatal anatomies presented here under a nonhomogeneous process.
+
+Despite the quantitative differences in the the spatially explicit and implicit models, they have similar qualitative properties when $H > 0$, which is reasonable since the leaf surface is a relatively hostile environment for most pathogens (see [Introduction]).
+In both models, $p_\text{colonize}$ increases with, but is higher than \fs.
+In the spatially explicit model, size-density scaling that preserves $p_\text{colonize}$ is 1 when $H = 0$ and slightly less than 1 otherwise (Figure \ref{fig:fig4}).
+In the spatially implicit model, the scaling coefficient is always 1.
+Rearranging the equation for $p_\text{colonize}$ above and substituting $f_\text{S} = DS$, the following relationship holds:
+
+$$SD = \frac{p_\text{colonize} H}{1 - p_\text{colonize} + H}$$
+Since $H$ is a constant the right-hand side of the equation above is constant for a given value of $p_\text{colonize}$.
+Hence the $\beta_p$ that would preserve the relationship above is simply 1 (Figure \ref{fig:figS3}).
+
+\clearpage
+
 # References {-}

### 2020-07-13 r/03_plot-spatially-explicit.R

@@ -2,20 +2,20 @@ source("r/header.R")
 
 df <- read_rds("objects/spatially-explicit-results.rds")
 
-# Effect of f_s and gmax om p_colonize ----
+# Figure 3: Effect of f_s and gmax on p_colonize ----
 
 df1 <- df %>%
-  filter(D %in% c(10, 100, 1000)) %>%
-  mutate(H = glue("H = {H}", H = H))
+  filter(D %in% c(100, 500)) %>%
+  mutate(H = glue("H = {H}", H = H), D = factor(round(D), levels = c("100", "500")))
 
-f3a <- ggplot(df1, aes(x = f_s, y = p_colonize, linetype = as.character(D))) +
+f3a <- ggplot(df1, aes(x = f_s, y = p_colonize, linetype = D)) +
   facet_grid(H ~ .) +
   geom_line(size = 1.1, lineend = "round") +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
   xlab(expression(paste("Stomatal cover (", italic(f)[s], ")"))) +
   ylab(expression(paste("Probability of colonization (", italic(p)[colonize], ")"))) +
   scale_linetype_manual(
-    values = 3:1, 
+    values = c(2, 1), 
     name = expression(paste("Stomatal density [m", m^-2, "]"))
   ) +
   theme_bw() +
@@ -23,6 +23,7 @@ f3a <- ggplot(df1, aes(x = f_s, y = p_colonize, linetype = as.character(D))) +
     axis.text = element_text(size = 10),
     axis.title = element_text(size = 12),
     axis.title.x = element_text(margin = margin(0.3, 0, 0, 0, unit = "cm")),
+    legend.key.size = unit(0.75, "in"),
     legend.position = "top",
     strip.background = element_rect(fill = "white"),
     strip.text = element_text(size = 10)
@@ -35,12 +36,12 @@ f3a <- f3a + theme(legend.position = "none")
 df2 <- df1 %>%
   filter(g_smax < 5)
 
-f3b <- ggplot(df2, aes(x = g_smax, y = p_colonize, linetype = as.character(D))) +
+f3b <- ggplot(df2, aes(x = g_smax, y = p_colonize, linetype = D)) +
   facet_wrap(H ~ ., strip.position = "left", ncol = 1) +
   geom_line(size = 1.1, lineend = "round") +
[…]

### 2020-07-13 r/04_find-scaling.R

@@ -1,5 +1,6 @@
 source("r/header.R")
 
+# Spatially explicit model ----
 df <- crossing(
   p_colonize = c(0.025, 0.05, 0.1, 0.2, 0.4),
   D = set_units(10 ^ seq(1, 3.5, by = 0.025), 1 / mm ^ 2),
@@ -8,4 +9,16 @@ df <- crossing(
   rowwise() %>%
   mutate(S = find_S(p_colonize, D, H))
 
-write_rds(df, "objects/scaling-results.rds")
+write_rds(df, "objects/spatially-explicit-scaling-results.rds")
+
+# Spatially implicit model ----
+df <- crossing(
+  p_colonize = c(0.25, 0.5, 0.75),
+  D = set_units(10 ^ seq(1, 3.5, by = 0.025), 1 / mm ^ 2),
+  H = c(0.01, 0.1)
+) %>%
+  rowwise() %>%
+  # Multiplying by 10^6 to convert mm^2 to um^2
+  mutate(S = 1e6 * p_colonize * H / (D * (1 - p_colonize + H)))
+
+write_rds(df, "objects/spatially-implicit-scaling-results.rds")

### 2020-07-13 r/05_plot-scaling.R

@@ -1,7 +1,9 @@
 source("r/header.R")
 
-df <- read_rds("objects/scaling-results.rds") %>%
+# Spatially explicit model ----
+df <- read_rds("objects/spatially-explicit-scaling-results.rds") %>%
   ungroup() %>%
+  filter(p_colonize %in% c(0.025, 0.1, 0.4)) %>%
   mutate(x = D * S) %>%
   mutate_if(~ {inherits(.x, "units")}, drop_units) %>%
   mutate(H = glue("H = {H}", H = H))
@@ -11,13 +13,17 @@ df_refline <- df %>%
   group_by(H) %>%
   summarize(min_D = min(D), max_D = max(D), D = median(D), S = median(S)) %>%
   mutate(d = log(D), s = log(S)) %>%
-  crossing(slope = c(-1, -2)) %>%
+  crossing(nesting(
+    slope = c(-1, -2),
+    `Scaling relationships that preserve:` = c("italic(f)[S]", "italic(g)[s,max]")
+  )) %>%
   mutate(
     intercept = s - slope * d,
-         min_S = exp(intercept + slope * log(min_D)),
-         max_S = exp(intercept + slope * log(max_D))
-    ) %>%
-  select(H, D = min_D, xend = max_D, S = min_S, yend = max_S) %>%
+    min_S = exp(intercept + slope * log(min_D)),
+    max_S = exp(intercept + slope * log(max_D))
+  ) %>%
+  select(H, D = min_D, xend = max_D, S = min_S, yend = max_S, 
+         `Scaling relationships that preserve:`) %>%
   mutate(p_colonize = NA)
 
 df_labels <- df %>%
@@ -29,8 +35,8 @@ fig4 <- ggplot(df, aes(D, S, group = p_colonize)) +
   facet_wrap(H ~ .) +
   geom_segment(
     data = df_refline, 
-    mapping = aes(xend = xend, yend = yend),
-    linetype = "dashed", color = "grey", size = 1.1
+    mapping = aes(xend = xend, yend = yend, linetype = `Scaling relationships that preserve:`),
+    lineend = "round", color = "darkgrey", size = 2
   ) +
   geom_line() +
   geom_text(
@@ -40,12 +46,16 @@ fig4 <- ggplot(df, aes(D, S, group = p_colonize)) +
   ) +
   scale_x_log10(breaks = 10 ^ (1:3), limits = c(1, 3500)) +
   scale_y_log10(breaks = 10 ^ (1:3)) +
[…]

### 2020-07-13 r/functions.R

@@ -35,14 +35,14 @@ U2D <- function(U) {
 S2R <- function(S) {
   
   # S must be in units of um ^ 2
-  sqrt(S / (2 * pi)) # units of um
+  sqrt(S / pi) # units of um
   
 }
 
 R2S <- function(R) {
   
   # r must be in units of um
-  2 * pi * R ^ 2 # units of um ^ 2
+  pi * R ^ 2 # units of um ^ 2
   
 }
 

### 2020-07-13 ms/appendix2.tex
r1

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

\documentclass[utf8]{frontiersSCNS}

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}

\linenumbers

% BELOW TAKEN FROM rticles plos template
%
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% hyperref package, useful for hyperlinks
\usepackage{hyperref}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% Sweave(-like)
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Below is from frontiers
%
\bibliographystyle{frontiersinSCNS}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace}
%\doublespacing

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{soul}
\usepackage{tikz}

% short-cut for fs, gmax, and gop
\newcommand{\fs}{$f_\text{S}$}
\newcommand{\gsmax}{$g_\text{s,max}$}
\newcommand{\gsop}{$g_\text{s,op}$}

\def\Authors{
  Christopher D. Muir\,\textsuperscript{1*}}

\def\Address{

  \textsuperscript{1} School of Life Sciences, University of Hawaii,  Honolulu,  Hawaii,  USA
  }

  \def\corrAuthor{Christopher D. Muir}\def\corrAddress{,  }\def\corrEmail{\href{mailto:cdmuir@hawaii.edu}{\nolinkurl{cdmuir@hawaii.edu}}}
  \def\firstAuthorLast{Muir {et~al.}}

\begin{document}
\onecolumn
\firstpage{1}

\title[]{Appendix 2: Additional R packages}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

\begin{longtable}[]{@{}lll@{}}
\toprule
package & version & reference\tabularnewline
\midrule
\endhead
assertthat & 0.2.1 & Wickham (2019a)\tabularnewline
bibtex & 0.4.2.2 & Francois (2020)\tabularnewline
BiocManager & 1.30.10 & Morgan (2019)\tabularnewline
cli & 2.0.2 & Cs\a'ardi (2020)\tabularnewline
codetools & 0.2-16 & Tierney (2018)\tabularnewline
colorspace & 1.4-1 & Zeileis et al. (2019)\tabularnewline
& & Zeileis et al. (2009)\tabularnewline
& & Stauffer et al. (2009)\tabularnewline
cowplot & 1.0.0 & Wilke (2019)\tabularnewline
crayon & 1.3.4 & Cs\a'ardi (2017)\tabularnewline
digest & 0.6.25 & Antoine Lucas et al. (2020)\tabularnewline
dplyr & 1.0.0 & Wickham et al. (2020a)\tabularnewline
ellipsis & 0.3.1 & Wickham (2020)\tabularnewline
evaluate & 0.14 & Wickham and Xie (2019)\tabularnewline
fansi & 0.4.1 & Gaslam (2020)\tabularnewline
farver & 2.0.3 & Pedersen et al. (2020)\tabularnewline
filehash & 2.4-2 & Peng (2006)\tabularnewline
furrr & 0.1.0 & Vaughan and Dancho (2018)\tabularnewline
future & 1.18.0 & Bengtsson (2020)\tabularnewline
generics & 0.0.2 & Kuhn et al. (2018)\tabularnewline
ggforce & 0.3.2 & Pedersen (2020)\tabularnewline
ggimage & 0.2.8 & Yu (2020a)\tabularnewline
gginnards & 0.0.3 & Aphalo (2019)\tabularnewline
ggplot2 & 3.3.2 & Wickham (2016)\tabularnewline
ggplotify & 0.0.5 & Yu (2020b)\tabularnewline
globals & 0.12.5 & Bengtsson (2019a)\tabularnewline
glue & 1.4.1 & Hester (2020)\tabularnewline
gridGraphics & 0.5-0 & Murrell and Wen (2020)\tabularnewline
gtable & 0.3.0 & Wickham and Pedersen (2019)\tabularnewline
hms & 0.5.3 & Müller (2020)\tabularnewline
htmltools & 0.5.0 & Cheng et al. (2020)\tabularnewline
httr & 1.4.1 & Wickham (2019b)\tabularnewline
jsonlite & 1.7.0 & Ooms (2014)\tabularnewline
knitcitations & 1.0.10 & Boettiger (2019)\tabularnewline
knitr & 1.29 & Xie (2020a)\tabularnewline
& & Xie (2015)\tabularnewline
& & Xie (2014)\tabularnewline
lifecycle & 0.2.0 & Henry (2020)\tabularnewline
listenv & 0.8.0 & Bengtsson (2019b)\tabularnewline
lubridate & 1.7.9 & Grolemund and Wickham (2011)\tabularnewline
magick & 2.4.0 & Ooms (2020)\tabularnewline
magrittr & 1.5 & Bache and Wickham (2014)\tabularnewline
MASS & 7.3-51.6 & Venables and Ripley (2002)\tabularnewline
munsell & 0.5.0 & Wickham (2018)\tabularnewline
pillar & 1.4.6 & Müller and Wickham (2020a)\tabularnewline
pkgconfig & 2.0.3 & Cs\a'ardi (2019)\tabularnewline
plyr & 1.8.6 & Wickham (2011)\tabularnewline
polyclip & 1.10-0 & Johnson and Baddeley (2019)\tabularnewline
pracma & 2.2.9 & Borchers (2019)\tabularnewline
purrr & 0.3.4 & Henry and Wickham (2020a)\tabularnewline
R6 & 2.4.1 & Chang (2019)\tabularnewline
Rcpp & 1.0.5 & Eddelbuettel and François (2011)\tabularnewline
& & Eddelbuettel (2013)\tabularnewline
& & Eddelbuettel and Balamuta (2017)\tabularnewline
readr & 1.3.1 & Wickham et al. (2018)\tabularnewline
RefManageR & 1.2.12 & McLean (2017)\tabularnewline
& & McLean (2014)\tabularnewline
rlang & 0.4.7 & Henry and Wickham (2020b)\tabularnewline
rmarkdown & 2.3 & Allaire et al. (2020a)\tabularnewline
& & Xie et al. (2018)\tabularnewline
rticles & 0.14 & Allaire et al. (2020b)\tabularnewline
rvcheck & 0.1.8 & Yu (2020c)\tabularnewline
scales & 1.1.1 & Wickham and Seidel (2020)\tabularnewline
sessioninfo & 1.1.1 & Cs\a'ardi et al. (2018)\tabularnewline
stringi & 1.4.6 & Gagolewski (2020)\tabularnewline
stringr & 1.4.0 & Wickham (2019c)\tabularnewline
tibble & 3.0.3 & Müller and Wickham (2020b)\tabularnewline
tidyr & 1.1.0 & Wickham and Henry (2020)\tabularnewline
tidyselect & 1.1.0 & Henry and Wickham (2020c)\tabularnewline
tikzDevice & 0.12.3.1 & Sharpsteen and Bracken (2020)\tabularnewline
tweenr & 1.0.1 & Pedersen (2018)\tabularnewline
units & 0.6-7 & Pebesma et al. (2016)\tabularnewline
vctrs & 0.3.1 & Wickham et al. (2020b)\tabularnewline
withr & 2.2.0 & Hester et al. (2020)\tabularnewline
xfun & 0.15 & Xie (2020b)\tabularnewline
xml2 & 1.3.2 & Wickham et al. (2020c)\tabularnewline
yaml & 2.2.1 & Stephens et al. (2020)\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-allaire2020rmarkdown}{}%
Allaire, J., Xie, Y., McPherson, J., Luraschi, J., Ushey, K., Atkins,
A., Wickham, H., Cheng, J., Chang, W., and Iannone, R. (2020a).
\emph{Rmarkdown: Dynamic documents for r}. Available at:
\url{https://github.com/rstudio/rmarkdown}.

\leavevmode\hypertarget{ref-allaire2020rticles}{}%
Allaire, J., Xie, Y., R Foundation, Wickham, H., Journal of Statistical
Software, Vaidyanathan, R., Association for Computing Machinery,
Boettiger, C., Elsevier, Broman, K., et al. (2020b). \emph{Rticles:
Article formats for r markdown}. Available at:
\url{https://CRAN.R-project.org/package=rticles}.

\leavevmode\hypertarget{ref-lucas2020digest}{}%
Antoine Lucas, D. E. with contributions by, Tuszynski, J., Bengtsson,
H., Urbanek, S., Frasca, M., Lewis, B., Stokely, M., Muehleisen, H.,
Murdoch, D., Hester, J., et al. (2020). \emph{Digest: Create compact
hash digests of r objects}. Available at:
\url{https://CRAN.R-project.org/package=digest}.

\leavevmode\hypertarget{ref-aphalo2019gginnards}{}%
Aphalo, P. J. (2019). \emph{Gginnards: Explore the innards of 'ggplot2'
objects}. Available at:
\url{https://CRAN.R-project.org/package=gginnards}.

\leavevmode\hypertarget{ref-bache2014magrittr}{}%
Bache, S. M., and Wickham, H. (2014). \emph{Magrittr: A forward-pipe
operator for r}. Available at:
\url{https://CRAN.R-project.org/package=magrittr}.

\leavevmode\hypertarget{ref-bengtsson2020future}{}%
Bengtsson, H. (2020). \emph{Future: Unified parallel and distributed
processing in r for everyone}. Available at:
\url{https://CRAN.R-project.org/package=future}.

\leavevmode\hypertarget{ref-bengtsson2019globals}{}%
Bengtsson, H. (2019a). \emph{Globals: Identify global objects in r
expressions}. Available at:
\url{https://CRAN.R-project.org/package=globals}.

\leavevmode\hypertarget{ref-bengtsson2019listenv}{}%
Bengtsson, H. (2019b). \emph{Listenv: Environments behaving (almost) as
lists}. Available at: \url{https://CRAN.R-project.org/package=listenv}.

\leavevmode\hypertarget{ref-boettiger2019knitcitations}{}%
Boettiger, C. (2019). \emph{Knitcitations: Citations for 'knitr'
markdown files}. Available at:
\url{https://CRAN.R-project.org/package=knitcitations}.

\leavevmode\hypertarget{ref-borchers2019pracma}{}%
Borchers, H. W. (2019). \emph{Pracma: Practical numerical math
functions}. Available at:
\url{https://CRAN.R-project.org/package=pracma}.

\leavevmode\hypertarget{ref-chang2019encapsulated}{}%
Chang, W. (2019). \emph{R6: Encapsulated classes with reference
semantics}. Available at: \url{https://CRAN.R-project.org/package=R6}.

\leavevmode\hypertarget{ref-cheng2020htmltools}{}%
Cheng, J., Sievert, C., Chang, W., Xie, Y., and Allen, J. (2020).
\emph{Htmltools: Tools for html}. Available at:
\url{https://CRAN.R-project.org/package=htmltools}.

\leavevmode\hypertarget{ref-csrdi2020helpers}{}%
Cs\a'ardi, G. (2020). \emph{Cli: Helpers for developing command line
interfaces}. Available at: \url{https://CRAN.R-project.org/package=cli}.

\leavevmode\hypertarget{ref-csrdi2017crayon}{}%
Cs\a'ardi, G. (2017). \emph{Crayon: Colored terminal output}. Available
at: \url{https://CRAN.R-project.org/package=crayon}.

\leavevmode\hypertarget{ref-csrdi2019pkgconfig}{}%
Cs\a'ardi, G. (2019). \emph{Pkgconfig: Private configuration for 'r'
packages}. Available at:
\url{https://CRAN.R-project.org/package=pkgconfig}.

\leavevmode\hypertarget{ref-csrdi2018sessioninfo}{}%
Cs\a'ardi, G., core, R., Wickham, H., Chang, W., Flight, R. M., Müller,
K., and Hester, J. (2018). \emph{Sessioninfo: R session information}.
Available at: \url{https://CRAN.R-project.org/package=sessioninfo}.

\leavevmode\hypertarget{ref-eddelbuettel2013seamless}{}%
Eddelbuettel, D. (2013). \emph{Seamless R and C++ integration with
Rcpp}. New York: Springer
doi:\href{https://doi.org/10.1007/978-1-4614-6868-4}{10.1007/978-1-4614-6868-4}.

\leavevmode\hypertarget{ref-eddelbuettel2017extending}{}%
Eddelbuettel, D., and Balamuta, J. J. (2017). Extending extitR with
extitC++: A Brief Introduction to extitRcpp. \emph{PeerJ Preprints} 5,
e3188v1.
doi:\href{https://doi.org/10.7287/peerj.preprints.3188v1}{10.7287/peerj.preprints.3188v1}.

\leavevmode\hypertarget{ref-eddelbuettel2011rcpp}{}%
Eddelbuettel, D., and François, R. (2011). Rcpp: Seamless R and C++
integration. \emph{Journal of Statistical Software} 40, 1--18.
doi:\href{https://doi.org/10.18637/jss.v040.i08}{10.18637/jss.v040.i08}.

\leavevmode\hypertarget{ref-francois2020bibtex}{}%
Francois, R. (2020). \emph{Bibtex: Bibtex parser}. Available at:
\url{https://CRAN.R-project.org/package=bibtex}.

\leavevmode\hypertarget{ref-gagolewski2020package}{}%
Gagolewski, M. (2020). \emph{R package stringi: Character string
processing facilities}. Available at:
\url{http://www.gagolewski.com/software/stringi/}.

\leavevmode\hypertarget{ref-gaslam2020fansi}{}%
Gaslam, B. (2020). \emph{Fansi: ANSI control sequence aware string
functions}. Available at:
\url{https://CRAN.R-project.org/package=fansi}.

\leavevmode\hypertarget{ref-grolemund2011dates}{}%
Grolemund, G., and Wickham, H. (2011). Dates and times made easy with
lubridate. \emph{Journal of Statistical Software} 40, 1--25. Available
at: \url{http://www.jstatsoft.org/v40/i03/}.

\leavevmode\hypertarget{ref-henry2020lifecycle}{}%
Henry, L. (2020). \emph{Lifecycle: Manage the life cycle of your package
functions}. Available at:
\url{https://CRAN.R-project.org/package=lifecycle}.

\leavevmode\hypertarget{ref-henry2020purrr}{}%
Henry, L., and Wickham, H. (2020a). \emph{Purrr: Functional programming
tools}. Available at: \url{https://CRAN.R-project.org/package=purrr}.

\leavevmode\hypertarget{ref-henry2020rlang}{}%
Henry, L., and Wickham, H. (2020b). \emph{Rlang: Functions for base
types and core r and 'tidyverse' features}. Available at:
\url{https://CRAN.R-project.org/package=rlang}.

\leavevmode\hypertarget{ref-henry2020tidyselect}{}%
Henry, L., and Wickham, H. (2020c). \emph{Tidyselect: Select from a set
of strings}. Available at:
\url{https://CRAN.R-project.org/package=tidyselect}.

\leavevmode\hypertarget{ref-hester2020glue}{}%
Hester, J. (2020). \emph{Glue: Interpreted string literals}. Available
at: \url{https://CRAN.R-project.org/package=glue}.

\leavevmode\hypertarget{ref-hester2020withr}{}%
Hester, J., Müller, K., Ushey, K., Wickham, H., and Chang, W. (2020).
\emph{Withr: Run code 'with' temporarily modified global state}.
Available at: \url{https://CRAN.R-project.org/package=withr}.

\leavevmode\hypertarget{ref-johnson2019polyclip}{}%
Johnson, A., and Baddeley, A. (2019). \emph{Polyclip: Polygon clipping}.
Available at: \url{https://CRAN.R-project.org/package=polyclip}.

\leavevmode\hypertarget{ref-kuhn2018generics}{}%
Kuhn, M., Wickham, H., and Vaughan, D. (2018). \emph{Generics: Common s3
generics not provided by base r methods related to model fitting}.
Available at: \url{https://CRAN.R-project.org/package=generics}.

\leavevmode\hypertarget{ref-mclean2017refmanager}{}%
McLean, M. W. (2017). RefManageR: Import and manage bibtex and biblatex
references in r. \emph{The Journal of Open Source Software}.
doi:\href{https://doi.org/10.21105/joss.00338}{10.21105/joss.00338}.

\leavevmode\hypertarget{ref-mclean2014straightforward}{}%
McLean, M. W. (2014). \emph{Straightforward bibliography management in r
using the refmanager package}. Available at:
\url{https://arxiv.org/abs/1403.2036}.

\leavevmode\hypertarget{ref-morgan2019biocmanager}{}%
Morgan, M. (2019). \emph{BiocManager: Access the bioconductor project
package repository}. Available at:
\url{https://CRAN.R-project.org/package=BiocManager}.

\leavevmode\hypertarget{ref-murrell2020gridgraphics}{}%
Murrell, P., and Wen, Z. (2020). \emph{GridGraphics: Redraw base
graphics using 'grid' graphics}. Available at:
\url{https://CRAN.R-project.org/package=gridGraphics}.

\leavevmode\hypertarget{ref-mller2020pretty}{}%
Müller, K. (2020). \emph{Hms: Pretty time of day}. Available at:
\url{https://CRAN.R-project.org/package=hms}.

\leavevmode\hypertarget{ref-mller2020pillar}{}%
Müller, K., and Wickham, H. (2020a). \emph{Pillar: Coloured formatting
for columns}. Available at:
\url{https://CRAN.R-project.org/package=pillar}.

\leavevmode\hypertarget{ref-mller2020tibble}{}%
Müller, K., and Wickham, H. (2020b). \emph{Tibble: Simple data frames}.
Available at: \url{https://CRAN.R-project.org/package=tibble}.

\leavevmode\hypertarget{ref-ooms2020magick}{}%
Ooms, J. (2020). \emph{Magick: Advanced graphics and image-processing in
r}. Available at: \url{https://CRAN.R-project.org/package=magick}.

\leavevmode\hypertarget{ref-ooms2014jsonlite}{}%
Ooms, J. (2014). The jsonlite package: A practical and consistent
mapping between json data and r objects. \emph{arXiv:1403.2805
{[}stat.CO{]}}. Available at: \url{https://arxiv.org/abs/1403.2805}.

\leavevmode\hypertarget{ref-pebesma2016measurement}{}%
Pebesma, E., Mailund, T., and Hiebert, J. (2016). Measurement units in
R. \emph{R Journal} 8, 486--494.
doi:\href{https://doi.org/10.32614/RJ-2016-061}{10.32614/RJ-2016-061}.

\leavevmode\hypertarget{ref-pedersen2020ggforce}{}%
Pedersen, T. L. (2020). \emph{Ggforce: Accelerating 'ggplot2'}.
Available at: \url{https://CRAN.R-project.org/package=ggforce}.

\leavevmode\hypertarget{ref-pedersen2018tweenr}{}%
Pedersen, T. L. (2018). \emph{Tweenr: Interpolate data for smooth
animations}. Available at:
\url{https://CRAN.R-project.org/package=tweenr}.

\leavevmode\hypertarget{ref-pedersen2020farver}{}%
Pedersen, T. L., Nicolae, B., and François, R. (2020). \emph{Farver:
High performance colour space manipulation}. Available at:
\url{https://CRAN.R-project.org/package=farver}.

\leavevmode\hypertarget{ref-peng2006interacting}{}%
Peng, R. D. (2006). Interacting with data using the filehash package.
\emph{R News} 6, 19--24. Available at:
\url{https://cran.r-project.org/doc/Rnews/}.

\leavevmode\hypertarget{ref-sharpsteen2020tikzdevice}{}%
Sharpsteen, C., and Bracken, C. (2020). \emph{TikzDevice: R graphics
output in latex format}. Available at:
\url{https://CRAN.R-project.org/package=tikzDevice}.

\leavevmode\hypertarget{ref-stauffer2009somewhere}{}%
Stauffer, R., Mayr, G. J., Dabernig, M., and Zeileis, A. (2009).
Somewhere over the rainbow: How to make effective use of colors in
meteorological visualizations. \emph{Bulletin of the American
Meteorological Society} 96, 203--216.
doi:\href{https://doi.org/10.1175/BAMS-D-13-00155.1}{10.1175/BAMS-D-13-00155.1}.

\leavevmode\hypertarget{ref-stephens2020yaml}{}%
Stephens, J., Simonov, K., Xie, Y., Dong, Z., Wickham, H., Horner, J.,
reikoch, Beasley, W., O'Connor, B., and Warnes, G. R. (2020).
\emph{Yaml: Methods to convert r data to yaml and back}. Available at:
\url{https://CRAN.R-project.org/package=yaml}.

\leavevmode\hypertarget{ref-tierney2018codetools}{}%
Tierney, L. (2018). \emph{Codetools: Code analysis tools for r}.
Available at: \url{https://CRAN.R-project.org/package=codetools}.

\leavevmode\hypertarget{ref-vaughan2018furrr}{}%
Vaughan, D., and Dancho, M. (2018). \emph{Furrr: Apply mapping functions
in parallel using futures}. Available at:
\url{https://CRAN.R-project.org/package=furrr}.

\leavevmode\hypertarget{ref-venables2002modern}{}%
Venables, W. N., and Ripley, B. D. (2002). \emph{Modern applied
statistics with s}. Fourth. New York: Springer Available at:
\url{http://www.stats.ox.ac.uk/pub/MASS4}.

\leavevmode\hypertarget{ref-wickham2018munsell}{}%
Wickham, C. (2018). \emph{Munsell: Utilities for using munsell colours}.
Available at: \url{https://CRAN.R-project.org/package=munsell}.

\leavevmode\hypertarget{ref-wickham2019assertthat}{}%
Wickham, H. (2019a). \emph{Assertthat: Easy pre and post assertions}.
Available at: \url{https://CRAN.R-project.org/package=assertthat}.

\leavevmode\hypertarget{ref-wickham2020ellipsis}{}%
Wickham, H. (2020). \emph{Ellipsis: Tools for working with ...}
Available at: \url{https://CRAN.R-project.org/package=ellipsis}.

\leavevmode\hypertarget{ref-wickham2016ggplot2}{}%
Wickham, H. (2016). \emph{Ggplot2: Elegant graphics for data analysis}.
Springer-Verlag New York Available at:
\url{https://ggplot2.tidyverse.org}.

\leavevmode\hypertarget{ref-wickham2019httr}{}%
Wickham, H. (2019b). \emph{Httr: Tools for working with urls and http}.
Available at: \url{https://CRAN.R-project.org/package=httr}.

\leavevmode\hypertarget{ref-wickham2019stringr}{}%
Wickham, H. (2019c). \emph{Stringr: Simple, consistent wrappers for
common string operations}. Available at:
\url{https://CRAN.R-project.org/package=stringr}.

\leavevmode\hypertarget{ref-wickham2011split}{}%
Wickham, H. (2011). The split-apply-combine strategy for data analysis.
\emph{Journal of Statistical Software} 40, 1--29. Available at:
\url{http://www.jstatsoft.org/v40/i01/}.

\leavevmode\hypertarget{ref-wickham2020dplyr}{}%
Wickham, H., François, R., Henry, L., and Müller, K. (2020a).
\emph{Dplyr: A grammar of data manipulation}. Available at:
\url{https://CRAN.R-project.org/package=dplyr}.

\leavevmode\hypertarget{ref-wickham2020tidyr}{}%
Wickham, H., and Henry, L. (2020). \emph{Tidyr: Tidy messy data}.
Available at: \url{https://CRAN.R-project.org/package=tidyr}.

\leavevmode\hypertarget{ref-wickham2020vctrs}{}%
Wickham, H., Henry, L., and Vaughan, D. (2020b). \emph{Vctrs: Vector
helpers}. Available at: \url{https://CRAN.R-project.org/package=vctrs}.

\leavevmode\hypertarget{ref-wickham2018readr}{}%
Wickham, H., Hester, J., and Francois, R. (2018). \emph{Readr: Read
rectangular text data}. Available at:
\url{https://CRAN.R-project.org/package=readr}.

\leavevmode\hypertarget{ref-wickham2020xml2}{}%
Wickham, H., Hester, J., and Ooms, J. (2020c). \emph{Xml2: Parse xml}.
Available at: \url{https://CRAN.R-project.org/package=xml2}.

\leavevmode\hypertarget{ref-wickham2019gtable}{}%
Wickham, H., and Pedersen, T. L. (2019). \emph{Gtable: Arrange 'grobs'
in tables}. Available at:
\url{https://CRAN.R-project.org/package=gtable}.

\leavevmode\hypertarget{ref-wickham2020scales}{}%
Wickham, H., and Seidel, D. (2020). \emph{Scales: Scale functions for
visualization}. Available at:
\url{https://CRAN.R-project.org/package=scales}.

\leavevmode\hypertarget{ref-wickham2019evaluate}{}%
Wickham, H., and Xie, Y. (2019). \emph{Evaluate: Parsing and evaluation
tools that provide more details than the default}. Available at:
\url{https://CRAN.R-project.org/package=evaluate}.

\leavevmode\hypertarget{ref-wilke2019cowplot}{}%
Wilke, C. O. (2019). \emph{Cowplot: Streamlined plot theme and plot
annotations for 'ggplot2'}. Available at:
\url{https://CRAN.R-project.org/package=cowplot}.

\leavevmode\hypertarget{ref-xie2015dynamic}{}%
Xie, Y. (2015). \emph{Dynamic documents with R and knitr}. 2nd ed. Boca
Raton, Florida: Chapman; Hall/CRC Available at:
\url{https://yihui.org/knitr/}.

\leavevmode\hypertarget{ref-xie2014knitr}{}%
Xie, Y. (2014). ``Knitr: A comprehensive tool for reproducible research
in R,'' in \emph{Implementing reproducible computational research}, eds.
V. Stodden, F. Leisch, and R. D. Peng (Chapman; Hall/CRC). Available at:
\url{http://www.crcpress.com/product/isbn/9781466561595}.

\leavevmode\hypertarget{ref-xie2020knitr}{}%
Xie, Y. (2020a). \emph{Knitr: A general-purpose package for dynamic
report generation in r}. Available at: \url{https://yihui.org/knitr/}.

\leavevmode\hypertarget{ref-xie2020xfun}{}%
Xie, Y. (2020b). \emph{Xfun: Miscellaneous functions by 'yihui xie'}.
Available at: \url{https://CRAN.R-project.org/package=xfun}.

\leavevmode\hypertarget{ref-xie2018markdown}{}%
Xie, Y., Allaire, J. J., and Grolemund, G. (2018). \emph{R markdown: The
definitive guide}. Boca Raton, Florida: Chapman; Hall/CRC Available at:
\url{https://bookdown.org/yihui/rmarkdown}.

\leavevmode\hypertarget{ref-yu2020ggimage}{}%
Yu, G. (2020a). \emph{Ggimage: Use image in 'ggplot2'}. Available at:
\url{https://CRAN.R-project.org/package=ggimage}.

\leavevmode\hypertarget{ref-yu2020ggplotify}{}%
Yu, G. (2020b). \emph{Ggplotify: Convert plot to 'grob' or 'ggplot'
object}. Available at:
\url{https://CRAN.R-project.org/package=ggplotify}.

\leavevmode\hypertarget{ref-yu2020rvcheck}{}%
Yu, G. (2020c). \emph{Rvcheck: R/package version check}. Available at:
\url{https://CRAN.R-project.org/package=rvcheck}.

\leavevmode\hypertarget{ref-zeileis2019colorspace}{}%
Zeileis, A., Fisher, J. C., Hornik, K., Ihaka, R., McWhite, C. D.,
Murrell, P., Stauffer, R., and Wilke, C. O. (2019). colorspace: A
toolbox for manipulating and assessing colors and palettes. arXiv.org
E-Print Archive Available at: \url{http://arxiv.org/abs/1903.06490}.

\leavevmode\hypertarget{ref-zeileis2009escaping}{}%
Zeileis, A., Hornik, K., and Murrell, P. (2009). Escaping RGBland:
Selecting colors for statistical graphics. \emph{Computational
Statistics \& Data Analysis} 53, 3259--3270.
doi:\href{https://doi.org/10.1016/j.csda.2008.11.033}{10.1016/j.csda.2008.11.033}.

\end{document}

### 2020-07-13 ms/ms.tex
r1

@@
 
+\usepackage{booktabs}
 \usepackage{longtable}
+\usepackage{soul}
 \usepackage{tikz}
@@
 
-\title[Stomata tradeoff photosynthesis for defense]{A stomatal model of anatomical tradeoffs between photosynthesis and
-pathogen defense}
+\title[Stomatal anatomy and pathogen colonization]{A stomatal model of anatomical tradeoffs between gas exchange and
+pathogen colonization}
 \author[\firstAuthorLast]{\Authors}
@@
 (Hetherington and Woodward, 2003; Berry et al., 2010; Chater et al.,
-2017), but many plant pathogens take advantage of these chinks in the
+2017), but many foliar pathogens take advantage of these chinks in the
 leaf cuticular armor to infect prospective hosts (Zeng et al., 2010;
-McLachlan et al., 2014; Melotto et al., 2017). The density and size of
-stomata set the anatomical maximum rate of stomatal conductance to
-CO\(_2\) and water vapor (Brown and Escombe, 1900; Parlange and
-Waggoner, 1970; Franks and Farquhar, 2001; Franks and Beerling, 2009b;
-Lehmann and Or, 2015; Sack and Buckley, 2016; Harrison et al., 2019),
-but the pore area shrinks and expands in response to internal and
-external factors to regulate gas exchange dynamically (Buckley, 2019).
-Many plant pathogens, including viruses (Murray et al., 2016), bacteria
-(Melotto et al., 2006; Underwood et al., 2007), protists (Fawke et al.,
-2015), and fungi (Hoch et al., 1987; Zeng et al., 2010) use stomatal
-pores to gain entry into the leaf. Since stomatal conductance is a major
-limitation on photosynthesis (Farquhar and Sharkey, 1982; Jones, 1985)
-while pathogens reduce fitness, this sets up a potential tradeoff
-between increased photosynthesis and defense against pathogens. Although
-there have been many empirical studies on the effect that pathogens have
-on stomata, there is no theoretical framework in which to place these
-findings. Lack of a theoreatical framework makes it difficult to answer
-general questions about how selection for pathogen defense constrains
-maximum rates of gas exchange.
-
-Stomatal anatomy is the key link between gas exchange and pathogen
-colonization. The density and size of stomata not only determines the
-theoretical maximum stomatal conductance (\gsmax), but is also
-proportional to the operational stomatal conductance (\gsop) in many
-circumstances (Franks et al., 2009, 2014; Dow et al., 2014a; McElwain et
-al., 2016; Murray et al., 2019). Therefore, I use anatomical \gsmax{} as
-a proxy for \gsop{} and do not address dynamic changes in stomatal
-aperture (see \protect\hyperlink{discussion}{Discussion}). Harrison et
-al. (2019) recently reviewed the relationship between stomatal anatomy
-and gas exchange in detail.
-
-Many pathogens rely on stomata to gain entry into the leaf, but it is
-unclear how anatomical traits that determine \gsmax{} (size and density)
-affect the ability of pathogens to colonize leaves. The impact of
-pathogens on host fitness is complex, but after a pathogen reaches a
-host, the first major step is colonization. Once infected, a pathogen
-can reduce fitness. Susceptible hosts can lose much of their biomass or
-die, but even resistant hosts must allocate resources to defense or
-reduce photosynthesis because of necrosis around sites of infection.
-Plants can limit colonization physiologically by closing stomata after
-they recognize pathogens, called stomatal defense (Melotto et al.,
-2017). Anatomy may be another layer of defense; plants can reduce
-pathogen colonization by developing leaves with lower stomatal density
-and/or size. Infection increases in leaves with higher stomatal density
-(McKown et al., 2014; Tateda et al., 2019; Dutton et al., 2019; Fetter
-et al., 2019), which suggests that both anatomical and physiological
-responses (stomatal closure) affect host colonization. By the same
-logic, if stomatal density were held constant, larger stomata should
-also increase colonization because they occupy more area. One key
-difference between physiological and anatomical defenses is that
-stomatal closure reduces gas exchange transiently, whereas anatomy
-constrains gas exchange throught the life of the leaf. Because we do not
-understand the relationship between stomatal anatomy and colonization
-well, we cannot predict the relationship between \gsmax{} and infection.
-
-If stomatal size and density affect pathogen colonization, selection to
-limit colonization may shape stomatal size-density scaling
-relationships. Botanists have long recognized that stomatal size and
-density are inversely correlated (Weiss, 1865; Tichá, 1982; Hetherington
-and Woodward, 2003; Sack et al., 2003; Franks and Beerling, 2009a;
-Brodribb et al., 2013; de Boer et al., 2016), but the evolutionary
+McLachlan et al., 2014; Melotto et al., 2017). The stomatal and
+mesophyll conductance to CO\(_2\) are two major limits to photosynthesis
+(Lawson et al., 2018; Flexas et al., 2018) that are partially determined
+by stomatal anatomy. Since CO\(_2\) conductance limits photosynthesis
+(Farquhar and Sharkey, 1982; Jones, 1985) and pathogen infection can
+reduce fitness (Gilbert, 2002), this sets up a potential tradeoff
+between increased photosynthesis and defense against pathogens mediated
+by stomatal anatomy (McKown et al., 2014; Tateda et al., 2019; Dutton et
+al., 2019; Fetter et al., 2019). For example, plants could increase
+photosynthetic rate by developing more stomata, but more stomata could
+result in more pathogen colonization. The optimal stomatal density,
+size, and arrangement on the leaf will depend on the fitness gains from
+increased gas exchange and fitness losses imposed by foliar pathogens,
+both of which depend on the environment. In the next two paragraphs I
+will review the relationship between stomatal anatomy, gas exchange, and
+foliar pathogen colonization. Then I will discuss why two anatomical
+traits, stomatal size and density, might be crucial components of a
+broader tradeoff between photosynthesis and pathogen defense.
+
+The stomatal density and maximum pore area set an anatomical upper limit
+on stomatal conductance (Brown and Escombe, 1900; Parlange and Waggoner,
+1970; Franks and Farquhar, 2001; Franks and Beerling, 2009b; Lehmann and
+Or, 2015; Sack and Buckley, 2016; Harrison et al., 2019), but stomatal
+shape, distribution, and patterning also affect gas exchange. Smaller
+guard cells and dumbbell-shaped stomata of grasses can respond faster to
+environmental changes (Drake et al., 2013), but responsiveness is
+further modulated by subsidiary cell anatomy and physiology (Franks and
+Farquhar, 2007; Raissig et al., 2017; Gray et al., 2020). Stomatal
+clustering reduces gas exchange and photosynthesis because adjacent
+stomata interfere with one another (Dow et al., 2014b), diffusion shells
+overlap (Lehmann and Or, 2015), and limitations on lateral diffusion of
+CO\(_2\) in the mesophyll (Lawson and Blatt, 2014 and references
+therein). However, sparse clusters of small stomata could allow a leaf
+with low rates of gas exchange to have faster stomatal response compared
+to a leaf with large, low-density stomata (Papanatsiou et al., 2017).
+Leaves with stomata on both lower and upper surfaces (amphistomatic or
+amphistomatous) supply more CO\(_2\) to the mesophyll than hypostomatic
+(aka hypostomatous) leaves that only have stomata on the lower surface
+(Parkhurst, 1978; Gutschick, 1984; Parkhurst and Mott, 1990; Oguchi et
+al., 2018). In addition to anatomy, the pore area shrinks and expands in
+response to internal and external factors to regulate gas exchange
+dynamically (Buckley, 2019). For example, stomata typically open during
+the day and close at night in C\(_3\)/C\(_4\) plants, but the opposite
+is true for CAM plants. Shade, high vapor pressure deficits, dry soil
+and other factors can cause stomata to (partially) close even in the
+middle of the day. Variation in how stomata respond to internal and
+external signals may explain as much of the variation in gas exchange
+across leaves as anatomy (Lawson and Blatt, 2014).
+
+Many types of foliar pathogens, including viruses (Murray et al., 2016),
+bacteria (Melotto et al., 2006; Underwood et al., 2007), protists (Fawke
+et al., 2015), and fungi (Hoch et al., 1987; Zeng et al., 2010) use
+stomatal pores to gain entry into the leaf. For example, rust fungi
+hyphae recognize the angle at which guard cells project from the leaf
+surface and use it as a cue for appressorium formation (Allen et al.,
+1991). Oomycete pathogens can target open stomata on a leaf (Kiefer et
+al., 2002). Plants can limit colonization through innate immunity,
+called stomatal defense (recently reviewed in Melotto et al., 2017), by
+closing stomata after they recognize microbe-associated molecular
+patterns (MAMPs) on pathogen cells. Some bacterial pathogens have
+responded by evolving the ability to prevent stomatal closure,
+increasing their colonization of the leaf interior (Melotto et al.,
+2006). In addition to stomatal closure, anatomical changes in stomatal
+density and/or size might provide another layer of defense against
+pathogen colonization. For example, infection increases in leaves with
+higher stomatal density (McKown et al., 2014; Tateda et al., 2019;
+Dutton et al., 2019; Fetter et al., 2019). The positive effect of
+stomatal density on infection suggests that infection is limited by the
+number or size of locations for colonization, meaning that many
+individual pathogens must usually be unable to find stomata or other
+suitable locations for colonization. This is actually somewhat
+surprising given the ability of some pathogens to search for and sense
+stomata (see above).
+
+Stomatal anatomy could be a key link between gas exchange and pathogen
+colonization. Although many anatomical factors and stomatal movement
+affect gas exchange (see above), here I focus on the density and size of
+stomata in a hypostomatous leaf. Stomatal size refers to both the area
+of guard cells when fully open, from which one can calculate the pore
+area for gas exchange (see \protect\hyperlink{model}{Model}). For
+simplicity, I model a hypostomatous leaf, but consider the implications
+for amphistomatous leaves in the
+\protect\hyperlink{discussion}{Discussion}. Stomatal size and density
+not only determine the theoretical maximum stomatal conductance
+(\gsmax), but is also proportional to the operational stomatal
+conductance (\gsop) in many circumstances (Franks et al., 2009, 2014;
+Dow et al., 2014a; McElwain et al., 2016; Murray et al., 2019). \gsop{}
+is the actual stomatal conductance of plants in the field and is almost
+always below \gsmax{} because stomata are usually not fully open.
+Although they are not the same, the strong empirical relationship
+between \gsmax{} and \gsop{} means that anatomical \gsmax{} can be used
+as a proxy for \gsop{} without explicitly modeling dynamic changes in
+stomatal aperture (see \protect\hyperlink{discussion}{Discussion}).
+Stomatal size and density have also been measured on many more species
+than stomatal responsiveness, which may make it easier to test
+predictions.
+
+After a pathogen reaches a host, it must survive on the leaf surface and
+colonize the interior (Beattie and Lindow, 1995; Tucker and Talbot,
+2001). For analytical tractability, I restrict the focus here to
+colonization by a pathogen using a random search on a leaf without
+stomatal defense (i.e.~a leaf that cannot recognize pathogens and close
+stomata). Obviously, these simplifications ignore a lot of important
+plant-pathogen interaction biology. In the
+\protect\hyperlink{discussion}{Discussion}, I delve further into these
+limitations and suggest future work to overcome these limitations. In
+order for pathogen-mediated selection on stomatal anatomy, I assume that
+the pathogen reduces host fitness once it colonizes (Gilbert, 2002).
+Susceptible hosts can lose much of their biomass or die, but even
+resistant hosts must allocate resources to defense or reduce
+photosynthesis because of defoliation, biotrophy, or necrosis around
+sites of infection (Bastiaans, 1991; Mitchell, 2003).
+
+The purpose of this study is to develop a theoretical framework to test
+whether variation in stomatal size and density arises from a tradeoff
+between gas exchange and pathogen colonization. Since stomatal size and
+density affect both gas exchange and pathogen colonization, selection to
+balance these competing demands could shape stomatal size-density
+scaling relationships. Botanists have long recognized that stomatal size
+and density are inversely correlated (Weiss, 1865; Tichá, 1982;
+Hetherington and Woodward, 2003; Sack et al., 2003; Franks and Beerling,
+2009a; Brodribb et al., 2013; Boer et al., 2016), but the evolutionary
 origin of this relationship is not yet known. Here I argue that
-pathogens could shape selection on this relationship. Explanations for
-inverse size-density scaling are usually cast in terms of preserving
-\gsmax{} and/or total epidermal area allocated to stomata (\fs; de Boer
-et al., 2016) because there are many combinations of stomatal size and
+deleterious effects of pathogen infection could shape selection on this
+relationship. Explanations for inverse size-density scaling are usually
+cast in terms of preserving \gsmax{} and/or stomatal cover (\fs),
+defined at the fraction of epidermal area allocated to stomata (Boer et
+al., 2016), because there are many combinations of stomatal size and
 density that have same \gsmax{} or same \fs{}:
@@
 
-\(b\) and \(m\) are assumed to be biophysical and morphological
-constants, \emph{sensu} Sack and Buckley (2016) (see
-\protect\hyperlink{supplementary-material}{Supplementary Material}). If
-size and density also affect pathogen colonization, then selection from
-foliar pathogens could significantly alter the size-density scaling
+\(D\) and \(S\) are stomatal stomatal density and size, respectively
+(see Table 1 for a glossary of mathematical symbols and units). \(b\)
+and \(m\) are assumed to be biophysical and morphological constants,
+\emph{sensu} Sack and Buckley (2016) (see
+\protect\hyperlink{supplementary-material}{Supplementary Material}).
+\fs{} is proportional to the more widely used stomatal pore area index
+{[}Sack et al. (2003); see
+\protect\hyperlink{supplementary-material}{Supplementary Material}{]}.
+If size and density also affect pathogen colonization, then selection
+from foliar pathogens could significantly alter the size-density scaling
 relationship. The empirical size-density scaling relationship is linear
@@
 How would adding pathogens alter these predicted scaling relationships?
-For simplicity, imagine two environments, one without foliar pathogens
+For simplicity, consider two environments, one without foliar pathogens
 and one with lots. In the absence of foliar pathogens, we expect
 size-density scaling to preserve \gsmax{}, \fs{}, or some least-cost
-combination of them. What happens when we introduce pathogens? Assuming
-that stomatal size and density increase pathogen colonization, then
-selection will favor reduced size and/or density. This would change the
-intercept \(\alpha\) but not the slope. The effect of foliar pathogens
-on the slope depends on the relationship between size, density, and
-probability of colonization. If the probability of colonization is
-proportional to the product of \emph{linear} stomatal size
-(\(S ^ {0.5}\)) and density (\(\propto D S ^ {0.5}\) as for \gsmax) then
-it has the same effect on the slope as \gsmax{} because there are many
-combinations of \(D\) and \(S ^ {0.5}\) that have same probability of
-colonization. If the probability of colonization is proportional to the
-product of \emph{areal} stomatal size (\(S\)) and density
-(\(\propto D S\) as for \fs) then it has the same effect on the slope as
-\fs{} because there are many combinations of \(D\) and \(S\) that have
-same probability of colonization. Alternatively, the probability of
-colonization may have a different scaling relationsip (neither 0.5 nor
-1) or may be nonlinear on a log-log scale. Unlike \gsmax{} and \fs{}, we
-do not have theory to predict a stomatal size-density relationship that
-preserves the probability of colonization.
+combination of them. What happens when we introduce pathogens? If
+stomatal size and density increase pathogen colonization, then selection
+will favor reduced size and/or density. This would change the intercept
+\(\alpha\) but not the slope. The effect of foliar pathogens on the
+slope depends on the relationship between size, density, and probability
+of colonization. If the probability of colonization is proportional to
+the product of \emph{linear} stomatal size (\(S ^ {0.5}\)) and density
+(\(\propto D S ^ {0.5}\) as for \gsmax) then it has the same effect on
+the slope as \gsmax{} because there are many combinations of \(D\) and
+\(S ^ {0.5}\) that have same probability of colonization. If the
+probability of colonization is proportional to the product of
+\emph{areal} stomatal size (\(S\)) and density (\(\propto D S\) as for
+\fs) then it has the same effect on the slope as \fs{} because there are
+many combinations of \(D\) and \(S\) that have same probability of
+colonization. Alternatively, the probability of colonization may have a
+different scaling relationship (neither 0.5 nor 1) or may be nonlinear
+on a log-log scale. Unlike \gsmax{} and \fs{}, we do not have theory to
+predict a stomatal size-density relationship that preserves the
+probability of colonization.
 
 In summary, the physical relationship between stomatal size, density,
-and conductance is well established (Harrison et al., 2019). The same
-traits likely affect the probability of pathogen colonization, but we do
-not have a theoretical model that makes quantitative predictions. The
-inverse stomatal size-density relationship has usually been explained in
-terms of preserving stomatal conductance and/or stomatal cover, but
-selection by pathogens might alter scaling. To address these gaps, the
-goals of this study are to 1) introduce a spatially explicit model
-pathogen colonization on the leaf surface; 2) use the model to predict
-the relationship between \gsmax, \fs{}, and the probability of
+and conductance is well established (Harrison et al., 2019). Size and
+density also likely affect the probability of pathogen colonization, but
+we do not have a theoretical model that makes quantitative predictions.
+The inverse stomatal size-density relationship has usually been
+explained in terms of preserving stomatal conductance and/or stomatal
+cover, but selection by pathogens might alter scaling. To address these
+gaps, the goals of this study are to 1) introduce a spatially explicit
+model pathogen colonization on the leaf surface; 2) use the model to
+predict the relationship between \gsmax, \fs, and the probability of
 colonization; 3) work out what these relationships predict about
-stomatal size-density scaling.
+stomatal size-density scaling. I analyzed an idealized, spatially
+explicit \protect\hyperlink{model}{Model} of how a pathogen lands on a
+leaf and finds a stomate to colonize the leaf using a random search. To
+my knowledge, this is the first model that makes quantitative
+predictions about the relationship between stomatal anatomy, the
+probability of colonization, and their impact on stomatal size-density
+scaling.
 
@@
 
-In this section, I introduce a spatially explicit model of pathogen
-colonization on a leaf surface. I explain the model structure and
-assumptions here; the
+In this section, I introduce a spatially explicit model of successful
+pathogen colonization on a leaf surface. I explain the model structure
+and assumptions here; the
 \protect\hyperlink{materials-and-methods}{Materials and Methods} section
@@
 pathogen (virus, bacterium, fungus, etc.) and the specific biological
-details of how it moves (biotrophy). For example, motile bacterial cells
-can land and move around where fungi may germinate from a cyst and grow
-until they form an appresorium for infection. These very different
-biotropic movements on the leaf are treated identically here. I used
-Sympy version 1.3 (Meurer et al., 2017) for symbolic derivations.
+details of how it moves. For example, motile bacterial cells can land
+and move around (Beattie and Lindow, 1995) whereas fungi may germinate
+from a cyst and grow until they form an appresorium for infection
+(Tucker and Talbot, 2001). These very different tropic movements on the
+leaf are treated identically here. I do not model photosynthesis
+explicitly, but assume that stomatal conductance limits carbon fixation,
+even though the relationship is nonlinear. I used Sympy version 1.6.1
+(Meurer et al., 2017) for symbolic derivations.
 
@@
 R\strut
-\end{minipage} & \begin{minipage}[b]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\raggedright
 Units\strut
-\end{minipage} & \begin{minipage}[b]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{0.59\columnwidth}\raggedright
 Description\strut
@@
 \begin{minipage}[t]{0.09\columnwidth}\raggedright
-\(D,d\)\strut
+\(D\)\strut
 \end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
-\texttt{D,\ d}\strut
+\texttt{D}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
+\(\textrm{mm}^{-2}\)\strut
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
+stomatal density\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.09\columnwidth}\raggedright
+\(d\)\strut
 \end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\texttt{d}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\textrm{mm}^{-2}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
-stomatal density (\(d = \text{log}~D\))\strut
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
+stomatal density (log-scale, \(d = \text{log}~D\))\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.09\columnwidth}\raggedright
-\(f_s\)\strut
+\fs\strut
 \end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
 \texttt{f\_s}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 none\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
 stomatal cover (\(f_s = D S\))\strut
@@
 \texttt{g\_smax}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\text{mol}~\text{m}^{-2}~\text{s}^{-1}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
-anatomical maximum stomatal conductance\strut
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
+theoretical maximum stomatal conductance\strut
 \end{minipage}\tabularnewline
@@
 \texttt{g\_sop}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\text{mol}~\text{m}^{-2}~\text{s}^{-1}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
 operational stomatal conductance\strut
@@
 \texttt{H}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\mu\textrm{m} ^{-1}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
 death rate of pathogen on leaf surface\strut
@@
 \texttt{R}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
+\(\mu\textrm{m}\)\strut
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
+stomatal radius (\(S = \pi R ^ 2\))\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.09\columnwidth}\raggedright
+\(S\)\strut
 \end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
-\(\mu\textrm{m}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
-stomatal radius (\(S = 2 \pi R ^ 2\))\strut
+\texttt{S}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
+\(\mu\textrm{m}^2\)\strut
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
+stomatal size\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.09\columnwidth}\raggedright
-\(S,s\)\strut
+\(s\)\strut
 \end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
-\texttt{S,\ s}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\texttt{s}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\mu\textrm{m}^2\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
-stomatal size (\(s = \text{log}~S\))\strut
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
+stomatal size (log-scale, \(s = \text{log}~S\))\strut
 \end{minipage}\tabularnewline
@@
 \texttt{theta\_i}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 radians\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
 angles between pathogen (\(x_p,y_p\)) and lines tangent to the
-circumfrence of stomate \(i\)\strut
+circumference of stomate \(i\)\strut
 \end{minipage}\tabularnewline
@@
 \texttt{U}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\mu\textrm{m}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
 interstomatal distance\strut
@@
 \texttt{v\_i}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\mu\textrm{m}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
 distance between pathogen (\(x_p,y_p\)) and stomate \(i\)\strut
@@
 \texttt{x\_i,y\_i}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\mu\textrm{m}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
 position of stomate \(i\)\strut
@@
 \texttt{x\_p,y\_p}\strut
-\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
 \(\mu\textrm{m}\)\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.59\columnwidth}\raggedright
 starting position of pathogen\strut
@@
 2015). Here I assume that stomata are arrayed in an equilateral
-triangular grid with a density \(D\) and size (area) \(S\). This
+triangular grid with a density \(D\) and size (area) \(S\) on the
+abaxial surface only, since most leaves are hypostomatoues {[}Muir
+(2015); but see \protect\hyperlink{discussion}{Discussion}{]}. This
 assumption ignores veins, trichomes, and within-leaf variation in
@@
 diagonal of the hexagon in \(\mu\textrm{m}\) that forms an equal area
-boundary between neighbording stomata. The area of a hexagon is
+boundary between neighboring stomata. The area of a hexagon is
 \(A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2\). By definition the
@@
 result previously. I also make the simplifying assumption that stomata
-are perfectly circular with radius \(R\). This may be approximately true
-for fully open stomata with kidney-shaped guard cells. Although I assume
+are perfectly circular with radius \(R\) when fully open. This may be
+approximately true for fully open stomata with kidney-shaped guard cells
+(Sack and Buckley, 2016 and references therein). Although I assume
 stomata are circular here, in calculating \gsmax, I assume typical
-allometric relationships between length, width, and pore area (see
-\protect\hyperlink{supplementary-material}{Supplementary Material}).
+allometric relationships between length, width, and pore area {[}Sack
+and Buckley (2016); see
+\protect\hyperlink{supplementary-material}{Supplementary Material}{]}.
 
@@
 
-Now imagine that a pathogen lands at a uniform random position within
-the focal region and must arrive at a stomate to colonize. If it lands
-on a stomate, then it infects the leaf with probability 1; if it lands
-between stomata, then it infects the leaf with probability
-\(p_\text{locate}\). This is the probability that it locates a stomate,
-which I will derive below. The probabilities of landing on or between a
-stomate are \fs{} and \(1 - f_S\), respectively. Hence, the total
-probability of colonization is:
+Since stomata are arrayed in a homogeneous grid, we can focus on single
+focal triangle (Figure \ref{fig:fig1}b-c). Suppose that an individual
+pathogen (e.g.~bacterial cell or fungal spore) lands at a uniform random
+position within the focal triangle and must arrive at a stomate to
+colonize. If it lands on a stomate, then it infects the leaf with
+probability 1; if it lands between stomata, then it infects the leaf
+with probability \(p_\text{locate}\). This is the probability that it
+locates a stomate, which I will derive below. The probabilities of
+landing on or between a stomate are \fs{} and \(1 - f_S\), respectively.
+Hence, the total probability of colonization is:
 
@@
 form an equilateral triangular grid (see above), we can extrapolate what
-happens in a focal triangle (Figure \ref{fig:fig1}b) by symmetry.
+happens in the focal triangle (Figure \ref{fig:fig1}b) by symmetry.
 Further, since an equilateral triangle can be broken up into six
@@
 \(p_\text{locate}\) in this focal area. This implicitly assumes that the
-probability of colonzing stomata outside the focal area is 0 because
-they are too far away.
-
-Imagine that the pathogens lands in position (\(x_p,y_p\)) within the
+probability of colonizing stomata outside the focal area is 0 because
+they are too far away. This assumption may be unrealistic for larger
+pathogens, such as fungi, whose hyphae can travel longer distances on
+the leaf surface (Brand and Gow, 2012). In
+\protect\hyperlink{appendix-1-spatially-implicit-model}{Appendix 1:
+Spatially implicit model} I derive a simpler, but spatially
+\emph{implicit} model that relaxes the assumption the pathogens must
+colonize a stomate within their focal triangle.
+
+Consider a pathogen that lands in position (\(x_p,y_p\)) within the
 triangle. The centroid of the triangle is at position (\(x_c,y_c\)) and
@@
 at very low density, then a pathogen may die before it reaches a stomate
-because of UV, dessication, or another factor. I included this effect by
+because of UV, desiccation, or another factor. I included this effect by
 assuming the probability of reaching a stomate declines exponentially at
@@
 
-The probability of locating a stomate as a function of \(x_p\) and
-\(y_p\) (\(f_\text{locate}(x_p,y_p)\)) is the sum of the angles divided
-by \(2\pi\), discounted by their distance from the stomate:
+The probability of locating a stomate as a function of pathogen position
+(\(x_p\) and \(y_p\)) is the sum of the angles divided by \(2\pi\),
+discounted by their distance from the stomate:
 
@@
 
-When \(H = 0\), \(p_\text{locate}\) is the fraction of angles that lead
-from \((x_p,y_p)\) to a stomate. When \(H > 0\), \(p_\text{locate}\) is
-proportional to this fraction, but less than it depending on stomatal
-density, size, and starting location of the pathogen.
+When there is no pathogen death (\(H = 0\)), \(p_\text{locate}\) is the
+fraction of angles that lead from \((x_p,y_p)\) to a stomate. When
+\(H > 0\), \(p_\text{locate}\) is proportional to this fraction, but
+less than it depending on stomatal density, size, and starting location
+of the pathogen.
 
@@
 the midpoint of baseline \((U/2,0)\), and the centroid of the focal
-trinagle \((U/2, \sqrt{3} / 6 U)\) (Figure \ref{fig:fig1}c).
+triangle \((U/2, \sqrt{3} / 6 U)\) (Figure \ref{fig:fig1}c).
 Colonization occurs with probability 1 if the pathogen lands in the
@@
 The \protect\hyperlink{model}{Model} calculates a probability of host
-colonization (Equation \ref{eq:p_colonize}) as a function stomatal
+colonization (Equation \ref{eq:p_colonize}) as a function of stomatal
 density, size, and position of a pathogen on the leaf. I solved
-\(p_\text{colonize}\) using the \texttt{integral2()} function in the
-\textbf{pracma} package version 2.2.5 (Borchers, 2019) for numerical
-integration. I used R version 3.6.1 (R Core Team, 2019) for all analyses
-and wrote the paper in \textbf{rmarkdown} version 1.17 (Xie et al.,
-2018; Allaire et al., 2019). Source code is deposited on GitHub
-(\url{https://github.com/cdmuir/stomata-tradeoff}) and will be arcived
+\(p_\text{colonize}\) by importing symbolic derivations from Sympy into
+R with \textbf{reticulate} version 1.16 (Ushey et al., 2020) and used
+the \texttt{integral2()} function in the \textbf{pracma} package version
+2.2.9 (Borchers, 2019) for numerical integration. I used R version 4.0.2
+(R Core Team, 2020) for all analyses and wrote the paper in
+\textbf{rmarkdown} version 2.3 (Xie et al., 2018; Allaire et al., 2020).
+Citations for additional R software packages are in Appendix 2. Source
+code is deposited on GitHub
+(\url{https://github.com/cdmuir/stomata-tradeoff}) and will be archived
 on Zenodo upon publication.
@@
 I calculated \(p_\text{colonize}\) over a biologically plausible grid of
-stomatal size and density for hypostomatous species based on de Boer et
-al. (2016). Stomatal density ranges from
-\(10^1 - 10^{3.5}~\text{mm}^{-2}\); stomatal size ranges from
+stomatal size and density for hypostomatous species based on Boer et al.
+(2016). Stomatal density (\(D\)) ranges from
+\(10^1 - 10^{3.5}~\text{mm}^{-2}\); stomatal size (\(S\)) ranges from
 \(10^1 - 10^{3.5}~\mu\text{m}^2\). I only considered combinations of
-size and density where \fs{} was less than 1/3. For simplicity, I have
-not extended the current analysis to amphistomatous leaves. I crossed
-stomatal traits with three levels of \(H \in \{0, 0.01, 0.1\}\). When
-\(H = 0\), a pathogen perists indefinitely on the leaf surface.
+size and density where stomatal cover (\fs) was less than 1/3, which is
+close to the upper limit in terrestrial plants (Boer et al., 2016). I
+crossed stomatal traits with three levels of \(H \in \{0, 0.01, 0.1\}\).
+When \(H = 0\), a pathogen persists indefinitely on the leaf surface.
 \(H = 0.01\) and \(H = 0.1\) correspond to low and high death rates,
@@
 The stomatal size-density scaling relationship can be explained in terms
-of preserving a constant \gsmax{} that is proportional to
-\(D S ^ {0.5}\) when \(bm\) is constant (Equation \ref{eq:gsmax}). In
-other words, there are infinitely many combinations of \(D\) and
-\(S ^ {0.5}\) with the same \gsmax. If \gsmax{} is held constant at
-\(C_g\), then the resulting size-density scaling relationship on a
-log-log scale is:
+of preserving a constant stomatal conductance (\gsmax) that is
+proportional to \(D S ^ {0.5}\) when \(bm\) is constant (Equation
+\ref{eq:gsmax}). In other words, there are infinitely many combinations
+of \(D\) and \(S ^ {0.5}\) with the same \gsmax. If \gsmax{} is held
+constant at \(C_g\), then the resulting size-density scaling
+relationship on a log-log scale is:
 
 \[ d = c_g - 0.5 s \] where lowercase variables are log-transformed
-equivalents of their uppercase counterparts. The scaling exponent
-\(\beta_g = 0.5\) preseves \(C_g\).
-
-Next, imagine there is similarly a scaling exponent \(\beta_p\) that
-preserves \(p_\text{colonize}\) for the product \(D S^{\beta_p}\). If
+equivalents of their uppercase counterparts (Table 1). The scaling
+exponent \(\beta_g = 0.5\) preserves \(C_g\).
+
+Next, suppose there is a scaling exponent \(\beta_p\) that preserves
+\(p_\text{colonize}\) for the product \(D S^{\beta_p}\). If
 \(\beta_p = 0.5\), then \(p_\text{colonize}\) is always proportional to
 \gsmax. If \(\beta_p > 0.5\), small, densely packed stomata would be
-better defended (lower \(p_\text{colonize}\)) compared to larger,
-sparsely spaced stomata with the same \gsmax. If \(\beta_p < 0.5\),
-small, densely packed stomata would be less defended (higher
-\(p_\text{colonize}\)) compared to larger, sparsely spaced stomata with
-the same \gsmax. I refer to the three outcomes (\(\beta_p = 0.5\),
-\(\beta_p < 0.5\), and \(\beta_p > 0.5\)) as iso-, hypo-, and
-hyper-conductance, respectively. I was unable to solve analytically for
-\(\beta_p\), so I numerically calculated isoclines of
+more resistant to colonization (lower \(p_\text{colonize}\)) compared to
+larger, sparsely spaced stomata with the same \gsmax. If
+\(\beta_p < 0.5\), small, densely packed stomata would be less defended
+(higher \(p_\text{colonize}\)) compared to larger, sparsely spaced
+stomata with the same \gsmax. I refer to the three outcomes
+(\(\beta_p = 0.5\), \(\beta_p < 0.5\), and \(\beta_p > 0.5\)) as iso-,
+hypo-, and hyper-conductance, respectively. I was unable to solve
+analytically for \(\beta_p\), so I numerically calculated isoclines of
 \(p_\text{colonize}\) over the grid of \(D\) and \(S\) values described
@@
 relationships at a constant
-\(p_\text{colonize} \in \{0.025, 0.05, 0.1, 0.2, 0.4\}\) for
+\(p_\text{colonize} \in \{0.025, 0.1, 0.4\}\) for
 \(H \in \{0, 0.01, 0.1\}\).
@@
 \addcontentsline{toc}{section}{Results}
-
-I analyzed an idealized, spatially explicit
-\protect\hyperlink{model}{Model} of how a pathogen lands on a leaf and
-finds a stomate to colonize the leaf using a random search. To my
-knowledge, this is the first model that makes quantitative predictions
-about the relationship between stomatal anatomy, the probability of
-colonization, and their impact on stomatal size-density scaling.
 
@@
 
-The probability of colonization (\(p_\text{colonize}\)) is not simply a
-one-to-one relationship between the fraction of epidermal area allocated
-to stomata (\fs). At low \fs, \(p_\text{colonize}\) increases faster
-rapdily relative to \fs{} at first Figure \ref{fig:fig3}a). At higher
-\fs, the \(p_\text{colonize}\) increases linearly with \fs. When
-\(H = 0\), any combination of stomatal size (\(S\)) and density (\(D\))
-with the same \fs{} have the same effect on \(p_\text{colonize}\). When
-\(H > 0\), pathogens are less likely to land close enough to a stomate
-to infect before dying, so \(p_\text{colonize}\) is closer to \fs{}
-(Figure \ref{fig:fig3}a). Furthermore, \(p_\text{colonize}\) depends on
-\(D\) and \(S\), not just \fs. For the same \fs, leaves with greater
-\(D\) have higher \(p_\text{colonize}\) (Figure \ref{fig:fig3}a).
-Holding \fs{} constant, leaves with lower \(D\) and higher \(S\) will
-have a greater distance (\(v_i\)) between a pathogen and its stomata.
-When \(H > 0\), this extra distance leads more pathogens to die before
-they can find a stomate. In contrast to \fs, \(p_\text{colonize}\)
-increases at a greater than linear rate with \gsmax. Greater \(D\)
-(smaller \(S\)) is associated with lower \(p_\text{colonize}\) when
-\gsmax{} is held constant (Figure \ref{fig:fig3}b). This happens because
+The probability of colonization (\(p_\text{colonize}\)) is not simply
+proportional to stomatal cover (\fs). At low \fs, \(p_\text{colonize}\)
+increases rapidly relative to \fs{} at first Figure \ref{fig:fig3}a). At
+higher \fs, \(p_\text{colonize}\) increases linearly with \fs. When
+pathogens persist indefinitely (\(H = 0\)), any combination of stomatal
+size (\(S\)) and density (\(D\)) with the same \fs{} have the same
+effect on \(p_\text{colonize}\). When \(H > 0\), pathogens are less
+likely to land close enough to a stomate to infect before dying, so
+\(p_\text{colonize}\) is closer to \fs{} (Figure \ref{fig:fig3}a). The
+maximum \(p_\text{colonize}\) under the range of parameters considered
+was \(\sim 0.6\) when \(H = 0\) and \fs{} is at its maximum value of
+1/3. When \fs{} is low, \(p_\text{colonize}\) is also low. The
+relationship between \(p_\text{colonize}\), \fs, and \gsmax is
+qualitatively similar in the spatially implicit model, but the values
+for \(p_\text{colonize}\) are substantially higher because pathogens can
+potentially colonize any stomate on the leaf rather than only those in
+the focal triangle (see
+\protect\hyperlink{appendix-1-spatially-implicit-model}{Appendix 1:
+Spatially implicit model} for more detail). Bear in mind that this is
+the probability for a single individual searching randomly; if enough
+individuals reach the leaf and/or they can actively find stomata, it's
+almost certain that at least some will colonize the leaf. However,
+reducing \(p_\text{colonize}\) may help plants limit the damage since
+fewer total individual pathogens will colonize the leaf interior.
+
+\(p_\text{colonize}\) is not directly proportional to \fs{} because it
+depends on \(D\) and \(S\) in quantitatively different ways (Figure
+\ref{fig:figS1}). For the same \fs, leaves with greater \(D\) have
+higher \(p_\text{colonize}\) (Figure \ref{fig:fig3}a). Holding \fs{}
+constant, leaves with lower \(D\) and higher \(S\) will have a greater
+distance (\(v_i\)) between a pathogen and its stomata. When \(H > 0\),
+this extra distance leads more pathogens to die before they can find a
+stomate. However, this result is inconsistent with the spatially
+implicit model (Appendix 1) because \(S\) and \(D\) have identical
+effects on \fs.
+
+In contrast to \fs, \(p_\text{colonize}\) increases at a greater than
+linear rate with stomatal conductance (\gsmax). Greater \(D\) (smaller
+\(S\)) is associated with lower \(p_\text{colonize}\) for a given value
+of \gsmax{} (Figure \ref{fig:fig3}b). This happens because
 \(p_\text{colonize}\) increases approximately linearly with \(S\)
-whereas \gsmax{} is proportional to \(S^{0.5}\).
+whereas \gsmax{} is proportional to \(S^{0.5}\). Therefore,
+\(p_\text{colonize}\) increases exponentially with \gsmax{} at all
+stomatal densities, but the rate of growth is lower at greater \(D\) for
+a given value of \gsmax.
 
@@
     \includegraphics{../figures/fig3.pdf}
-    \caption{\textbf{The probability of colonization increases with both stomatal cover and conductance} \textbf{a.} The probability of colonization ($p_\text{colonize}$, $y$-axis) initially increases rapidly with stomatal cover (\fs), then slows down to a linear relationship.
+    \caption{\textbf{The probability of colonization increases with both stomatal cover and conductance} I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of stomatal densities and sizes (see [Materials and Methods]), but a subset of results are shown here.
+Stomatal size and density determine stomatal cover (\fs; Equation \ref{eq:fs}) and theoretical maximum stomatal conductance (\gsmax; Equation \ref{eq:gsmax}).
+\textbf{a.} $p_\text{colonize}$ initially increases rapidly with \fs{} ($x$-axis), then slows down to a linear relationship.
 Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$).
-The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density when $H = 0$ (upper facet).
+The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density when $H = 0$ (upper facet), which is why the lines overlap.
 When $H > 0$, higher density (solid lines) increase $p_\text{colonize}$ (lower facets).
-\textbf{b.} $p_\text{colonize}$ increases expoentially with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is much lower at higher densities for a given \gsmax.
+\textbf{b.} $p_\text{colonize}$ increases exponentially with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is much lower at higher densities for a given \gsmax.
 The relationship between \gsmax{} and $p_\text{colonize}$ is similar for all values of $H$.}
@@
 The scaling relationship between \(S\) and \(D\) that preserves
-\(p_\text{colonize}\) is always greater 0.5 (hyper-conductance), but
-usually less than 1. When \(H = 0\), the scaling relationship is
+\(p_\text{colonize}\) is always greater than 0.5 (hyper-conductance),
+but usually less than 1. When \(H = 0\), the scaling relationship is
 essentially 1 (Figure \ref{fig:fig4}), which means that an increase
@@
 while \(p_\text{colonize}\) is constant, but this will still result in
-lower \gsmax.
+lower \gsmax. In the spatially implicit model, the size-density scaling
+exponent was always exactly 1 except when \(H = 0\) (Appendix 1).
 
@@
     \includegraphics{../figures/fig4.pdf}
-    \caption{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$).
-In each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.025 (lowest line), 0.05, 0.1, 0.2, or 0.4 (highest line).
+    \caption{\textbf{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$).} In each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.025 (lowest line), 0.1, or 0.4 (highest line).
 For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.5$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region.
@@
 stomatal conductance in nature (Murray et al., 2019). Despite the fact
-that many ecologically and economically significant plant pathogens
-infect through stomata, the relationship between stomatal anatomy and
-susceptibility to foliar pathogens is less clear than it is for gas
-exchange. To develop testable predictions, we need mathematical models
-that can clarify the potential for tradeoffs between stomatal
-conductance, stomatal cover, and disease resistance. I used a spatially
-explicit model of a pathogen searching for a stomate to colonize a host.
-From this \protect\hyperlink{model}{Model}, I derived predictions about
-the relationship between stomatal anatomy and disease resistance for the
-first time. The model predicts that the probability of colonization is
-not always proportional to the surface area of leaf covered by stomata
-(\fs), as one might intuitively predict. If the leaf surface is a
-hostile environment and pathogens have a limited time to search, lower
-stomatal density decreases the probability of colonization even if \fs{}
-is constant. However, \gsmax{} decreases proportionally more than the
-probability of colonization. The model reveals the potential for
-conflicting demands of maximizing disease resistance, minimizing
-stomatal cover, and maintaining stomatal conductance. Including the
-effect of anatomy on disease resistance therefore has the potential to
-change our understading of how stomatal size-density scaling evolves in
-land plants.
+that many foliar pathogens infect through stomata, the relationship
+between stomatal anatomy and resistance to foliar pathogens is less
+clear than it is for gas exchange. I used a spatially explicit model of
+a pathogen searching for a stomate to colonize a host. From this
+\protect\hyperlink{model}{Model}, I derived predictions about the
+relationship between stomatal anatomy and the probability of
+colonization, a component of disease resistance. The model predicts that
+the probability of colonization is not always proportional to the
+surface area of leaf covered by stomata (\fs), as one might intuitively
+predict. If the leaf surface is a hostile environment and pathogens have
+a limited time to search, lower stomatal density decreases the
+probability of colonization even if \fs{} is constant. However, \gsmax{}
+decreases proportionally more than the probability of colonization. The
+model highlights the potential for conflicting demands of minimizing
+pathogen colonization, minimizing stomatal cover, and maintaining
+stomatal conductance. Including the effect of anatomy on pathogen
+colonization therefore has the potential to change our understanding of
+how stomatal size-density scaling evolves in land plants.
 
 The model predicts that in most cases, increasing stomatal cover should
-lead to a proportonal increase in susceptibility, which is the implicit
-assumption of some empirical studies (e.g.~McKown et al. (2014); Tateda
-et al. (2019); Dutton et al. (2019); Fetter et al. (2019)). It also
-makes new, testable predictions that are less intuitive. At very low
-\fs, there is a rapid increase in susceptibility (Figure
-\ref{fig:fig3}a). If there are no stomata, the probability of
-colonization is 0, so the first few stomata dramatically increase the
-probability. This is unlikely to be significant for abaxial (lower) leaf
-surfaces, which usually have most of the stomata (Salisbury, 1928;
-Metcalfe and Chalk, 1950; Mott et al., 1984; Peat and Fitter, 1994;
-Jordan et al., 2014; Muir, 2015; Bucher et al., 2017; Drake et al.,
-2019). However, many adaxial (upper) leaf surfaces have zero or very few
-stomata. Using adaxial leaf surfaces, it should be possible to test if
-small changes in stomatal size or density have a larger effect on
-disease susceptibility when \fs{} is low. The nonlinear increase in
+lead to a proportional increase in colonization, which agrees with
+empirical studies (e.g.~McKown et al. (2014); Tateda et al. (2019);
+Dutton et al. (2019); Fetter et al. (2019)). It also makes new, testable
+predictions that are less intuitive (Table 2). At very low \fs, there is
+a rapid increase in colonization (Figure \ref{fig:fig3}a). If there are
+no stomata, the probability of colonization is 0, so the first few
+stomata dramatically increase the probability. This is less likely to be
+significant for abaxial (lower) leaf surfaces, which usually have most
+of the stomata (Salisbury, 1928; Metcalfe and Chalk, 1950; Mott et al.,
+1984; Peat and Fitter, 1994; Jordan et al., 2014; Muir, 2015; Bucher et
+al., 2017; Drake et al., 2019). However, many adaxial (upper) leaf
+surfaces have zero or very few stomata. Using adaxial leaf surfaces, it
+should be possible to test if small changes in stomatal size or density
+have a larger effect on pathogen colonization when \fs{} is low. Such
+experiments could use natural genetic variation (McKown et al., 2014) or
+mutant lines (Dow et al., 2014b). The nonlinear increase in
 \(p_\text{colonize}\) is less apparent when \(H > 0\) (Figure
 \ref{fig:fig3}a). A more hostile microenvironment (e.g.~drier, higher
-UV) should therefore reduce the effect of increased size or density as
-low \fs. If true, the dimishing marginal effect of \fs{} on colonization
-could explain why stomatal ratio on the upper and lower surface is
-bimodal (Muir, 2015). The initial cost of adaxial (upper) stomata is
-high, but if the benefits outweigh the costs, then equal stomatal
-densities on each surface maximize CO\(_2\) supply for photosynthesis
-(Parkhurst, 1978; Gutschick, 1984; Parkhurst and Mott, 1990).
-
-An effect of stomatal size and density on susceptibility to foliar
-pathogens could change our understanding of stomatal size-density
-scaling. Since allocating leaf epidermis to stomata may be costly
-(Franks and Farquhar, 2007; Assmann and Zeiger, 1987; Dow et al., 2014b;
-Lehmann and Or, 2015; Baresch et al., 2019), selection should favor
-leaves that achieve a desired \gsmax{} while minimizing \fs{} (de Boer
-et al., 2016). Because of their different scaling exponents (Equation
-\ref{eq:gsmax}, \ref{eq:fs}), smaller, densely packed stomata can
-achieve the same \gsmax{} at minimum \fs. However, many leaves have
-larger, sparsely packed stomata. Incorporating susceptibility to disease
-may explain why. If pathogens have a limited time to find stomata before
-dying (\(H > 0\)), then the scaling exponent between size and density
-that keeps \(p_\text{colonize}\) constant is between 0.5 and 1, the
-scaling exponents for \gsmax{} and \fs, respectively (Figure
-\ref{fig:fig4}). Greater density of smaller stomata can increase
-\gsmax{} while keeping \(p_\text{colonize}\) constant, but this will
-increase \fs. Conversely, \fs{} could decrease while keeping
-\(p_\text{colonize}\) constant, but this will decrease \gsmax. This sets
-up the potential for conflict between competing goals. The optimal
-stomatal size and density will therefore depend on the precise costs and
-benefits of infection, stomatal conductance, and stomatal cover. This
-may explain why many leaves have large, sparsely packed stomata despite
-the fact that they could achieve the same \gsmax{} and lower \fs{} with
-smaller, more densely packed stomata.
+UV) should therefore reduce the effect of increased size or density at
+low \fs. If true, the diminishing marginal effect of \fs{} on
+colonization could explain why stomatal ratio on the upper and lower
+surface is bimodal (Muir, 2015). The initial cost of adaxial (upper)
+stomata is relatively high, but if the benefits outweigh the costs, then
+equal stomatal densities on each surface maximize CO\(_2\) supply for
+photosynthesis (Parkhurst, 1978; Gutschick, 1984; Parkhurst and Mott,
+1990). The costs and benefits will certainly vary with environmental
+conditions as well. Future work should extend this model, which
+considered hypostomatous leaves, to address stomatal size and density in
+amphistomatous leaves, since leaf surfaces may differ in the type of
+pathogens present and microenvironment (McKown et al., 2014; Fetter et
+al., 2019).
+
+\textbf{Table 2} \textbar{} New testable model predictions and suggested
+experiments to test them.
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+\begin{minipage}[b]{0.50\columnwidth}\raggedright
+Model Prediction\strut
+\end{minipage} & \begin{minipage}[b]{0.44\columnwidth}\raggedright
+How to test it\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.50\columnwidth}\raggedright
+Increasing stomatal size and/or density will have a larger effect on
+pathogen colonization in leaves with low stomatal cover.\strut
+\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
+Compare the effect of changing stomatal size and/or density on pathogen
+colonization in leaves with low and high stomatal cover.\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.50\columnwidth}\raggedright
+Increasing stomatal size and/or density will have a smaller effect on
+pathogen colonization in ephiphytic environments more hostile to
+pathogens.\strut
+\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
+Compare the effect of changing stomatal size and/or density on pathogen
+colonization in more hostile environments (e.g.~drier, higher
+light)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.50\columnwidth}\raggedright
+When selection against pathogen colonization stronger, the stomatal
+size-density scaling exponent should be lower\strut
+\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
+Measure stomatal anatomy in environments that differ in pathogen
+colonization using comparative or experimental approaches\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+An effect of stomatal size and density on foliar pathogen colonization
+could change our understanding of stomatal size-density scaling. Since
+allocating leaf epidermis to stomata may be costly (Franks and Farquhar,
+2007; Assmann and Zeiger, 1987; Dow et al., 2014b; Lehmann and Or, 2015;
+Baresch et al., 2019), selection should favor leaves that achieve a
+desired \gsmax{} while minimizing \fs{} (Boer et al., 2016). Because of
+their different scaling exponents (Equation \ref{eq:gsmax},
+\ref{eq:fs}), smaller, densely packed stomata can achieve the same
+\gsmax{} at minimum \fs. However, many leaves have larger, sparsely
+packed stomata. Incorporating pathogen colonization may explain why. If
+pathogens have a limited time to find stomata before dying (\(H > 0\)),
+then the scaling exponent between size and density that keeps
+\(p_\text{colonize}\) constant is between 0.5 and 1, the scaling
+exponents for \gsmax{} and \fs, respectively (Figure \ref{fig:fig4}).
+Greater density of smaller stomata can increase \gsmax{} while keeping
+\(p_\text{colonize}\) constant, but this will increase \fs. Conversely,
+\fs{} could decrease while keeping \(p_\text{colonize}\) constant, but
+this will decrease \gsmax. This sets up the potential for conflict
+between competing goals. The optimal stomatal size and density will
+therefore depend on the precise costs and benefits of infection,
+stomatal conductance, and stomatal cover. This may explain why many
+leaves have large, sparsely packed stomata despite the fact that they
+could achieve the same \gsmax{} and lower \fs{} with smaller, more
+densely packed stomata.
 
@@
 interpreted as exact predictions, but rather as depicting qualitative
-relationships between stomatal anatomy and infection severity. The model
-is most applicable to diseases where the host has some resistance. The
+relationships between stomatal anatomy and infection severity. The
 energetic cost and lost photosynthetic capacity (closed stomata,
@@
 to the amount of infection. The actual fitness cost will be modulated by
-the number of pathogens landing on the leaf and the cost of infection.
-In environments with fewer or less virulent pathogens, the fitness cost
-of infection will be less than in environments with more abundant,
-virulent pathogens. The model is less relevant to very susceptible host
-plants that can be severely damaged or killed by a small number of
-colonizations that spread unchecked throughout the host tissue.
-
-The purpose of this model is to provide a general foundation to examine
-the relationship between stomatal size, density, and defense against
-foliar pathogens. In its generality, it overlooks interesting natural
-history and biologically important features of specific plants and their
-pathogens. For example, some pathogens actively seek out and find
-stomata (Kiefer et al., 2002), whereas I assumed that pathogens randomly
-orient themselves on the leaf. Including sensing should increase the
-probability of colonization. I also assumed that the plant does not
-respond by, for example, closing stomata when it senses a pathogen.
-However, plants can sense and close stomata when pathogens land on the
-leaf, but pathogens can pry stomata back open (Melotto et al., 2006).
-Future work on specific plant-pathogen interactions could build on this
-model by adding more biological realism to provide more precise
-predictions.
+the number of pathogens landing on the leaf and the cost of infection,
+all else being equal. In environments with fewer or less virulent
+pathogens, the fitness cost of infection will be less than in
+environments with more abundant, virulent pathogens. The model is less
+relevant to very susceptible host plants that can be severely damaged or
+killed by a small number of colonizations that spread unchecked
+throughout the host tissue.
 
@@
 resistance sets up a potential conflict between minimizing stomatal
-cover and maximizing stomatal conductance. The first prediction may be
-relatively straightforward to test experimentally with adaxial (upper)
-stomata that occur at low and moderate densities within the same or
-closely related species (Muir et al., 2014; McKown et al., 2014; Fetter
-et al., 2019). The second prediction about size-density scaling is more
-complex because we would need to know the relationships between
-colonization, stomatal cover, stomatal conductance, and fitness in
-natural conditions. Testing these predictions in a variety of species
-would help determine whether pathogens have played an important role
-shaping stomatal anatomy in land plants.
+cover and maximizing stomatal conductance. The first prediction is
+consistent with results in \emph{Populus trichocarpa} (McKown et al.,
+2014) and may be relatively straightforward to test experimentally with
+adaxial (upper) stomata that occur at low and moderate densities within
+the same or closely related species (Muir et al., 2014; Fetter et al.,
+2019). The second prediction about size-density scaling is more complex
+because we would need to know the relationships between colonization,
+stomatal cover, stomatal conductance, and fitness in natural conditions.
+There is growing evidence that stomata mediate tradeoffs between
+photosynthesis and defense in \emph{Populus trichocarpa} (McKown et al.,
+2019), but testing these predictions in a variety of species will help
+determine whether pathogens have played an important role shaping
+stomatal anatomy in land plants.
+
+\hypertarget{acknowledgments}{%
+\section*{Acknowledgments}\label{acknowledgments}}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+I would like to thank Athena McKown and an anonymous referee for
+valuable feedback that improved this manuscript.
 
@@
 
+\clearpage
+
 \hypertarget{supplementary-material}{%
@@
 \addcontentsline{toc}{section}{Supplementary Material}
+
+\renewcommand{\theequation}{S\arabic{equation}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
+\setcounter{equation}{0}    
+\setcounter{figure}{0}    
+\setcounter{table}{0}
+
+\hypertarget{g_textsmax-calculation}{%
+\subsection*{\texorpdfstring{\(g_\text{s,max}\)
+calculation}{g\_\textbackslash text\{s,max\} calculation}}\label{g_textsmax-calculation}}
+\addcontentsline{toc}{subsection}{\(g_\text{s,max}\) calculation}
 
@@
 
+\hypertarget{is-proportional-to-the-stomatal-pore-area-index}{%
+\subsection*{\texorpdfstring{\fs{} is proportional to the stomatal pore
+area
+index}{ is proportional to the stomatal pore area index}}\label{is-proportional-to-the-stomatal-pore-area-index}}
+\addcontentsline{toc}{subsection}{\fs{} is proportional to the stomatal
+pore area index}
+
+The stomatal pore area index (SPI; Sack et al., 2003) is calculated as
+the product of the stomatal density and guard cell length (GCL) squared:
+
+\begin{equation*}
+  \mathrm{SPI} = D \times \mathrm{GCL} ^ 2
+\end{equation*}
+
+Assuming that the stomatal radius \(R\) is half the GCL, then stomatal
+size \(S\) is equivalent to:
+
+\begin{equation*}
+  S = \pi R ^ 2 = \frac{\pi \times \mathrm{GCL} ^ 2}{4}
+\end{equation*}
+
+Based on equation \ref{eq:fs}, it follows that \fs{} and SPI are
+proportional:
+
+\begin{equation*}
+  f_\text{S} = \frac{\pi \times \mathrm{SPI}}{4}
+\end{equation*}
+
+\clearpage
+
+\hypertarget{figures}{%
+\subsection*{Figures}\label{figures}}
+\addcontentsline{toc}{subsection}{Figures}
+
+\begin{figure}
+  \centering
+    \includegraphics{../figures/figS1.pdf}
+    \caption{\textbf{The probability of colonization increases with both stomatal size ($S$) and density $D$} I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of $S$, $D$, and $H$ (see [Materials and Methods]) \textbf{a.} Each line shows how $p_\text{colonize}$ increases with $S$ ($x$-axis, log-scale) for selected values of $D \in \{10, 100, 1000\}~\text{mm}^{-2}$.
+\textbf{b.} Each line shows how $p_\text{colonize}$ increases with $D$ ($x$-axis, log-scale) for selected values of $S \in \{10, 100, 1000\}~\mu\text{m}^2$.
+The facets show results for different values of $H$}
+    \label{fig:figS1}
+\end{figure}
+
+\begin{figure}
+  \centering
+    \includegraphics{../figures/figS2.pdf}
+    \caption{\textbf{The probability of colonization increases with both stomatal cover and conductance in the spatially implicit model.} As in Figure \ref{fig:fig3}, I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of stomatal densities and sizes (see [Materials and Methods]), but a subset of results are shown here.
+Stomatal size and density determine stomatal cover (\fs; Equation \ref{eq:fs}) and theoretical maximum stomatal conductance (\gsmax; Equation \ref{eq:gsmax}).
+\textbf{a.} $p_\text{colonize}$ initially increases rapidly with \fs{} ($x$-axis), then slows down to a linear relationship.
+Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$).
+The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density for all values of $H$, which is why the lines overlap.
+\textbf{b.} $p_\text{colonize}$ increases sigmoidally with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is lower at higher densities for a given \gsmax.
+The relationship between \gsmax{} and $p_\text{colonize}$ is similar for all values of $H > 0$.}
+    \label{fig:figS2}
+\end{figure}
+
+\begin{figure}
+  \centering
+    \includegraphics{../figures/figS3.pdf}
+    \caption{\textbf{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$) in the spatially implicit model}.
+As in Figure \ref{fig:fig4}, in each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.25 (lowest line), 0.5, or 0.75 (highest line).
+For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.5$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region.
+The scaling exponent is unity $\beta = 1$ when $H > 0$.}
+    \label{fig:figS3}
+\end{figure}
+
+\clearpage
+
+\hypertarget{appendix-1-spatially-implicit-model}{%
+\section*{Appendix 1: Spatially implicit
+model}\label{appendix-1-spatially-implicit-model}}
+\addcontentsline{toc}{section}{Appendix 1: Spatially implicit model}
+
+A limitation of the spatially explicit model is that a pathogen could
+only infect stomata in the focal triangle where it landed. Here I
+analyze an alternative, spatially implicit, model that relaxes this
+assumptions. Instead, I assume that a pathogen can potentially infect
+any stomate on the leaf. It searches through a random walk and has a
+continuous, constant probability of encountering a stomate that is
+determined by stomatal cover (\fs). If \(fs << 1\), this can be modeled
+as a homogeneous Poisson process and the distance \(x\) a pathogen must
+travel before reaching a stomate follows an exponential distribution:
+
+\begin{equation*}
+  f(x) = f_\text{S} e ^ {-f_\text{S} x}
+\end{equation*}
+
+Given the constant death rate per unit distance \(H\), the probability
+of surviving to distance \(x\) is \(e^{-H x}\). The probability of
+locating a stomate the probability of surviving to distance \(x\)
+multiplied by \(f(x)\) and integrated over all \(x\) from 0 to
+\(\infty\):
+
+\begin{align*}
+  p_\text{locate} & = \int_0^\infty f(x) e^{-Hx} dx \\
+                  & = \int_0^\infty f_\text{S} e^{-(f_\text{S} + H)x} dx \\
+                  & = \frac{f_\text{S}}{H + f_\text{S}}
+\end{align*}
+
+Substituting \(p_\text{locate}\) above into Equation
+\ref{eq:p_colonize}:
+
+\begin{align*}
+  p_\text{colonize} & = f_\text{S} + (1 - f_\text{S}) p_\text{locate} \\
+                    & = f_\text{S} + (1 - f_\text{S}) \frac{f_\text{S}}{H + f_\text{S}} \\
+                    & = f_\text{S} (1 + \frac{1 - f_\text{S}}{H + f_\text{S}})
+\end{align*}
+
+With the spatially implicit model, because pathogens can potentially
+reach any stomate on the leaf, \(p_\text{colonize}\) is greater than
+that in the spatially explicit model for the same value of \(H\). For
+example, if the pathogen can search forever (\(H = 0\)), then it will
+always colonize (\(p_\text{colonize} = 1\); Figure \ref{fig:figS2}a).
+But even when \(H > 0\), \(p_\text{colonize}\) is significantly higher
+than in the spatially implicit than spatially explicit model for the
+same \fs{} because pathogens can potentially colonize any stomate on the
+leaf.
+
+Whereas the spatially explicit model probably underestimates
+\(p_\text{locate}\) for pathogens that can search over long distances,
+the implicit model overestimates because it assumes that the probability
+of encountering a stomate is constant (i.e.~homogeneous Poisson
+process). This is not true because stomata are discrete areas on the
+leaf. If a pathogen is searching far away from a stomate, its
+probability of encountering a stomate in the near future is lower than
+that for a pathogen searching near a stomate. This should be modeled as
+a \emph{nonhomogenous} Poisson process. Future work should derive
+\(p_\text{locate}\) for the stomatal anatomies presented here under a
+nonhomogeneous process.
+
+Despite the quantitative differences in the the spatially explicit and
+implicit models, they have similar qualitative properties when
+\(H > 0\), which is reasonable since the leaf surface is a relatively
+hostile environment for most pathogens (see
+\protect\hyperlink{introduction}{Introduction}). In both models,
+\(p_\text{colonize}\) increases with, but is higher than \fs. In the
+spatially explicit model, size-density scaling that preserves
+\(p_\text{colonize}\) is 1 when \(H = 0\) and slightly less than 1
+otherwise (Figure \ref{fig:fig4}). In the spatially implicit model, the
+scaling coefficient is always 1. Rearranging the equation for
+\(p_\text{colonize}\) above and substituting \(f_\text{S} = DS\), the
+following relationship holds:
+
+\[SD = \frac{p_\text{colonize} H}{1 - p_\text{colonize} + H}\] Since
+\(H\) is a constant the right-hand side of the equation above is
+constant for a given value of \(p_\text{colonize}\). Hence the
+\(\beta_p\) that would preserve the relationship above is simply 1
+(Figure \ref{fig:figS3}).
+
+\clearpage
+
 \hypertarget{references}{%
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-allaire_rmarkdown:_2019}{}%
+\leavevmode\hypertarget{ref-allaire_rmarkdown:_2020}{}%
 Allaire, J., Xie, Y., McPherson, J., Luraschi, J., Ushey, K., Wickham,
-H., Cheng, J., Chang, W., and Iannone, R. (2019). Rmarkdown: Dynamic
+H., Cheng, J., Chang, W., and Iannone, R. (2020). Rmarkdown: Dynamic
 Documents for R. Available at:
 \url{https://github.com/rstudio/rmarkdown}.
+
+\leavevmode\hypertarget{ref-allen_appressorium_1991}{}%
+Allen, E. A., Hazen, B. E., Hoch, H. C., Kwon, Y., Leinhos, G. M. E.,
+Staples, R. C., Stumpf, M. A., and Terhune, B. T. (1991). Appressorium
+Formation in Response to Topographical Signals by 27 Rust Species.
+\emph{Phytopathology} 81, 323.
+doi:\href{https://doi.org/10.1094/Phyto-81-323}{10.1094/Phyto-81-323}.
 
@@
 
+\leavevmode\hypertarget{ref-bastiaans_ratio_1991}{}%
+Bastiaans, L. (1991). Ratio Between Virtual and Visual Lesion Size as a
+Measure to Describe Reduction in Leaf Photosynthesis of Rice Due to Leaf
+Blast. \emph{Phytopathology} 81, 611.
+doi:\href{https://doi.org/10.1094/Phyto-81-611}{10.1094/Phyto-81-611}.
+
+\leavevmode\hypertarget{ref-beattie_secret_1995}{}%
+Beattie, G. A., and Lindow, S. E. (1995). The secret life of foliar
+bacterial pathogens on leaves. \emph{Annual Review of Phytopathology}
+33, 145--172.
+
 \leavevmode\hypertarget{ref-berry_stomata:_2010}{}%
@@
 
+\leavevmode\hypertarget{ref-de_boer_optimal_2016}{}%
+Boer, H. J. de, Price, C. A., Wagner-Cremer, F., Dekker, S. C., Franks,
+P. J., and Veneklaas, E. J. (2016). Optimal allocation of leaf epidermal
+area for gas exchange. \emph{New Phytologist} 210, 1219--1228.
+doi:\href{https://doi.org/10.1111/nph.13929}{10.1111/nph.13929}.
+
 \leavevmode\hypertarget{ref-borchers_pracma:_2019}{}%
 Borchers, H. W. (2019). Pracma: Practical Numerical Math Functions. R
-package version 2.2.5. Available at:
+package version 2.2.9. Available at:
 \url{https://CRAN.R-project.org/package=pracma}.
+
+\leavevmode\hypertarget{ref-perez-martin_tropic_2012}{}%
+Brand, A., and Gow, N. A. R. (2012). ``Tropic Orientation Responses of
+Pathogenic Fungi,'' in \emph{Morphogenesis and Pathogenicity in Fungi},
+eds. J. Pérez-Martín and A. Di Pietro (Berlin, Heidelberg: Springer
+Berlin Heidelberg), 21--41.
+doi:\href{https://doi.org/10.1007/978-3-642-22916-9_2}{10.1007/978-3-642-22916-9\_2}.
 
@@
 doi:\href{https://doi.org/10.1104/pp.17.00183}{10.1104/pp.17.00183}.
-
-\leavevmode\hypertarget{ref-de_boer_optimal_2016}{}%
-de Boer, H. J., Price, C. A., Wagner-Cremer, F., Dekker, S. C., Franks,
-P. J., and Veneklaas, E. J. (2016). Optimal allocation of leaf epidermal
-area for gas exchange. \emph{New Phytologist} 210, 1219--1228.
-doi:\href{https://doi.org/10.1111/nph.13929}{10.1111/nph.13929}.
 
@@
 
+\leavevmode\hypertarget{ref-drake_smaller_2013}{}%
+Drake, P. L., Froend, R. H., and Franks, P. J. (2013). Smaller, faster
+stomata: Scaling of stomatal size, rate of response, and stomatal
+conductance. \emph{Journal of Experimental Botany} 64, 495--505.
+doi:\href{https://doi.org/10.1093/jxb/ers347}{10.1093/jxb/ers347}.
+
 \leavevmode\hypertarget{ref-dutton_bacterial_2019}{}%
@@
 selection conflicts in hybrid poplars indicate the stomatal ratio as an
-important trait regulating disease resistance.
+important trait regulating disease resistance. University of Vermont
 doi:\href{https://doi.org/10.1101/814046}{10.1101/814046}.
+
+\leavevmode\hypertarget{ref-flexas_co2_2018}{}%
+Flexas, J., Cano, F. J., Carriquí, M., Coopman, R. E., Mizokami, Y.,
+Tholen, D., and Xiong, D. (2018). ``CO2 Diffusion Inside Photosynthetic
+Organs,'' in \emph{The Leaf: A Platform for Performing Photosynthesis}
+Advances in Photosynthesis and Respiration., eds. W. W. Adams III and I.
+Terashima (Cham: Springer International Publishing), 163--208.
+doi:\href{https://doi.org/10.1007/978-3-319-93594-2_7}{10.1007/978-3-319-93594-2\_7}.
 
@@
 
+\leavevmode\hypertarget{ref-gilbert_evolutionary_2002}{}%
+Gilbert, G. S. (2002). Evolutionary Ecology of Plant Diseases in Natural
+Ecosystems. \emph{Annual Review of Phytopathology} 40, 13--43.
+doi:\href{https://doi.org/10.1146/annurev.phyto.40.021202.110417}{10.1146/annurev.phyto.40.021202.110417}.
+
+\leavevmode\hypertarget{ref-gray_flanking_2020}{}%
+Gray, A., Liu, L., and Facette, M. (2020). Flanking Support: How
+Subsidiary Cells Contribute to Stomatal Form and Function.
+\emph{Frontiers in Plant Science} 11.
+doi:\href{https://doi.org/10.3389/fpls.2020.00881}{10.3389/fpls.2020.00881}.
+
 \leavevmode\hypertarget{ref-gutschick_photosynthesis_1984}{}%
@@
 doi:\href{https://doi.org/10.1007/s00425-002-0760-2}{10.1007/s00425-002-0760-2}.
+
+\leavevmode\hypertarget{ref-lawson_stomatal_2014}{}%
+Lawson, T., and Blatt, M. R. (2014). Stomatal Size, Speed, and
+Responsiveness Impact on Photosynthesis and Water Use Efficiency.
+\emph{Plant Physiology} 164, 1556--1570.
+doi:\href{https://doi.org/10.1104/pp.114.237107}{10.1104/pp.114.237107}.
+
+\leavevmode\hypertarget{ref-lawson_coordination_2018}{}%
+Lawson, T., Terashima, I., Fujita, T., and Wang, Y. (2018).
+``Coordination Between Photosynthesis and Stomatal Behavior,'' in
+\emph{The Leaf: A Platform for Performing Photosynthesis} Advances in
+Photosynthesis and Respiration., eds. W. W. Adams III and I. Terashima
+(Cham: Springer International Publishing), 141--161.
+doi:\href{https://doi.org/10.1007/978-3-319-93594-2_6}{10.1007/978-3-319-93594-2\_6}.
 
@@
 
+\leavevmode\hypertarget{ref-mckown_role_2019}{}%
+McKown, A. D., Klápště, J., Guy, R. D., Corea, O. R. A., Fritsche, S.,
+Ehlting, J., El‐Kassaby, Y. A., and Mansfield, S. D. (2019). A role for
+\emph{speechless} in the integration of leaf stomatal patterning with
+the growth vs disease trade‐off in poplar. \emph{New Phytologist},
+nph.15911.
+doi:\href{https://doi.org/10.1111/nph.15911}{10.1111/nph.15911}.
+
 \leavevmode\hypertarget{ref-mclachlan_gate_2014}{}%
@@
 
+\leavevmode\hypertarget{ref-mitchell_trophic_2003}{}%
+Mitchell, C. E. (2003). Trophic control of grassland production and
+biomass by pathogens. \emph{Ecology Letters} 6, 147--155.
+doi:\href{https://doi.org/10.1046/j.1461-0248.2003.00408.x}{10.1046/j.1461-0248.2003.00408.x}.
+
 \leavevmode\hypertarget{ref-monteith_principles_2013}{}%
@@
 Morison, J. I. L., Emily Gallouët, Lawson, T., Cornic, G., Herbin, R.,
-and work(s): N. R. B. R. (2005). Lateral Diffusion of CO₂ in Leaves Is
-Not Sufficient to Support Photosynthesis. \emph{Plant Physiology} 139,
-254--266. Available at: \url{http://www.jstor.org/stable/4281859}.
+and work(s): N. R. B. R. (2005). Lateral Diffusion of CO\$\_2\$ in
+Leaves Is Not Sufficient to Support Photosynthesis. \emph{Plant
+Physiology} 139, 254--266. Available at:
+\url{http://www.jstor.org/stable/4281859}.
 
@@
 
+\leavevmode\hypertarget{ref-oguchi_leaf_2018}{}%
+Oguchi, R., Onoda, Y., Terashima, I., and Tholen, D. (2018). ``Leaf
+Anatomy and Function,'' in \emph{The Leaf: A Platform for Performing
+Photosynthesis} Advances in Photosynthesis and Respiration., eds. W. W.
+Adams III and I. Terashima (Cham: Springer International Publishing),
+97--139.
+doi:\href{https://doi.org/10.1007/978-3-319-93594-2_5}{10.1007/978-3-319-93594-2\_5}.
+
+\leavevmode\hypertarget{ref-papanatsiou_stomatal_2017}{}%
+Papanatsiou, M., Amtmann, A., and Blatt, M. R. (2017). Stomatal
+clustering in Begonia associates with the kinetics of leaf gaseous
+exchange and influences water use efficiency. \emph{Journal of
+Experimental Botany} 68, 2309--2315.
+doi:\href{https://doi.org/10.1093/jxb/erx072}{10.1093/jxb/erx072}.
+
 \leavevmode\hypertarget{ref-parkhurst_diffusion_1994}{}%
@@
 
-\leavevmode\hypertarget{ref-r_core_team_r:_2019}{}%
-R Core Team (2019). \emph{R: A Language and Environment for Statistical
+\leavevmode\hypertarget{ref-raissig_mobile_2017}{}%
+Raissig, M. T., Matos, J. L., Anleu Gil, M. X., Kornfeld, A.,
+Bettadapur, A., Abrash, E., Allison, H. R., Badgley, G., Vogel, J. P.,
+Berry, J. A., et al. (2017). Mobile MUTE specifies subsidiary cells to
+build physiologically improved grass stomata. \emph{Science} 355,
+1215--1218.
+doi:\href{https://doi.org/10.1126/science.aal3254}{10.1126/science.aal3254}.
+
+\leavevmode\hypertarget{ref-r_core_team_r:_2020}{}%
+R Core Team (2020). \emph{R: A Language and Environment for Statistical
 Computing}. Vienna, Austria: R Foundation for Statistical Computing
@@
 Sack, L., and Buckley, T. N. (2016). The developmental basis of stomatal
-density and flux. \emph{Plant Physiology}, pp.00476.2016.
+density and flux. \emph{Plant Physiology} 171, 2358--2363.
 doi:\href{https://doi.org/10.1104/pp.16.00476}{10.1104/pp.16.00476}.
@@
 
+\leavevmode\hypertarget{ref-tucker_surface_2001}{}%
+Tucker, S. L., and Talbot, N. J. (2001). Surface attachment and
+pre-penetration stage development by plant pathogenic fungi.
+\emph{Annual Review of Phytopathology} 39, 385--417.
+doi:\href{https://doi.org/10.1146/annurev.phyto.39.1.385}{10.1146/annurev.phyto.39.1.385}.
+
 \leavevmode\hypertarget{ref-underwood_role_2007}{}%
@@
 doi:\href{https://doi.org/10.1111/j.1462-5822.2007.00938.x}{10.1111/j.1462-5822.2007.00938.x}.
+
+\leavevmode\hypertarget{ref-ushey_reticulate_2020}{}%
+Ushey, K., Allaire, J. J., and Tang, Y. (2020). Reticulate: Interface to
+'Python'. Available at:
+\url{https://CRAN.R-project.org/package=reticulate}.
 

### 2020-07-13 r/00_run-python.R

@@ -24,3 +24,9 @@ get_tangent_points <-
      }}", x1 = x1, x2 = x2, y1 = y1, y2 = y2) 
 
 cat(get_tangent_slopes, "\n", get_tangent_points, file = "r/functions-from-python.R")
+
+
+library(glue)
+library(reticulate)
+# use_python('/anaconda3/bin/python', required = TRUE)
+source_python("python/spatially-implicit-model.py")

### 2020-07-13 r/02_spatially-explicit-model.R

@@ -2,7 +2,7 @@ source("r/header.R")
 
 df <- crossing(
   H = c(0, 0.01, 0.1),
-  D = set_units(10 ^ seq(1, 3.5, by = 0.025), 1 / mm ^ 2),
+  D = set_units(sort(c(25, 500, 10 ^ seq(1, 3.5, by = 0.025))), 1 / mm ^ 2),
   S = set_units(10 ^ seq(1, 3.5, by = 0.025), um ^ 2)
 ) %>%
   mutate(f_s = D * S) %>%
@@ -16,15 +16,12 @@ df %<>%
   mutate(
     b = biophysical_constant(D_wv, v),
     m = morphological_constant(0.5, 0.5, 0.5),
-    # CHeck if sqrt(s) is right here
     g_smax = set_units(b * m * D * sqrt(S), mol / m ^ 2 / s)
   )
 
-# p_colonize(df$D[1], df$S[1], df$H[1])
-
 safe_pcolonize <- safely(p_colonize)
-tictoc::tic()
-plan(multiprocess)
+
+plan(multisession)
 df$p_colonize <- df %>% 
   select(D, S, H) %>%
   furrr::future_pmap_dbl(~ {
@@ -35,7 +32,6 @@ df$p_colonize <- df %>%
       return(res$result)
     }
   }, .progress = TRUE)
-tictoc::toc()
 
 df %<>% 
   filter(!is.na(p_colonize)) %>%

### 2020-07-13 r/06_plot-spatially-implicit.R

@@ -1,0 +1,58 @@
+source("r/header.R")
+
+df1 <- read_rds("objects/spatially-explicit-results.rds") %>%
+  mutate(p_colonize = f_s * (1 + (1 - f_s) / (H + f_s))) %>%
+  filter(D %in% c(100, 500)) %>%
+  mutate(H = glue("H = {H}", H = H), D = factor(round(D), levels = c("100", "500")))
+
+
+fS2a <- ggplot(df1, aes(x = f_s, y = p_colonize, linetype = D)) +
+  facet_grid(H ~ .) +
+  geom_line(size = 1.1, lineend = "round") +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+  xlab(expression(paste("Stomatal cover (", italic(f)[s], ")"))) +
+  ylab(expression(paste("Probability of colonization (", italic(p)[colonize], ")"))) +
+  scale_linetype_manual(
+    values = c(2, 1), 
+    name = expression(paste("Stomatal density [m", m^-2, "]"))
+  ) +
+  theme_bw() +
+  theme(
+    axis.text = element_text(size = 10),
+    axis.title = element_text(size = 12),
+    axis.title.x = element_text(margin = margin(0.3, 0, 0, 0, unit = "cm")),
+    legend.key.size = unit(0.75, "in"),
+    legend.position = "top",
+    strip.background = element_rect(fill = "white"),
+    strip.text = element_text(size = 10)
+  )
+
+l <- get_legend(fS2a)
+
+fS2a <- fS2a + theme(legend.position = "none")
+
+df2 <- df1 %>%
+  filter(g_smax < 5)
+
+fS2b <- ggplot(df2, aes(x = g_smax, y = p_colonize, linetype = D)) +
+  facet_wrap(H ~ ., strip.position = "left", ncol = 1) +
+  geom_line(size = 1.1, lineend = "round") +
+  xlab(expression(paste("Stomatal conductance (", italic(g)[list(s,max)], ") [mol ", m^-2~s^-1, "]"))) +
+  ylab(expression(paste("Probability of colonization (", italic(p)[colonize], ")"))) +
+  scale_linetype_manual(values = c(2, 1)) +
+  scale_y_continuous(position = "right") +
+  theme_bw() +
+  theme(
+    axis.text = element_text(size = 10),
+    axis.title = element_text(size = 12),
+    legend.position = "none",
[…]

### 2020-07-13 r/07_make-appendix2.R

@@ -1,0 +1,23 @@
+source("r/header.R")
+
+library(knitcitations)
+library(RefManageR)
+library(sessioninfo)
+si <- session_info(include_base = FALSE)
+
+r_package_table <- map2_dfr(si$packages$package, si$packages$loadedversion, ~ {
+  data.frame(
+    key = as.character(as.BibEntry(citation(.x))$key)
+  ) %>%
+    mutate(
+      package = c(.x, rep("", nrow(.) - 1L)),
+      version = c(.y, rep("", nrow(.) - 1L)),
+      key = paste0("@", key, "")
+    )
+})
+
+file.create("ms/appendix.bib")
+si$packages$package %>%
+  walk(~ {
+    write.bibtex(citation(.x), file = "ms/appendix.bib", append = TRUE)
+  })

### 2020-07-13 r/header.R

@@ -23,4 +23,3 @@ source("r/functions.R")
 source("r/functions-from-python.R")
 
 palette(colorRampPalette(c("tomato", "steelblue"), alpha = TRUE)(5))
-

### 2020-10-18 ms/appendix2.tex
pre-release

@@
 
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {\setlength{\parindent}{0pt}%
+  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
+  {\par}
+
 \usepackage{booktabs}
@@
 
-  \textsuperscript{1} School of Life Sciences, University of Hawaii,  Honolulu,  Hawaii,  USA
+  \textsuperscript{1} School of Life Sciences, University of
+Hawaii,  Honolulu,  Hawaii,  USA
   }
 
-  \def\corrAuthor{Christopher D. Muir}\def\corrAddress{,  }\def\corrEmail{\href{mailto:cdmuir@hawaii.edu}{\nolinkurl{cdmuir@hawaii.edu}}}
+  \def\corrAuthor{Christopher D.
+Muir}\def\corrAddress{,  }\def\corrEmail{\href{mailto:cdmuir@hawaii.edu}{\nolinkurl{cdmuir@hawaii.edu}}}
   \def\firstAuthorLast{Muir {et~al.}}
@@
 assertthat & 0.2.1 & Wickham (2019a)\tabularnewline
-bibtex & 0.4.2.2 & Francois (2020)\tabularnewline
+bibtex & 0.4.2.3 & Francois (2020)\tabularnewline
 BiocManager & 1.30.10 & Morgan (2019)\tabularnewline
-cli & 2.0.2 & Cs\a'ardi (2020)\tabularnewline
+cli & 2.1.0 & Cs\a'ardi (2020)\tabularnewline
 codetools & 0.2-16 & Tierney (2018)\tabularnewline
@@
 & & Stauffer et al. (2009)\tabularnewline
-cowplot & 1.0.0 & Wilke (2019)\tabularnewline
+cowplot & 1.1.0 & Wilke (2020)\tabularnewline
 crayon & 1.3.4 & Cs\a'ardi (2017)\tabularnewline
 digest & 0.6.25 & Antoine Lucas et al. (2020)\tabularnewline
-dplyr & 1.0.0 & Wickham et al. (2020a)\tabularnewline
-ellipsis & 0.3.1 & Wickham (2020)\tabularnewline
+dplyr & 1.0.2 & Wickham et al. (2020a)\tabularnewline
+ellipsis & 0.3.1 & Wickham (2020a)\tabularnewline
 evaluate & 0.14 & Wickham and Xie (2019)\tabularnewline
@@
 filehash & 2.4-2 & Peng (2006)\tabularnewline
-furrr & 0.1.0 & Vaughan and Dancho (2018)\tabularnewline
-future & 1.18.0 & Bengtsson (2020)\tabularnewline
+furrr & 0.2.0 & Vaughan and Dancho (2020)\tabularnewline
+future & 1.19.1 & Bengtsson (2020a)\tabularnewline
 generics & 0.0.2 & Kuhn et al. (2018)\tabularnewline
@@
 ggplotify & 0.0.5 & Yu (2020b)\tabularnewline
-globals & 0.12.5 & Bengtsson (2019a)\tabularnewline
-glue & 1.4.1 & Hester (2020)\tabularnewline
+globals & 0.13.1 & Bengtsson (2020b)\tabularnewline
+glue & 1.4.2 & Hester (2020)\tabularnewline
 gridGraphics & 0.5-0 & Murrell and Wen (2020)\tabularnewline
@@
 htmltools & 0.5.0 & Cheng et al. (2020)\tabularnewline
-httr & 1.4.1 & Wickham (2019b)\tabularnewline
-jsonlite & 1.7.0 & Ooms (2014)\tabularnewline
+httr & 1.4.2 & Wickham (2020b)\tabularnewline
+jsonlite & 1.7.1 & Ooms (2014)\tabularnewline
 knitcitations & 1.0.10 & Boettiger (2019)\tabularnewline
-knitr & 1.29 & Xie (2020a)\tabularnewline
+knitr & 1.30 & Xie (2020a)\tabularnewline
 & & Xie (2015)\tabularnewline
@@
 lifecycle & 0.2.0 & Henry (2020)\tabularnewline
-listenv & 0.8.0 & Bengtsson (2019b)\tabularnewline
+listenv & 0.8.0 & Bengtsson (2019)\tabularnewline
 lubridate & 1.7.9 & Grolemund and Wickham (2011)\tabularnewline
@@
 magrittr & 1.5 & Bache and Wickham (2014)\tabularnewline
-MASS & 7.3-51.6 & Venables and Ripley (2002)\tabularnewline
+MASS & 7.3-53 & Venables and Ripley (2002)\tabularnewline
 munsell & 0.5.0 & Wickham (2018)\tabularnewline
@@
 & & Eddelbuettel and Balamuta (2017)\tabularnewline
-readr & 1.3.1 & Wickham et al. (2018)\tabularnewline
+readr & 1.4.0 & Wickham and Hester (2020)\tabularnewline
 RefManageR & 1.2.12 & McLean (2017)\tabularnewline
 & & McLean (2014)\tabularnewline
-rlang & 0.4.7 & Henry and Wickham (2020b)\tabularnewline
-rmarkdown & 2.3 & Allaire et al. (2020a)\tabularnewline
+rlang & 0.4.8 & Henry and Wickham (2020b)\tabularnewline
+rmarkdown & 2.4 & Allaire et al. (2020a)\tabularnewline
 & & Xie et al. (2018)\tabularnewline
-rticles & 0.14 & Allaire et al. (2020b)\tabularnewline
+& & Xie et al. (2020)\tabularnewline
+rticles & 0.16 & Allaire et al. (2020b)\tabularnewline
 rvcheck & 0.1.8 & Yu (2020c)\tabularnewline
@@
 sessioninfo & 1.1.1 & Cs\a'ardi et al. (2018)\tabularnewline
-stringi & 1.4.6 & Gagolewski (2020)\tabularnewline
-stringr & 1.4.0 & Wickham (2019c)\tabularnewline
-tibble & 3.0.3 & Müller and Wickham (2020b)\tabularnewline
-tidyr & 1.1.0 & Wickham and Henry (2020)\tabularnewline
+stringi & 1.5.3 & Gagolewski (2020)\tabularnewline
+stringr & 1.4.0 & Wickham (2019b)\tabularnewline
+tibble & 3.0.4 & Müller and Wickham (2020b)\tabularnewline
+tidyr & 1.1.2 & Wickham (2020c)\tabularnewline
 tidyselect & 1.1.0 & Henry and Wickham (2020c)\tabularnewline
 tikzDevice & 0.12.3.1 & Sharpsteen and Bracken (2020)\tabularnewline
+tinytex & 0.26 & Xie (2020b)\tabularnewline
+& & Xie (2019)\tabularnewline
 tweenr & 1.0.1 & Pedersen (2018)\tabularnewline
 units & 0.6-7 & Pebesma et al. (2016)\tabularnewline
-vctrs & 0.3.1 & Wickham et al. (2020b)\tabularnewline
-withr & 2.2.0 & Hester et al. (2020)\tabularnewline
-xfun & 0.15 & Xie (2020b)\tabularnewline
+vctrs & 0.3.4 & Wickham et al. (2020b)\tabularnewline
+withr & 2.3.0 & Hester et al. (2020)\tabularnewline
+xfun & 0.18 & Xie (2020c)\tabularnewline
 xml2 & 1.3.2 & Wickham et al. (2020c)\tabularnewline
@@
 \hypertarget{refs}{}
+\begin{cslreferences}
 \leavevmode\hypertarget{ref-allaire2020rmarkdown}{}%
@@
 
-\leavevmode\hypertarget{ref-bengtsson2020future}{}%
-Bengtsson, H. (2020). \emph{Future: Unified parallel and distributed
-processing in r for everyone}. Available at:
-\url{https://CRAN.R-project.org/package=future}.
+\leavevmode\hypertarget{ref-bengtsson2020unifying}{}%
+Bengtsson, H. (2020a). A unifying framework for parallel and distributed
+processing in r using futures. Available at:
+\url{https://arxiv.org/abs/2008.00553}.
 
-\leavevmode\hypertarget{ref-bengtsson2019globals}{}%
-Bengtsson, H. (2019a). \emph{Globals: Identify global objects in r
+\leavevmode\hypertarget{ref-bengtsson2020globals}{}%
+Bengtsson, H. (2020b). \emph{Globals: Identify global objects in r
 expressions}. Available at:
@@
 \leavevmode\hypertarget{ref-bengtsson2019listenv}{}%
-Bengtsson, H. (2019b). \emph{Listenv: Environments behaving (almost) as
+Bengtsson, H. (2019). \emph{Listenv: Environments behaving (almost) as
 lists}. Available at: \url{https://CRAN.R-project.org/package=listenv}.
@@
 
-\leavevmode\hypertarget{ref-vaughan2018furrr}{}%
-Vaughan, D., and Dancho, M. (2018). \emph{Furrr: Apply mapping functions
+\leavevmode\hypertarget{ref-vaughan2020furrr}{}%
+Vaughan, D., and Dancho, M. (2020). \emph{Furrr: Apply mapping functions
 in parallel using futures}. Available at:
@@
 statistics with s}. Fourth. New York: Springer Available at:
-\url{http://www.stats.ox.ac.uk/pub/MASS4}.
+\url{http://www.stats.ox.ac.uk/pub/MASS4/}.
 
@@
 \leavevmode\hypertarget{ref-wickham2020ellipsis}{}%
-Wickham, H. (2020). \emph{Ellipsis: Tools for working with ...}
+Wickham, H. (2020a). \emph{Ellipsis: Tools for working with ...}
 Available at: \url{https://CRAN.R-project.org/package=ellipsis}.
@@
 
-\leavevmode\hypertarget{ref-wickham2019httr}{}%
-Wickham, H. (2019b). \emph{Httr: Tools for working with urls and http}.
+\leavevmode\hypertarget{ref-wickham2020httr}{}%
+Wickham, H. (2020b). \emph{Httr: Tools for working with urls and http}.
 Available at: \url{https://CRAN.R-project.org/package=httr}.
@@
 \leavevmode\hypertarget{ref-wickham2019stringr}{}%
-Wickham, H. (2019c). \emph{Stringr: Simple, consistent wrappers for
+Wickham, H. (2019b). \emph{Stringr: Simple, consistent wrappers for
 common string operations}. Available at:
@@
 
+\leavevmode\hypertarget{ref-wickham2020tidyr}{}%
+Wickham, H. (2020c). \emph{Tidyr: Tidy messy data}. Available at:
+\url{https://CRAN.R-project.org/package=tidyr}.
+
 \leavevmode\hypertarget{ref-wickham2020dplyr}{}%
-Wickham, H., François, R., Henry, L., and Müller, K. (2020a).
+Wickham, H., François, R., ? Henry, L., and Müller, K. (2020a).
 \emph{Dplyr: A grammar of data manipulation}. Available at:
 \url{https://CRAN.R-project.org/package=dplyr}.
-
-\leavevmode\hypertarget{ref-wickham2020tidyr}{}%
-Wickham, H., and Henry, L. (2020). \emph{Tidyr: Tidy messy data}.
-Available at: \url{https://CRAN.R-project.org/package=tidyr}.
 
@@
 
-\leavevmode\hypertarget{ref-wickham2018readr}{}%
-Wickham, H., Hester, J., and Francois, R. (2018). \emph{Readr: Read
-rectangular text data}. Available at:
-\url{https://CRAN.R-project.org/package=readr}.
+\leavevmode\hypertarget{ref-wickham2020readr}{}%
+Wickham, H., and Hester, J. (2020). \emph{Readr: Read rectangular text
+data}. Available at: \url{https://CRAN.R-project.org/package=readr}.
 
@@
 
-\leavevmode\hypertarget{ref-wilke2019cowplot}{}%
-Wilke, C. O. (2019). \emph{Cowplot: Streamlined plot theme and plot
+\leavevmode\hypertarget{ref-wilke2020cowplot}{}%
+Wilke, C. O. (2020). \emph{Cowplot: Streamlined plot theme and plot
 annotations for 'ggplot2'}. Available at:
@@
 
+\leavevmode\hypertarget{ref-xie2019tinytex}{}%
+Xie, Y. (2019). TinyTeX: A lightweight, cross-platform, and
+easy-to-maintain latex distribution based on tex live. \emph{TUGboat},
+30--32. Available at:
+\url{http://tug.org/TUGboat/Contents/contents40-1.html}.
+
+\leavevmode\hypertarget{ref-xie2020tinytex}{}%
+Xie, Y. (2020b). \emph{Tinytex: Helper functions to install and maintain
+tex live, and compile latex documents}. Available at:
+\url{https://github.com/yihui/tinytex}.
+
 \leavevmode\hypertarget{ref-xie2020xfun}{}%
-Xie, Y. (2020b). \emph{Xfun: Miscellaneous functions by 'yihui xie'}.
+Xie, Y. (2020c). \emph{Xfun: Miscellaneous functions by 'yihui xie'}.
 Available at: \url{https://CRAN.R-project.org/package=xfun}.
@@
 \url{https://bookdown.org/yihui/rmarkdown}.
+
+\leavevmode\hypertarget{ref-xie2020markdown}{}%
+Xie, Y., Dervieux, C., and Riederer, E. (2020). \emph{R markdown
+cookbook}. Boca Raton, Florida: Chapman; Hall/CRC Available at:
+\url{https://bookdown.org/yihui/rmarkdown-cookbook}.
 
@@
 doi:\href{https://doi.org/10.1016/j.csda.2008.11.033}{10.1016/j.csda.2008.11.033}.
+\end{cslreferences}
 

### 2020-10-18 ms/ms.tex
pre-release

@@
 
-  \textsuperscript{1} School of Life Sciences, University of Hawaii,  Honolulu,  Hawaii,  USA
+  \textsuperscript{1} School of Life Sciences, University of
+Hawai'i,  Honolulu,  Hawaii,  USA
   }
 
-  \def\corrAuthor{Christopher D. Muir}\def\corrAddress{,  }\def\corrEmail{\href{mailto:cdmuir@hawaii.edu}{\nolinkurl{cdmuir@hawaii.edu}}}
+  \def\corrAuthor{Christopher D.
+Muir}\def\corrAddress{,  }\def\corrEmail{\href{mailto:cdmuir@hawaii.edu}{\nolinkurl{cdmuir@hawaii.edu}}}
   \def\firstAuthorLast{Muir {et~al.}}
@@
 
-\title[Stomatal anatomy and pathogen colonization]{A stomatal model of anatomical tradeoffs between gas exchange and
-pathogen colonization}
+\title[Stomatal anatomy and pathogen colonization]{A stomatal model of
+anatomical tradeoffs between gas exchange and pathogen colonization}
 \author[\firstAuthorLast]{\Authors}
@@
 
-Stomatal pores control both leaf gas exchange and are one route for infection of internal plant tissues by many foliar pathogens, setting up the potential for tradeoffs between photosynthesis and defense.
+Stomatal pores control leaf gas exchange and are one route for infection of internal plant tissues by many foliar pathogens, setting up the potential for tradeoffs between photosynthesis and pathogen colonization.
 Anatomical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes could permanently reduce the gas exchange capacity for the life of the leaf.
@@
 A limitation of the model is that it does not include variation in innate immunity and stomatal closure in response to pathogens.
-Nevertheless, the model makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants.
+Nevertheless, the model makes predictions that can be tested with experiments and may explain variation in stomatal size and density among plants.
 The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for precise predictions.
@@
 to a leaf with large, low-density stomata (Papanatsiou et al., 2017).
-Leaves with stomata on both lower and upper surfaces (amphistomatic or
-amphistomatous) supply more CO\(_2\) to the mesophyll than hypostomatic
-(aka hypostomatous) leaves that only have stomata on the lower surface
-(Parkhurst, 1978; Gutschick, 1984; Parkhurst and Mott, 1990; Oguchi et
-al., 2018). In addition to anatomy, the pore area shrinks and expands in
-response to internal and external factors to regulate gas exchange
-dynamically (Buckley, 2019). For example, stomata typically open during
-the day and close at night in C\(_3\)/C\(_4\) plants, but the opposite
-is true for CAM plants. Shade, high vapor pressure deficits, dry soil
-and other factors can cause stomata to (partially) close even in the
-middle of the day. Variation in how stomata respond to internal and
-external signals may explain as much of the variation in gas exchange
-across leaves as anatomy (Lawson and Blatt, 2014).
+Leaves with stomata on both lower and upper surfaces (amphistomatous)
+supply more CO\(_2\) to the mesophyll than hypostomatous leaves that
+only have stomata on the lower surface (Parkhurst, 1978; Gutschick,
+1984; Parkhurst and Mott, 1990; Oguchi et al., 2018). In addition to
+anatomy, the pore area shrinks and expands in response to internal and
+external factors to regulate gas exchange dynamically (Buckley, 2019).
+For example, stomata typically open during the day and close at night in
+C\(_3\)/C\(_4\) plants, but the opposite is true for CAM plants. Shade,
+high vapor pressure deficits, dry soil and other factors can cause
+stomata to (partially) close even in the middle of the day. Variation in
+how stomata respond to internal and external signals may explain as much
+of the variation in gas exchange across leaves as anatomy (Lawson and
+Blatt, 2014).
 
@@
 not only determine the theoretical maximum stomatal conductance
-(\gsmax), but is also proportional to the operational stomatal
+(\gsmax), but are also proportional to the operational stomatal
 conductance (\gsop) in many circumstances (Franks et al., 2009, 2014;
@@
 
-In this section, I introduce a spatially explicit model of successful
-pathogen colonization on a leaf surface. I explain the model structure
-and assumptions here; the
-\protect\hyperlink{materials-and-methods}{Materials and Methods} section
-below describes how I analyzed the model to address the goals of the
-study. For generality, I refer to a generic ``pathogen'' that lands on
-leaf and moves to a stomate. The model is agnostic to the type of
-pathogen (virus, bacterium, fungus, etc.) and the specific biological
-details of how it moves. For example, motile bacterial cells can land
-and move around (Beattie and Lindow, 1995) whereas fungi may germinate
-from a cyst and grow until they form an appresorium for infection
-(Tucker and Talbot, 2001). These very different tropic movements on the
-leaf are treated identically here. I do not model photosynthesis
-explicitly, but assume that stomatal conductance limits carbon fixation,
-even though the relationship is nonlinear. I used Sympy version 1.6.1
-(Meurer et al., 2017) for symbolic derivations.
+For generality, I refer to a generic ``pathogen'\,' that lands on leaf
+and moves to a stomate. The model is agnostic to the type of pathogen
+(virus, bacterium, fungus, etc.) and the specific biological details of
+how it moves. For example, motile bacterial cells can land and move
+around (Beattie and Lindow, 1995) whereas fungi may germinate from a
+cyst and grow until they form an appresorium for infection (Tucker and
+Talbot, 2001). These very different tropic movements on the leaf are
+treated identically here. I do not model photosynthesis explicitly, but
+assume that stomatal conductance limits carbon fixation, even though the
+relationship is nonlinear. I used Sympy version 1.6.1 (Meurer et al.,
+2017) for symbolic derivations.
 
@@
 triangular grid with a density \(D\) and size (area) \(S\) on the
-abaxial surface only, since most leaves are hypostomatoues {[}Muir
+abaxial surface only, since most leaves are hypostomatous {[}Muir
 (2015); but see \protect\hyperlink{discussion}{Discussion}{]}. This
@@
 (R Core Team, 2020) for all analyses and wrote the paper in
-\textbf{rmarkdown} version 2.3 (Xie et al., 2018; Allaire et al., 2020).
+\textbf{rmarkdown} version 2.4 (Xie et al., 2018; Allaire et al., 2020).
 Citations for additional R software packages are in Appendix 2. Source
@@
 proportional to stomatal cover (\fs). At low \fs, \(p_\text{colonize}\)
-increases rapidly relative to \fs{} at first Figure \ref{fig:fig3}a). At
-higher \fs, \(p_\text{colonize}\) increases linearly with \fs. When
+increases rapidly relative to \fs{} at first (Figure \ref{fig:fig3}a).
+At higher \fs, \(p_\text{colonize}\) increases linearly with \fs. When
 pathogens persist indefinitely (\(H = 0\)), any combination of stomatal
@@
 \begin{minipage}[t]{0.50\columnwidth}\raggedright
-When selection against pathogen colonization stronger, the stomatal
+When selection against pathogen colonization is stronger, the stomatal
 size-density scaling exponent should be lower\strut
@@
 
-I would like to thank Athena McKown and an anonymous referee for
-valuable feedback that improved this manuscript.
+I would like to thank Athena McKown and Karl Fetter for valuable
+feedback that improved this manuscript.
 
@@
 
-I am grateful startup funds from the University of Hawaii for supporting
-this work.
+I am grateful startup funds from the University of Hawai'i for
+supporting this work.
 
@@
 \clearpage
+
+\hypertarget{appendix-1-spatially-implicit-model}{%
+\section*{Appendix 1: Spatially implicit
+model}\label{appendix-1-spatially-implicit-model}}
+\addcontentsline{toc}{section}{Appendix 1: Spatially implicit model}
+
+A limitation of the spatially explicit model is that a pathogen could
+only infect stomata in the focal triangle where it landed. Here I
+analyze an alternative, spatially implicit, model that relaxes this
+assumptions. Instead, I assume that a pathogen can potentially infect
+any stomate on the leaf. It searches through a random walk and has a
+continuous, constant probability of encountering a stomate that is
+determined by stomatal cover (\fs). If \(fs << 1\), this can be modeled
+as a homogeneous Poisson process and the distance \(x\) a pathogen must
+travel before reaching a stomate follows an exponential distribution:
+
+\begin{equation*}
+  f(x) = f_\text{S} e ^ {-f_\text{S} x}
+\end{equation*}
+
+Given the constant death rate per unit distance \(H\), the probability
+of surviving to distance \(x\) is \(e^{-H x}\). The probability of
+locating a stomate the probability of surviving to distance \(x\)
+multiplied by \(f(x)\) and integrated over all \(x\) from 0 to
+\(\infty\):
+
+\begin{align*}
+  p_\text{locate} & = \int_0^\infty f(x) e^{-Hx} dx \\
+                  & = \int_0^\infty f_\text{S} e^{-(f_\text{S} + H)x} dx \\
+                  & = \frac{f_\text{S}}{H + f_\text{S}}
+\end{align*}
+
+Substituting \(p_\text{locate}\) above into Equation
+\ref{eq:p_colonize}:
+
+\begin{align*}
+  p_\text{colonize} & = f_\text{S} + (1 - f_\text{S}) p_\text{locate} \\
+                    & = f_\text{S} + (1 - f_\text{S}) \frac{f_\text{S}}{H + f_\text{S}} \\
+                    & = f_\text{S} (1 + \frac{1 - f_\text{S}}{H + f_\text{S}})
+\end{align*}
+
+With the spatially implicit model, because pathogens can potentially
+reach any stomate on the leaf, \(p_\text{colonize}\) is greater than
+that in the spatially explicit model for the same value of \(H\). For
+example, if the pathogen can search forever (\(H = 0\)), then it will
+always colonize (\(p_\text{colonize} = 1\); Figure \ref{fig:figS2}a).
+But even when \(H > 0\), \(p_\text{colonize}\) is significantly higher
+than in the spatially implicit than spatially explicit model for the
+same \fs{} because pathogens can potentially colonize any stomate on the
+leaf.
+
+Whereas the spatially explicit model probably underestimates
+\(p_\text{locate}\) for pathogens that can search over long distances,
+the implicit model overestimates because it assumes that the probability
+of encountering a stomate is constant (i.e.~homogeneous Poisson
+process). This is not true because stomata are discrete areas on the
+leaf. If a pathogen is searching far away from a stomate, its
+probability of encountering a stomate in the near future is lower than
+that for a pathogen searching near a stomate. This should be modeled as
+a \emph{nonhomogenous} Poisson process. Future work should derive
+\(p_\text{locate}\) for the stomatal anatomies presented here under a
+nonhomogeneous process.
+
+Despite the quantitative differences in the the spatially explicit and
+implicit models, they have similar qualitative properties when
+\(H > 0\), which is reasonable since the leaf surface is a relatively
+hostile environment for most pathogens (see
+\protect\hyperlink{introduction}{Introduction}). In both models,
+\(p_\text{colonize}\) increases with, but is higher than \fs. In the
+spatially explicit model, size-density scaling that preserves
+\(p_\text{colonize}\) is 1 when \(H = 0\) and slightly less than 1
+otherwise (Figure \ref{fig:fig4}). In the spatially implicit model, the
+scaling coefficient is always 1. Rearranging the equation for
+\(p_\text{colonize}\) above and substituting \(f_\text{S} = DS\), the
+following relationship holds:
+
+\[SD = \frac{p_\text{colonize} H}{1 - p_\text{colonize} + H}\] Since
+\(H\) is a constant the right-hand side of the equation above is
+constant for a given value of \(p_\text{colonize}\). Hence the
+\(\beta_p\) that would preserve the relationship above is simply 1
+(Figure \ref{fig:figS3}).
 
@@
 
-\hypertarget{appendix-1-spatially-implicit-model}{%
-\section*{Appendix 1: Spatially implicit
-model}\label{appendix-1-spatially-implicit-model}}
-\addcontentsline{toc}{section}{Appendix 1: Spatially implicit model}
-
-A limitation of the spatially explicit model is that a pathogen could
-only infect stomata in the focal triangle where it landed. Here I
-analyze an alternative, spatially implicit, model that relaxes this
-assumptions. Instead, I assume that a pathogen can potentially infect
-any stomate on the leaf. It searches through a random walk and has a
-continuous, constant probability of encountering a stomate that is
-determined by stomatal cover (\fs). If \(fs << 1\), this can be modeled
-as a homogeneous Poisson process and the distance \(x\) a pathogen must
-travel before reaching a stomate follows an exponential distribution:
-
-\begin{equation*}
-  f(x) = f_\text{S} e ^ {-f_\text{S} x}
-\end{equation*}
-
-Given the constant death rate per unit distance \(H\), the probability
-of surviving to distance \(x\) is \(e^{-H x}\). The probability of
-locating a stomate the probability of surviving to distance \(x\)
-multiplied by \(f(x)\) and integrated over all \(x\) from 0 to
-\(\infty\):
-
-\begin{align*}
-  p_\text{locate} & = \int_0^\infty f(x) e^{-Hx} dx \\
-                  & = \int_0^\infty f_\text{S} e^{-(f_\text{S} + H)x} dx \\
-                  & = \frac{f_\text{S}}{H + f_\text{S}}
-\end{align*}
-
-Substituting \(p_\text{locate}\) above into Equation
-\ref{eq:p_colonize}:
-
-\begin{align*}
-  p_\text{colonize} & = f_\text{S} + (1 - f_\text{S}) p_\text{locate} \\
-                    & = f_\text{S} + (1 - f_\text{S}) \frac{f_\text{S}}{H + f_\text{S}} \\
-                    & = f_\text{S} (1 + \frac{1 - f_\text{S}}{H + f_\text{S}})
-\end{align*}
-
-With the spatially implicit model, because pathogens can potentially
-reach any stomate on the leaf, \(p_\text{colonize}\) is greater than
-that in the spatially explicit model for the same value of \(H\). For
-example, if the pathogen can search forever (\(H = 0\)), then it will
-always colonize (\(p_\text{colonize} = 1\); Figure \ref{fig:figS2}a).
-But even when \(H > 0\), \(p_\text{colonize}\) is significantly higher
-than in the spatially implicit than spatially explicit model for the
-same \fs{} because pathogens can potentially colonize any stomate on the
-leaf.
-
-Whereas the spatially explicit model probably underestimates
-\(p_\text{locate}\) for pathogens that can search over long distances,
-the implicit model overestimates because it assumes that the probability
-of encountering a stomate is constant (i.e.~homogeneous Poisson
-process). This is not true because stomata are discrete areas on the
-leaf. If a pathogen is searching far away from a stomate, its
-probability of encountering a stomate in the near future is lower than
-that for a pathogen searching near a stomate. This should be modeled as
-a \emph{nonhomogenous} Poisson process. Future work should derive
-\(p_\text{locate}\) for the stomatal anatomies presented here under a
-nonhomogeneous process.
-
-Despite the quantitative differences in the the spatially explicit and
-implicit models, they have similar qualitative properties when
-\(H > 0\), which is reasonable since the leaf surface is a relatively
-hostile environment for most pathogens (see
-\protect\hyperlink{introduction}{Introduction}). In both models,
-\(p_\text{colonize}\) increases with, but is higher than \fs. In the
-spatially explicit model, size-density scaling that preserves
-\(p_\text{colonize}\) is 1 when \(H = 0\) and slightly less than 1
-otherwise (Figure \ref{fig:fig4}). In the spatially implicit model, the
-scaling coefficient is always 1. Rearranging the equation for
-\(p_\text{colonize}\) above and substituting \(f_\text{S} = DS\), the
-following relationship holds:
-
-\[SD = \frac{p_\text{colonize} H}{1 - p_\text{colonize} + H}\] Since
-\(H\) is a constant the right-hand side of the equation above is
-constant for a given value of \(p_\text{colonize}\). Hence the
-\(\beta_p\) that would preserve the relationship above is simply 1
-(Figure \ref{fig:figS3}).
-
-\clearpage
-
 \hypertarget{references}{%
@@
 \hypertarget{refs}{}
+\begin{cslreferences}
 \leavevmode\hypertarget{ref-allaire_rmarkdown:_2020}{}%
@@
 doi:\href{https://doi.org/10.1016/j.copbio.2010.05.006}{10.1016/j.copbio.2010.05.006}.
+\end{cslreferences}
 

### 2020-10-18 r/00_run-python.R

@@ -1,7 +1,9 @@
 # The output from this script is included with the repo, so there is no need to rerun, but I have provided the code for transparency
+# install.packages("glue")
+# install.packages("reticulate")
 library(glue)
 library(reticulate)
-use_python('/anaconda3/bin/python', required = TRUE)
+use_python('/anaconda3/bin/python', required = TRUE) # you may need to provide a different path to Python
 source_python("python/tangents.py")
 
 get_tangent_slopes <-
@@ -24,9 +26,3 @@ get_tangent_points <-
      }}", x1 = x1, x2 = x2, y1 = y1, y2 = y2) 
 
 cat(get_tangent_slopes, "\n", get_tangent_points, file = "r/functions-from-python.R")
-
-
-library(glue)
-library(reticulate)
-# use_python('/anaconda3/bin/python', required = TRUE)
-source_python("python/spatially-implicit-model.py")

### 2020-10-18 r/01_make-fig1-2.R

@@ -212,8 +212,8 @@ fig2a <- pD %>%
     mapping = aes(label = label),
     hjust = 0.5, vjust = 0, nudge_x = 0, nudge_y = -2.75 * R
   ) +
-  scale_x_continuous(expand = expand_scale(mult = 0.125)) +
-  scale_y_continuous(expand = expand_scale(mult = 0.125)) +
+  scale_x_continuous(expand = expansion(mult = 0.125)) +
+  scale_y_continuous(expand = expansion(mult = 0.125)) +
   geom_circle(
     data = df_stomata,
     mapping = aes(x0 = x, y0 = y, r = R),

### 2020-10-18 r/header.R

@@ -10,10 +10,13 @@ library(ggimage)
 library(gginnards)
 library(ggplot2)
 library(glue)
+library(knitcitations)
 library(magrittr)
 library(pracma)
 library(purrr)
 library(readr)
+library(RefManageR)
+library(sessioninfo)
 library(stringr)
 library(tikzDevice)
 library(tidyr)

### 2020-10-18 ms/ms.Rmd
pre-release

@@
     department: School of Life Sciences
-    institution: University of Hawaii
+    institution: University of Hawai'i
     city: Honolulu
@@
 However, sparse clusters of small stomata could allow a leaf with low rates of gas exchange to have faster stomatal response compared to a leaf with large, low-density stomata [@papanatsiou_stomatal_2017].
-Leaves with stomata on both lower and upper surfaces (amphistomatic or amphistomatous) supply more CO$_2$ to the mesophyll than hypostomatic (aka hypostomatous) leaves that only have stomata on the lower surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990; @oguchi_leaf_2018].
+Leaves with stomata on both lower and upper surfaces (amphistomatous) supply more CO$_2$ to the mesophyll than hypostomatous leaves that only have stomata on the lower surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990; @oguchi_leaf_2018].
 In addition to anatomy, the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019].
@@
 For simplicity, I model a hypostomatous leaf, but consider the implications for amphistomatous leaves in the [Discussion].
-Stomatal size and density not only determine the theoretical maximum stomatal conductance (\gsmax), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019].
+Stomatal size and density not only determine the theoretical maximum stomatal conductance (\gsmax), but are also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019].
 \gsop{} is the actual stomatal conductance of plants in the field and is almost always below \gsmax{} because stomata are usually not fully open.
@@
 
-In this section, I introduce a spatially explicit model of successful pathogen colonization on a leaf surface.
-I explain the model structure and assumptions here; the [Materials and Methods] section below describes how I analyzed the model to address the goals of the study.
 For generality, I refer to a generic ``pathogen'' that lands on leaf and moves to a stomate.
@@
 Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015].
-Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$ on the abaxial surface only, since most leaves are hypostomatoues [@muir_making_2015; but see [Discussion]].
+Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$ on the abaxial surface only, since most leaves are hypostomatous [@muir_making_2015; but see [Discussion]].
 This assumption ignores veins, trichomes, and within-leaf variation in stomatal density.
@@
 The probability of colonization ($p_\text{colonize}$) is not simply proportional to stomatal cover (\fs).
-At low \fs, $p_\text{colonize}$ increases rapidly relative to \fs{} at first Figure \ref{fig:fig3}a).
+At low \fs, $p_\text{colonize}$ increases rapidly relative to \fs{} at first (Figure \ref{fig:fig3}a).
 At higher \fs, $p_\text{colonize}$ increases linearly with \fs.
@@
 drier, higher light)
-When selection against pathogen colonization stronger, the stomatal size-density scaling exponent should be lower | Measure stomatal anatomy in environments that differ in pathogen colonization using comparative or experimental approaches
+When selection against pathogen colonization is stronger, the stomatal size-density scaling exponent should be lower | Measure stomatal anatomy in environments that differ in pathogen colonization using comparative or experimental approaches
 
@@
 
-I would like to thank Athena McKown and an anonymous referee for valuable feedback that improved this manuscript.
+I would like to thank Athena McKown and Karl Fetter for valuable feedback that improved this manuscript.
 
@@
 
-I am grateful startup funds from the University of Hawaii for supporting this work.
+I am grateful startup funds from the University of Hawai'i for supporting this work.
 
@@
 
-## Figures {-}
-
-\begin{figure}
-  \centering
-	\includegraphics{../figures/figS1.pdf}
-	\caption{\textbf{The probability of colonization increases with both stomatal size ($S$) and density $D$} I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of $S$, $D$, and $H$ (see [Materials and Methods]) \textbf{a.} Each line shows how $p_\text{colonize}$ increases with $S$ ($x$-axis, log-scale) for selected values of $D \in \{10, 100, 1000\}~\text{mm}^{-2}$.
-\textbf{b.} Each line shows how $p_\text{colonize}$ increases with $D$ ($x$-axis, log-scale) for selected values of $S \in \{10, 100, 1000\}~\mu\text{m}^2$.
-The facets show results for different values of $H$}
-	\label{fig:figS1}
-\end{figure}
-
-\begin{figure}
-  \centering
-	\includegraphics{../figures/figS2.pdf}
-	\caption{\textbf{The probability of colonization increases with both stomatal cover and conductance in the spatially implicit model.} As in Figure \ref{fig:fig3}, I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of stomatal densities and sizes (see [Materials and Methods]), but a subset of results are shown here.
-Stomatal size and density determine stomatal cover (\fs; Equation \ref{eq:fs}) and theoretical maximum stomatal conductance (\gsmax; Equation \ref{eq:gsmax}).
-\textbf{a.} $p_\text{colonize}$ initially increases rapidly with \fs{} ($x$-axis), then slows down to a linear relationship.
-Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$).
-The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density for all values of $H$, which is why the lines overlap.
-\textbf{b.} $p_\text{colonize}$ increases sigmoidally with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is lower at higher densities for a given \gsmax.
-The relationship between \gsmax{} and $p_\text{colonize}$ is similar for all values of $H > 0$.}
-	\label{fig:figS2}
-\end{figure}
-
-\begin{figure}
-  \centering
-	\includegraphics{../figures/figS3.pdf}
-	\caption{\textbf{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$) in the spatially implicit model}.
-As in Figure \ref{fig:fig4}, in each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.25 (lowest line), 0.5, or 0.75 (highest line).
-For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.5$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region.
-The scaling exponent is unity $\beta = 1$ when $H > 0$.}
-	\label{fig:figS3}
-\end{figure}
-
-\clearpage
-
 # Appendix 1: Spatially implicit model {-}
@@
 
+## Figures {-}
+
+\begin{figure}
+  \centering
+	\includegraphics{../figures/figS1.pdf}
+	\caption{\textbf{The probability of colonization increases with both stomatal size ($S$) and density $D$} I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of $S$, $D$, and $H$ (see [Materials and Methods]) \textbf{a.} Each line shows how $p_\text{colonize}$ increases with $S$ ($x$-axis, log-scale) for selected values of $D \in \{10, 100, 1000\}~\text{mm}^{-2}$.
+\textbf{b.} Each line shows how $p_\text{colonize}$ increases with $D$ ($x$-axis, log-scale) for selected values of $S \in \{10, 100, 1000\}~\mu\text{m}^2$.
+The facets show results for different values of $H$}
+	\label{fig:figS1}
+\end{figure}
+
+\begin{figure}
+  \centering
+	\includegraphics{../figures/figS2.pdf}
+	\caption{\textbf{The probability of colonization increases with both stomatal cover and conductance in the spatially implicit model.} As in Figure \ref{fig:fig3}, I simulated the probability of colonization ($p_\text{colonize}$, $y$-axis) over a range of stomatal densities and sizes (see [Materials and Methods]), but a subset of results are shown here.
+Stomatal size and density determine stomatal cover (\fs; Equation \ref{eq:fs}) and theoretical maximum stomatal conductance (\gsmax; Equation \ref{eq:gsmax}).
+\textbf{a.} $p_\text{colonize}$ initially increases rapidly with \fs{} ($x$-axis), then slows down to a linear relationship.
+Overall, $p_\text{colonize}$ is lower when pathogens can die on the leaf surface ($H > 0$).
+The relationship between \fs{} and $p_\text{colonize}$ is the same regardless of stomatal density for all values of $H$, which is why the lines overlap.
+\textbf{b.} $p_\text{colonize}$ increases sigmoidally with \gsmax{} at all stomatal densities, but $p_\text{colonize}$ is lower at higher densities for a given \gsmax.
+The relationship between \gsmax{} and $p_\text{colonize}$ is similar for all values of $H > 0$.}
+	\label{fig:figS2}
+\end{figure}
+
+\begin{figure}
+  \centering
+	\includegraphics{../figures/figS3.pdf}
+	\caption{\textbf{Log-log scaling relationships between stomatal density ($D$, \textit{x}-axis) and size ($S$, \textit{y}-axis) that preserve the probability of colonization ($p_\text{colonize}$) in the spatially implicit model}.
+As in Figure \ref{fig:fig4}, in each panel, solid lines indicate values of $D$ and $S$ where $p_\text{colonize}$ is 0.25 (lowest line), 0.5, or 0.75 (highest line).
+For reference, dashed grey lines show scaling relationships that preserve \fs{} ($\beta = 1$, $\text{slope} = - 1 / \beta = -1$) and \gsmax{} ($\beta = 0.5$, $\text{slope} = - 1 / \beta = -2$) drawn through the centroid of the plotting region.
+The scaling exponent is unity $\beta = 1$ when $H > 0$.}
+	\label{fig:figS3}
+\end{figure}
+
 \clearpage

### 2020-10-18 r/05_plot-scaling.R

@@ -128,5 +128,5 @@ figS3 <- ggplot(df, aes(D, S, group = p_colonize)) +
 figS3
 
 # ggsave("figures/figS3.jpeg", w = 4.5, h = 4, units = "in")
-ggsave("figures/figS3.pdf", w = 4.5, h = 4, units = "in")
+ggsave("figures/figS3.pdf", w = 5.5, h = 4, units = "in")
 

### 2020-10-18 r/07_make-appendix2.R

@@ -1,8 +1,5 @@
 source("r/header.R")
 
-library(knitcitations)
-library(RefManageR)
-library(sessioninfo)
 si <- session_info(include_base = FALSE)
 
 r_package_table <- map2_dfr(si$packages$package, si$packages$loadedversion, ~ {

### 2020-10-18 r/08_write-bib.R

@@ -1,0 +1,4 @@
+source("r/header.R")
+
+# Download bibliography from Zotero
+bib <- ReadZotero("5814932", .params = list(key = "ED0yHWd4Qtd6ereWuiLgYN2p", collection = "T93F8SNM", limit = 1000), temp.file = "ms/stomata-tradeoff.bib", delete.file = FALSE)

### 2020-10-18 r/09_render-ms.R

@@ -1,0 +1,9 @@
+rmarkdown::render(
+  input = "ms/ms.Rmd",
+  output_dir = "ms"
+)
+
+rmarkdown::render(
+  input = "ms/appendix2.Rmd",
+  output_dir = "ms"
+)

### 2020-10-18 r/functions.R

@@ -149,7 +149,6 @@ get_theta <- function(x_p, y_p, x_i, y_i, R) {
 # Probability of locating stomate at starting position x_p, y_p
 plocate_xy <- function(x_p, y_p, R, U, H) {
   
-  # params <- list(...)
   x_c <- c(0, U / 2, U)
   y_c <- c(0, sqrt(3) / 2 * U, 0)
   ymin <- get_ymin(x_p, R, U)

### 2020-10-18 r/install-packages.R

@@ -3,6 +3,7 @@ install.packages("rmarkdown")
 install.packages("rticles")
 
 # Packages for running R scripts
+install.packages("checkmate")
 install.packages("cowplot")
 install.packages("dplyr")
 install.packages("furrr")
@@ -11,10 +12,13 @@ install.packages("ggimage")
 install.packages("gginnards")
 install.packages("ggplot2")
 install.packages("glue")
+install.packages("knitcitations")
 install.packages("magrittr")
 install.packages("pracma")
 install.packages("purrr")
 install.packages("readr")
+install.packages("RefManageR")
+install.packages("sessioninfo")
 install.packages("stringr")
 install.packages("tikzDevice")
 install.packages("tidyr")
